### 2023-03-06 BB_endo.Rmd
first commit

---
title: "Endogeneous infant mortality"
author: "Göran Broström"
date: "January 6, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

It is convenient to transform data according to the "log-cube transformation"

$$
g(T) = \log^3(1 + 365 T), \quad 0 \le T \le 1,
$$
before fitting survival models, because on that scale, the *BB* model prescribes the *exponential* distribution. Furthermore, we are interested in splitting the first year of life into two periods, the first month and the last eleven. On the new time scale, the cut point is $\log^3(1 + 30)$ =  `r round((log(1 + 30))^3, 1)`.

```{r transform}
g <- function(t) (log(1 + 365 * t))^3 # the log-cube transform 
library(skum)
boys <- obs[obs$birthdate > 1832 & obs$birthdate < 1890 & obs$sex == "male", ]
boys$event <- boys$sluttyp == 2 # death
boys <- boys[, c("id", "region", "birthdate", "enter", "exit", "event")]
boys <- age.window(boys, c(0, 1)) # first year of life
boys$enter <- g(boys$enter) # transform 'enter'
boys$exit <- g(boys$exit) # transform 'exit'
boys1 <- age.window(boys, c(0, g(1/12))) # first month
boys2 <- age.window(boys, c(g(1/12), g(1))) # last 11 months
```

## The Bengtsson-Broström model

So let us check the *Bengtsson-Broström* (BB) model on *boys2*, the last eleven months.

```{r checkbb2}
fitw <- phreg(Surv(enter - g(1/12), exit - g(1/12), event) ~ I(birthdate - 1900) + region, 
              data = boys2, dist = "weibull")
fitc <- coxreg(Surv(enter - g(1/12), exit - g(1/12), event) ~ I(birthdate - 1900) + region,
               data = boys2)
check.dist(fitc, fitw, col = c("blue", "red"))
```

An extraordinary excellent fit to the *Weibull* distribution! But the BB model prescribes an exponential distribution, which corresponds to a Weibull with *shape = 1*. So let us look at the parameter estimates.

```{r}
summary(fitw)
```

the shape parameter is estimated to `r exp(fitw$coef[4])`, which is as close to one as you can get. A complete success for the *BB* model.

Now let us look at the first month of life:

```{r checkbb1}
fitw1 <- phreg(Surv(enter, exit, event) ~ I(birthdate - 1900) + region, 
               data = boys1, dist = "weibull")
fitc1 <- coxreg(Surv(enter, exit, event) ~ I(birthdate - 1900) + region, 
                data = boys1)
check.dist(fitc1, fitw1, col = c("blue", "red"))
```

Not so convincing, (i) the fit to a Weibull is slightly worse, and (2) the cumulative hazards are not straight lines, implying deviation from the exponential distribution.

## The Bourgeouis-Pichat suggestion about exogeneity

But this is what Bourgeouis-Pichat is predicting: During months 2--12 mortality is strictly exogeneous, and we can expect a uniform (BB: exponential) distribution, while during the first month there are both exogeneous and endogeneous causes of death, and deviation from the exponential distribution is to be expected.

The estimated coefficients are

```{r resend}
summary(fitw1)
```

an we can see that the *shape* parameter is significantly different from one.

One way, the PB way(?), to estimate the contribution of the endogeneous deaths is to assume that the exogeneous mortality is exponential during all twelve months, and extrapolate.

```{r extrapo, fig.cap = "Skellefteå and Umeå boys, 1820-1950."}
fitex <- phreg(Surv(enter, exit, event) ~ I(birthdate - 1900) + region, 
               data = boys2, dist = "weibull")
fitee <- coxreg(Surv(enter, exit, event) ~ I(birthdate - 1900) + region, 
                data = boys1)
plot(fitex, xlim = c(0, g(1/12)), col = "blue", fn = "cum", 
     ylim = c(0, 0.03), main = "Neonatal mortality")
y <- c(0, cumsum(fitee$hazards[[1]][, 2]))
x <- c(0, fitee$hazards[[1]][, 1])
lines(x, y, col = "red")
abline(h = 0)
text(10, 0.020, "Total mortality", col = "red")
text(25, 0.012, "Exogeneous mortality", col = "blue")
```

Obviously the endogeneous mortality dominates the first month of life. (We are still in the transformed time scale, with "40" representing 1 month.)

## Without the transformation

We repeat everything, but without transforming.

```{r transform2}
g <- function(t) t # the identity transform! 
library(skum)
boys <- obs[obs$birthdate > 1820 & obs$sex == "male", ]
boys$event <- boys$sluttyp == 2 # death
boys <- boys[, c("id", "region", "birthdate", "enter", "exit", "event")]
boys <- age.window(boys, c(0, 1)) # first year of life
boys$enter <- g(boys$enter) # transform 'enter'
boys$exit <- g(boys$exit) # transform 'exit'
boys1 <- age.window(boys, c(0, g(1/12))) # first month
boys2 <- age.window(boys, c(g(1/12), g(1))) # last 11 months
```

### The Bengtsson-Broström model, II

So let us check the *Bengtsson-Broström* (BB) model on *boys2*, the last eleven months.

```{r checkbb22}
fitw <- phreg(Surv(enter - g(1/12), exit - g(1/12), event) ~ I(birthdate - 1900) + region, 
              data = boys2, dist = "weibull")
fitc <- coxreg(Surv(enter - g(1/12), exit - g(1/12), event) ~ I(birthdate - 1900) + region,
               data = boys2)
check.dist(fitc, fitw, col = c("blue", "red"))
```

Not an equally excellent fit to the *Weibull* distribution!

And, add (12 jan 2020) an experiment with the Gompertz distribution:

```{r gomp}
fitg <- phreg(Surv(enter - g(1/12), exit - g(1/12), event) ~ I(birthdate - 1900) + region, 
              data = boys2, dist = "gompertz", param = "rate")
check.dist(fitc, fitg, col = c("blue", "red"), main = "Gompertz")
```

## Conclusion

Maybe it is too early to completely reject the log-cube transformation (and the BB model) for infant mortality.

### 2023-03-06 GBTB23.Rmd
first commit

---
title: "A hazards approach to the biometric analysis of infant mortality"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: bb.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage{a4wide}
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
- \subsectionfont{\normalfont\itshape}
abstract: "A variation of the Bourgeois-Pichat biometric analysis of infant mortality is suggested. In the original model, cumulative mortality in the last eleven months of infancy is assumed to follow a uniform distribution given a log-cube transformation of age. Instead, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. We argue that our assumption is more satisfactory from a theoretical point of view, since it focus on the conditional probability of dying. Our model gives different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is slightly better, especially in cases with higher levels of infant mortality."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 4)
##requireNamespace("eha")
suppressWarnings(library(kableExtra))
library(eha)
g <- function(x) (log(x+1))^3 * 365 / (log(366))^3
##ginv <- function(x) 1 - exp(x * ((log(366)^(1/3)) / 365))
ginv <- function(y) exp((y / 365)^(1/3) * log(366)) - 1
##g0 <- function(x) g(x) * 365 / g(365)
dg <- function(x) 3 * (log(1 + x))^2 / (1 + x) * 365 / (log(366))^3
```

# Introduction

```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.height = 7}
x <- seq(0, 365, length = 366)
plot(x, g(x), type = "l", xlab = "Days", ylab = "g(Days)", 
     col = "blue", axes = FALSE)
att <- c(0, 30, 183, 365)
axis(1, at = att, labels = att)
axis(2, las = 1, at = g(att), labels = c("0", "g(30)", "g(183)", "365"))
box()
lines(c(0, 365), c(0, 365), lty = 2)
abline(v = c(0, 365), h = c(0, 365))
abline(v = 30, h = g(30), col = "darkgreen", lty = 3)
```

Lois Bourgeois-Pichat saw the first year of life not only as the period when mortality is highest, but also the period when improvements should be possible 
[@bp51a; @bp51b; @bp52]. He argued that infant deaths should be divided into two categories, endogenous and exogenous deaths. Endogenous deaths are, by definition, due to inherited factors, or acquired during gestation or delivery. They typically occur at the initial period of life, though sometimes much later and include congenital debility, prematurity, malformations, and disease in early life. Exogenous infant mortality, which Bourgeois-Pichat regards as accidental, are deaths for which the society must hold itself responsible and to reduce endogenous mortality, medical intervention is essential [@bp52]. Since it is, based on causes of deaths, difficult to distinguish between endogenous and exogenous factors, either because they are inaccurate or difficult to make due to the diseases themself, or the information do not exist, Bourgeois-Pichat offered a solution to the problem.

Bourgeois-Pichat’s biometric model to differentiate between endogenous and exogenous mortality is based on an idea of a universal law governing the
distribution of deaths in the first year of life [@bp52]. He assumes that all deaths taking place in the last eleven months of the first year are exogenous. He argues that, although there are also some endogenous mortalities during this period, they do not affect the development of mortality much. Supported by examples from mid-twentieth century Western and Southern Europe, the US and Canada, New Zeeland, and other countries, Bourgeois-Pichat argued that the cumulative infant deaths after the first month follows a linear development given a log-cube transformation of age [@bp52]. Mortality in the first month of life is then divided into an endogenous and an exogenous assuming that the same linearity also exists in this period. It leads to the conclusion that exogenous deaths in the first month of life represent 25 percent of the deaths from the second to twelfth months [@bp52].

Applying this method, Bourgeois-Pichat also finds deviations from the linear death pattern after the first month. To take one example, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12]. To take another example, in the case of Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8]. A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec. The curve for Sweden 1910--1946 show, however, no such downward bend [@bp51b, Figure 9; @bp52].

Bourgeois-Pichat's biometric method has also been used to identify problems with data recording [@wea77; @weasr81]. While a high level of endogenous mortality cannot be used as a criterion for high data quality, the opposite holds true. A very low level of endogenous mortality indeed indicates data problems, just like large deviations from the normal sex ratio at birth do. These two criteria are, in fact, often used in historical studies to evaluate data recording (see Bengtsson et al 2004). They become even more precise if they are applied to different social strata, since under-recording of early deaths often have a social gradient, possibly due to the costs involved in a burial [@tb99; @tbcl99].

We suggest a variation of the Bourgeois-Pichat biometric analysis of infant mortality. Instead of assuming that the cumulative mortality in the last eleven months of infancy follows a uniform distribution, given a log-cube transformation of age, we assume an exponential distribution. The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk. This assumption is more satisfactory from a theoretical point of view, since it models the conditional probability of dying. The growth in birth weight also follows this distribution [@bp51a; @bp51b]. In addition, the model ﬁt is slightly better for our model, especially in cases with high level of infant mortality, common in pre-modern societies. The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.1 In addition, we give examples from eighteenth Sweden, where the curve do not follow the uniform distribution during the last eleven month of infancy, despite high levels of mortality in the first month of life.

# The Bourgeois-Pichat procedure

Central in the procedure suggested by @bp51a is *the log-cube transform*,
see Equation \@ref(eq:logcube).

\begin{equation} (\#eq:logcube)
g(t) = C \log^3(t + 1), \; 0 \le t \le 365,
\end{equation}
where $t$ is age measured in days and $C$ is a normalizing constant,
\begin{equation*}
C = \frac{365}{\log^3(366)}.
\end{equation*}

The constant $C$ is chosen so that $g(365) = 365$, see Figure \@ref(fig:logcube).
Note that $C$ is *not* part of the original definition of the log-cube transform,
but provided here only to make graphical comparisons easier to interpret. 
It makes no difference otherwise.

Assume that a cohort of infants is followed over time from birth to age one. There are no drop-outs 
(no right censoring). The exact age in days at each observed death is noted and transformed by $g$, 
and due to imperfect time measuring there may be tied death ages. As an illustrative example, 
we use a data set from northern Sweden, covering the years 1861--1950.
```{r readdata}
infant <- readRDS("Rdata/infant.rds")
infant$period <- cut(infant$birthdate, 
                   breaks = as.Date(c("1860-12-31", "1890-12-31", 
                                      "1920-12-31", "1951-01-01")),
                   labels = c("1861-1890", "1891-1920", "1921-1950"))
infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "icd.chapter")]
ginfant <- infant
ginfant$enter <- g(infant$enter)
ginfant$exit <- g(infant$exit)
```
For the first and last 30 years in this data set, we have the results shown in Figure \@ref(fig:first).

```{r first, fig.cap = "The periods 1861--1890 and 1921--1950, Bourgeois-Pichat method."}
source("R/drawBP.R")
oldpar <- par(mfrow = c(1,2))
##
r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
r1 <- eha::risksets(r1, members = FALSE)
drawBP(r1, main = "1861-1890", ylim = c(0, 0.1))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
r3 <- eha::risksets(r3, members = FALSE)
drawBP(r3, main = "1921-1950", ylim = c(0, 0.1))
par(oldpar)
```

The cumulative numbers of death are plotted against the
corresponding death ages on the $g$ time scale, and as is seen, the fit to linearity
after 30 days (72 on the $g$ scale) is excellent for both time periods.

# The hazard-based procedure

Instead of using the total number of births throughout in the denominator oas in  the
Bourgeois-Pichat procedure, we suggest 
using the present riskset size, that is, the the total number of infants still alive 
immediately prior to the death age under investigation. It is easily achieved by the use of 
the function `risksets` in the package `eha`. See Table \@ref(tab:risksets) for 
an illustration.
```{r risksets, echo = FALSE, results = 'asis'}
source("R/tbl.R")
lt1 <- with(r1, cbind(ginv(risktimes), round(risktimes, 1), size, n.events))
lt1 <- data.frame(lt1)
names(lt1) <- c("Day", "g(Day)", "Population", "Deaths")
tbl(lt1[1:7, ], caption = "Life table, 1861-1890.")
```
It starts with `r as.integer(lt1[1, 3])` live births, of which `r lt1[1, 4]` dies on the day of birth, 
on avareage at the age of six hours (a quarter of a day), and so on.

The plot corresponding to Figure \@ref(fig:first) is shown in Figure \@ref(fig:second).

```{r second, fig.cap = "The periods 1861--1890 and 1921--1950, hazard method."}
source("R/drawBB.R")
oldpar <- par(mfrow = c(1,2))
##
##r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
##r1 <- eha::risksets(r1, members = FALSE)
drawBB(r1, main = "1861-1890", ylim = c(0, 0.115))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
##r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
##r3 <- eha::risksets(r3, members = FALSE)
drawBB(r3, main = "1921-1950", ylim = c(0, 0.115))
par(oldpar)
```

See Figure \@ref(fig:both) for a comparison of the two curves.

```{r both, fig.cap = "Bourgeois-Pichat and 'hazard method' plots."}
oldpar = par(mfrow = c(1, 2))
plot(r1$risktimes, cumsum(r1$n.events) / r1$size[1], type = "l", col = "blue",
     main = "1861-1890", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
lines(r1$risktimes, cumsum(r1$n.events / r1$size), type = "l", col = "red")
abline(h = 0, v = 0)
text(120, 0.08, "Hazard method", col = "red")
text(200, 0.02,"Bourgeois-Pichat", col = "blue")
##
plot(r3$risktimes, cumsum(r3$n.events) / r3$size[1], type = "l", col = "blue",
     main = "1921-1950", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
lines(r3$risktimes, cumsum(r3$n.events / r3$size), type = "l", col = "red")
abline(h = 0, v = 0)
text(120, 0.05, "Hazard method", col = "red")
text(200, 0.015,"Bourgeois-Pichat", col = "blue")
##
##
par(oldpar)
```

The difference between the two curves increases as infant mortality increases, 
but both methods show an excellent fit to linearity.

# Theoretical considerations

We note that the Bourgeois-Pichat method requires that no right censorings (infants lost 
to follow-up before one year of age) are present, and no left truncations (infants under observation 
only from an age later than birth). The hazards approach, on the other hand, allows
left truncation and right censoring as long as they are *non-informative*. Often, though,
this not very important, since new-born babies tend to be geographically stationary
for their first year of life.

## Post-neonatal mortality on the $g$ scale

The following discussion assumes that age is measured on the $g$ scale. We concentrate on the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).

In order to see this, the data set from above is *left truncated* at age 30 days (at 72 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 30.

```{r postneo, fig.cap = "Exponential fits to postneonatal data."}
oldpar <- par(mfrow = c(1, 2))
post <- age.window(ginfant, c(g(30), g(365)))
p1 <- post[post$period == "1861-1890", ]
p3 <- post[post$period == "1921-1950", ]
fit1 <- coxreg(Surv(exit - enter, event) ~ 1, data = p1)
wfit1 <- phreg(Surv(exit - enter, event) ~ 1, data = p1, shape = 1)
check.dist(fit1, wfit1, main = "1861-1890")
##
fit3 <- coxreg(Surv(exit - enter, event) ~ 1, data = p3)
wfit3 <- phreg(Surv(exit - enter, event) ~ 1, data = p3, shape = 1)
check.dist(fit3, wfit3, main = "1921-1950")
par(oldpar)
```

The *Exponential* fit is excellent, with a slight edge for the early period data, 
where postneonatal mortality is high. However, one question pops up: If the exponential 
assumption is in fact true, will it also hold for subgroups of the data?

The theoretical ansver is "No", because a mixture of two exponential populations with
different mortality rates cannot itself be exponential. A heuristic argument for that: Assume two equally-sized 
groups at "birth", but with different levels of mortality. As time passes, there will be more deaths 
in the high-mortality group, and that means that the proportion survivors in the low-level group
will increase. Therefore, the population-level mortality will seem to decrease over time, 
and not being constant, as the exponentiality prescribes.

We can see what happens when the two periods in Figure \@ref(fig:postneo) are joined, see Figure
\@ref(fig:joined).

```{r joined, fig.cap = "Exponential fit for two periods."}
p13 <- post[post$period != "1891-1920", ]
p13$period <- factor(p13$period)
fit13 <- coxreg(Surv(exit - enter, event) ~ 1, data = p13)
wfit13 <- phreg(Surv(exit - enter, event) ~ 1, data = p13, shape = 1)
check.dist(fit13, wfit13, main = "1861-1890, 1921-1950")
#
```

Not what we expected: The fit seems to be extremely good in both figures! However,
the rate is close to zero, and so differences are small and hard to notice. And for practical use, they are not interesting.

What about the original Bourgeois-Pichat model, is it sensitive to taking subsets?
Let us see.

```{r letse, fig.cap = "The period 1861-1890, Bourgeouis-Pichat method by sex."}
source("R/drawBP.R")
oldpar <- par(mfrow = c(1,2))
##
r1 <- with(post[post$sex == "girl", ], eha::Surv(enter, exit, event))
r1 <- eha::risksets(r1, members = FALSE)
drawBP(r1, main = "Girls", ylim = c(0, 0.055))
##
##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
##r2 <- eha::risksets(r2, members = FALSE)
##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
##
r3 <- with(post[post$sex == "boy", ], eha::Surv(enter, exit, event))
r3 <- eha::risksets(r3, members = FALSE)
drawBP(r3, main = "Boys", ylim = c(0, 0.055))
par(oldpar)
```

In Figure \@ref(fig:letse) we notice a weak tendency towards convexity for boys.
Not a big deal.

## Exogeneous mortality on the $g$ scale

We now consider the whole infant age span (on the $g$ scale), and compare the uniform and exponential densities, especially of interest is the fraction of deaths that occur in the neonatal phase. @manfred04 argues that a less satisfactory property of the *B-P* model is that this fraction is constant, 
19.7 percent, independent of the overall level of infant mortality. This is not the case for the hazards model, see Figure \@ref(fig:exog).
For $p = 0.2$, the fraction in question is 21.5 percent.

```{r exog, fig.height = 8, fig.cap = "Densities for uniform and exponential distributions for varying death probability p."}
source("R/denhaz.R")
p <- c(0.01, 0.05, 0.1, 0.2)
oldpar <- par(mfrow = c(2, 2))
for (i in 1:4){
   rr <- denhaz(p[i], header = paste("p =", p[i]))
}
par(oldpar)
```

## Endogeneous mortality on the $g$ scale

In order to calculate the *endogeneous* infant mortality, a simple three-step procedure leads
to the goal.

1.   Estimate the post-neonatal mortality following the exponential route. It reduces to a
simple occurrence/exposure calculation: The total number of post-neonatal deaths $D$ is divided by the
total postneonatal exposure $E$ on the $g$ scale. So
\begin{equation*}
\hat{\lambda} = \frac{D}{E}
\end{equation*}
is the estimated hazard function (constant), and the cumulative hazard function is
\begin{equation*}
\hat{H}(t) = \hat{\lambda} t, \; t > 0
\end{equation*}

2.   Translate the result from 1. to the natural time scale.
We have found that the exogenous mortality distribution on the $g$ scale
has a constant hazard function. Now we want to translate this back to the original, natural time scale.
Apart from obvious reasons, we want to calculate the endogenous mortality by estimating 
the total mortality in the usual way [@nelson72; @aalen78] and subtracting the exogenous mortality, 
all takes of course place on the neonatal part on the natural time scale.
First, it can be established that the relations between the natural and the $g$ time scales
regarding hazard and cumulative hazard functions are given by Equations \@ref(eq:haz) and \@ref(eq:cumhaz).
\begin{align} (\#eq:haz)
h(t) &= \lambda \frac{d}{dt} g(t), \quad 0 \le t \le 365, \\
(\#eq:cumhaz) H(t) &= \lambda g(t), \quad 0 \le t \le 365, 
\end{align}
where $\lambda$ is the constant exogenous hazard on the $g$ scale.
See Figure \@ref(fig:hH).

```{r hH, fig.cap = "Exogenous infant mortality, hazard and cumulative hazard functions."}
lam <- 0.0002 ##rate
oldpar <- par(mfrow = c(1, 2), cex.main = 0.8)
x <- seq(0, 365, length = 1000)
plot(x, lam * dg(x) * 1000, type = "l", main = "Hazard function", col = "blue", 
     axes = FALSE, xlab = "Days", ylab = "per 1000")
axis(1, at = c(0, 30, 183, 365))
axis(2, las = 1)
abline(h = 0, v = 0)
abline(v = 30, lty = 3, col = "darkgreen")
box()
plot(x, lam * g(x) * 1000, type = "l", main = "Cumulative hazard function", col = "blue",
     axes = FALSE, xlab = "Days", ylab = "per 1000")
axis(1, at = c(0, 30, 183, 365))
axis(2, las = 1)
abline(h = 0, v = 0)
abline(v = 30, lty = 3, col = "darkgreen")
box()
par(oldpar)
```

3.   On the full infant interval, estimate the total cumulative hazard rate $A(t)$ with the usual 
Nelson-Aalen estimator $\hat{A}(t)$ [@nelson72; @aalen78]. Then, subtract $H(t)$ from 
$\hat{A}(t)$ to get $\hat{E}(t)$.
\begin{equation}
\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t, 0 < t < g(30).
\end{equation}
Note that subtraction and addition of competing risks are okay on the hazards scale, 
but not with probabilities, which is yet another argument in favor of the hazards approach.

Let us do it with the given data, and the first period 1861--1890.

### Step 1: Estimate post-neonatal (exogenous) mortality on the $g$ scale

Post-neonatal mortality is the same as exogenous mortality in the sense that the
hazard functions are the same on the post-neonatal age interval. So the first step
involves data left truncated at age 30 days (or at `r round(g(30))`) on the $g$ scale.
We exemplify the numerical procedure with the aid of the **eha** [@eha; @ehar2].
The first lines of the data set are (Table \@ref(tab:firstlines))
```{r firstlines, results='asis'}

tbl(infant[1:5, ], caption = "First five rows of data frame.")
nro <- nrow(infant)
deat <- as.integer(sum(infant$event))
```
and the number of rows is `r nro`, and the number of deaths is `r deat`,
resulting in a crude death probability of `r round(1000 * deat / nro)` per thousand
live births.

So, our first step is to truncate and rescale. We show how it is done in **R**:

```{r retrunc, echo = TRUE}
postneo <- eha::age.window(infant, c(30, 365))
gpostneo <- eha::age.window(ginfant, c(g(30), g(365)))
print(postneo[1:5, ], row.names = FALSE) # print first five rows
print(gpostneo[1:5, ], row.names = FALSE) # first five rows after...
```

Now, the post-neonatal hazard function is *constant*, and its ML estimator is simply 
the *occurrence/exposure* rate, or in **R**,

```{r occexp, echo = TRUE}
D <- sum(gpostneo$event) # No. of deaths
E <- with(gpostneo, sum(exit - enter)) # exposure
(rate <- D/E)
```

The rate is a very small number as a consequence of the very small time unit 
(implying large total exposure time), originally *day*.

### Transform back to the natural time scale

Use Equations \@ref(eq:haz) and \@ref(eq:cumhaz) to calculate the exogenous mortality distribution on 
the natural time scale.

```{r back}
H <- function(x, lambda = rate){
   lambda * g(x)
}

S <- function(x, lambda = rate){
   exp(-H(x, lambda))
}

getP <- function(lambda){
   days <- c(0, 0.25, 1:365)
   pd <- S(days, lambda)
   -diff(pd)
}
```

### Step 3: Estimate total and endogenous mortality

Now, we are primarily interested in the neonatal period, so let us zoom in on the 
neonatal age interval, Figure \@ref(fig:mag).

```{r mag, fig.cap = "Exogenous neonatal mortality, hazard and cumulative hazard functions."}
lam <- rate
oldpar <- par(mfrow = c(1, 2), cex.main = 0.9)
x <- seq(0, 30, length = 500)
plot(x, lam * dg(x) * 1000, type = "l", main = "Hazard function", col = "blue", 
     axes = FALSE, xlab = "Days", ylab = "per 1000")
axis(1, at = c(0, 15, 30))
axis(2, las = 1)
abline(h = 0, v = 0)
abline(v = 30, lty = 3, col = "darkgreen")
box()
plot(x, lam * g(x) * 1000, type = "l", main = "Cumulative hazard function", col = "blue",
     axes = FALSE, xlab = "Days", ylab = "per 1000")
axis(1, at = c(0, 15, 30))
axis(2, las = 1)
abline(h = 0, v = 0)
abline(v = 30, lty = 3, col = "darkgreen")
box()
par(oldpar)
```

The estimation of the cumulative hazard function for the total neonatal mortality is
easily achieved by calling the function `hazards` in the *eha* package, see also Figure \@ref(fig:hazards).

```{r hazards, fig.cap = "Cumulative hazard functions for neonatal mortality.", echo = TRUE}
neo <- age.window(infant, c(0, 30))
par(las = 1)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = neo)
hneo <- hazards(fit, cum = TRUE)
x <- c(0, hneo[[1]][, 1])
y <- c(0, 1000 * cumsum(hneo[[1]][, 2]))
plot(x, y, type = "l", xlab = "Day", ylab = "by 1000")
lines(x, 1000 * lam * g(x), col = "blue", lty = 2)
abline(h = 0, v = 0)
text(8, 15, "Total", col = "black")
text(8, 5, "Exogenous", col = "blue")
```

Next, take the difference between "Total" and "Exogenous" to get "Endogenous", Figure \@ref(fig:enof).

```{r enof, fig.cap = "Cumulative hazard function for endogenous neonatal mortality.", fig.width = 7}
yen <- y - 1000 * lam * g(x)
par(las = 1)
plot(x, yen, type = "l", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 25.4))
abline(h = 0, v = 0)
lines(x, 1000 * lam * g(x), col = "blue", lty = 3)
lines(x, y, lty = 3)
text(8, 15, "Total", cex = 0.8)
text(15, 14, "Endogenous", col = "red")
text(8, 5, "Exogenous", col = "blue", cex = 0.8)
n <- length(yen)
axis(4, at = c(0, 1000 * lam * g(x[n]), yen[n], y[n]), las = 1, cex.axis = 0.7)
```

# The hazards approach in practice

## Västerbotten 1921--1950

```{r lowume}
source("R/biometric.R")
ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
ume <- ume[ume$birthdate > "1920-12-31", ]
crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
crudprob <- round((1 - exp(-crudimr)) * 1000)
```

The crude IMR in Västerbotten 1921--1950 was around `r crudprob` per thousand, a rather 
low figure in context. Let us perform the biometric analysis with these data, see 
Figure \@ref(fig:lowumee).

```{r lowumee, fig.cap = "Västerbotten 1921--1950."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(30), g(365)))
##rs.post <- risksets()
oldpar <- par(mfrow = c(1, 2))
biometric(ume, graph = TRUE, C = 365 / (log(366))^3)
##
par(oldpar)
```

There is a good model fit, and we see that the endogenous mortality clearly dominates 
the early days of neonatal mortality, and almost vanishes towardsthe start of the post-neonatal period.

## Västerbotten 1890--1920

```{r medume}
source("R/biometric.R")
ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
ume <- ume[(ume$birthdate > "1890-12-31") & (ume$birthdate <= "1920-12-31"), ]
crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
crudprob <- round((1 - exp(-crudimr)) * 1000)
```

The crude IMR in Umeå 1891--1920 was around `r crudprob` per thousand, clearly higher than 
the later time period. Let us perform the biometric analysis with these data, see 
Figure \@ref(fig:bioume9120).

```{r bioume9120, fig.cap = "Västerbotten 1891--1920."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(30), g(365)))
##rs.post <- risksets()
oldpar <- par(mfrow = c(1, 2))
biometric(ume, graph = TRUE, C = 365 / (log(366))^3)
##
par(oldpar)
```

The dominance of endogenous mortality in early life is still clear.

## Västerbotten 1861--1890

```{r highume}
source("R/biometric.R")
ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
ume <- ume[(ume$birthdate <= "1890-12-31"), ]
crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
crudprob <- round((1 - exp(-crudimr)) * 1000)
```

The crude IMR in Umeå 1861--1890 was around `r crudprob` per thousand, highest of the three
time periods.
The biometric analysis is shown in 
Figure \@ref(fig:bioume6190).

```{r bioume6190, fig.cap = "Västerbotten 1861--1890."}
##source("R/compare.R")
##xx <- bio(ume, timescale = "day", graph = TRUE)
##post <- age.window(ume, c(g(30), g(365)))
##rs.post <- risksets()
oldpar <- par(mfrow = c(1, 2))
biometric(ume, graph = TRUE, C = 365 / (log(366))^3)
##
par(oldpar)
```

The conclusion here is almost the same as  for the later time periods, good model
fit and endogenous dominance in the very early days of life.

# Detecting under-registration

## Scania 1710--1800.

This is an example of under-registration of early deaths.

```{r scahigh, fig.cap = "Scania 1710--1800."}
sca1 <- readRDS("Rdata/sca1.rds")
sca1$exit <- sca1$exit * 366 / (log(366))^3
sca1$enter <- sca1$enter * 366 / (log(366))^3
oldpar <- par(mfrow = c(1, 2))
biometric(sca1, graph = TRUE)
par(oldpar)
```

The estimation for the neonatal period breaks down totally, see Figure \@ref(fig:scahigh).
A sign of severe under-registration of neonatal deaths.

## Skellefteå 1820--1835

Another example of severe under-registration, see Figure \@ref(fig:skum).

```{r skum, fig.cap = "Skellefteå 1821--1838."}
library(skum)
births <- obs[obs$starttyp == 2 &
                 obs$foddat > as.Date("1820-12-31") &
                 obs$foddat <= as.Date("1838-12-31"), ]
births$event <- births$sluttyp == 2
addup <- births$enter <= 0
births$enter[addup] <- births$enter[addup] + 0.0001
infa <- age.window(births, c(0, 1))
biometric(infa, timescale = "year", graph = TRUE)
```

# Known causes of death

In our data set U22003 we have some causes of death, distributed as in Table \@ref(tab:dcause).

```{r dcause, results='asis'}
deaths <- infant[infant$event == 1, ]
xx <- with(deaths, table(icd.chapter))
tt <- cbind(names(xx), as.vector(xx))
tbl(tt, rownames = FALSE, caption = "ICD-10 Category, frequencies.")
```

## Cumulative incidence analysis

Assuming that chapters XV, XVI, and XVII are endogenous and the rest is exogenous. 
Except for "missing value", which is a separate category, named "not available"
in Figure \@ref(fig:cuminc). This is a *competing risks analysis* [@ccr01].
The R package *cmprsk* [@cmprsk] was used to produce  Figure \@ref(fig:cuminc).

```{r cuminc, fig.cap = "Cumulative incidence functions, Västerbotten 1861-1950.", fig.height = 9}
par(mfrow = c(2, 2), lwd = 2)
library(survival)
infant$fstatus <- 0
endogen <- c("C", "D", "E", "G", "H", "O", "P", "Q")
endo <- (infant$icd.chapter %in% endogen) & (infant$event == 1)
exo <- (!(infant$icd.chapter %in% endogen)) & (infant$event == 1) &
   (!is.na(infant$icd.chapter))
unknown <- (is.na(infant$icd.chapter)) & (infant$event == 1)
infant$fstatus[endo] <- 1
infant$fstatus[exo] <- 2
infant$fstatus[unknown] <- 3
infant$fstatus <- factor(infant$fstatus, labels = c("censor", "endogenous", "exogenous", "not reported"))

cols <- c("magenta", "blue", "black")
att <- c(0, g(30), 365)
labs <- c("0", "g(30)", "365")
ltys <- c(1,2,3)
## All
fit <- survfit(Surv(exit, fstatus) ~ 1, data = infant)
plot(fit, xlab = "g(Day)", col = cols, axes = FALSE, main = "1861-1950", lty = ltys, lwd = 2)
axis(1, at = att, labels = labs)
axis(2, las = 1)
box()
legend("topleft", legend = c("Endogenous", "Exogenous", "Not reported"), col = cols, lty = ltys)
abline(h = 0, lwd = 1)
## 1861-1890
fit <- survfit(Surv(exit, fstatus) ~ 1, data = infant[infant$period == "1861-1890", ])
plot(fit, xlab = "g(Day)", col = cols, axes = FALSE, main = "1861-1890", lty = ltys, lwd = 2)
axis(1, at = att, labels = labs)
axis(2, las = 1)
box()
legend("topleft", legend = c("Endogenous", "Exogenous", "Not reported"), col = cols, lty = ltys)
abline(h = 0, lwd = 1)
## 1891-1920
fit <- survfit(Surv(exit, fstatus) ~ 1, data = infant[infant$period == "1891-1920", ])
plot(fit, xlab = "g(Day)", col = cols, axes = FALSE, main = "1891-1920", lty = ltys, lwd = 2)
axis(1, at = att, labels = labs)
axis(2, las = 1)
box()
legend("topleft", legend = c("Endogenous", "Exogenous", "Not reported"), col = cols, lty = ltys)
abline(h = 0, lwd = 1)
## 1921-1950
fit <- survfit(Surv(exit, fstatus) ~ 1, data = infant[infant$period == "1921-1950", ])
plot(fit, xlab = "g(Day)", col = cols, axes = FALSE, main = "1921-1950", lty = ltys, lwd = 2)
axis(1, at = att, labels = labs)
axis(2, las = 1)
box()
legend("topleft", legend = c("Endogenous", "Exogenous", "Not reported"), col = cols, lty = ltys)
abline(h = 0, lwd = 1)
##
par(oldpar)
```

# Conclusion

# References

### 2023-03-06 bb2.Rmd
first commit

---
title: "Old Scania Data"
author: "GB & TB"
date: '`r Sys.time()`'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: ../Lund2205/lund.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 4)
library(eha)
source("R/tbl.R")
```

# Introduction

An interesting paper is @lygrab98.

# Read data

```{r readdata}
sca <- read.table("Data/Copy of test_tommy 2.csv", header = TRUE, sep = ";",
                  fileEncoding = "latin1")
sca$EventDate = as.Date(sca$EventDate)
sca$BirthDate = as.Date(sca$BirthDate)
sca$DeathDate <- as.Date(sca$DeathDate)
sca$LastDate <- as.Date(sca$LastDate)
sca$SexID = NULL
sca$Sex <- factor(sca$Sex)
sca$wd <- weekdays(sca$BirthDate)
sca$death_wd <- weekdays(sca$DeathDate)
sca$year <- as.numeric(format(sca$BirthDate, format = "%Y"))
n1 <- NROW(sca)
cat("No. o f rows = ", n1, "\n")
```

# Calendar time restrictions

```{r timeres}
options(warn = -1)
hog <- sca[sca$ParishName == "Hög", ]
halm <- sca[sca$ParishName == "Halmstad", ]
kag <- sca[sca$ParishName == "Kågeröd", ]
kav <- sca[sca$ParishName == "Kävlinge", ]
sir <- sca[sca$ParishName == "Sireköpinge", ]
##
hog = hog[hog$year >= 1685, ]
halm <- halm[halm$year >= 1710, ]
kag <- kag[kag$year >= 1728, ]
kav <- kav[kav$year >= 1685, ]
sir <- sir[sir$year >=  1710, ]
##
sca <- rbind(hog, halm, kag, kav, sir)
sca$LastEvent <- factor(sca$LastEvent)
xxx <- with(sca, table(wd, ParishName))
n2 <- NROW(sca)
cat("No. of rows left = ", n2, ", removed", n1-n2, "\n")
tbl(xxx, caption = "Week day by Parish.", rownames = TRUE)
options(warn = 0)
```

# Last date

```{r lastdate}
weq <- with(sca, LastEvent %in% c("Född", "Hushåll In", "Inflytt", "vigd"))
sca <- sca[!weq, ]
sca <- sca[sca$Sex != "Unknown", ]
sca$Sex <- factor(sca$Sex)
n3 <- NROW(sca)
cat("No. of rows left = ", n3, ", removed another ", n2-n3, "\n")
```

# Sex ratio

See Table \@ref(tab:sexrat).

```{r sexrat}
xx <- with(sca, table(Sex, ParishName))
rr <- xx[2, ]  /  xx[1, ]
sex.ratio <- round(rr, 2)
tbl(sex.ratio, caption = "Sex ratio by Parish", rownames = TRUE)
```

# Survival data and cohort selection

We select a cohort consisting of all born later than 31 December 1709.

```{r surv}
sca$exit <- as.numeric(with(sca, LastDate - BirthDate))
sca$enter <- 0
sca$exit[sca$exit < 0.5] <- 0.25
sca$event <- !is.na(sca$DeathDate)
sca <- age.window(sca, c(0, 365))
## Temporarily change time scale to year:           NOTE!!!!
##sca$exit <- sca$exit / 365      NOTE: KEEP DAYS AS TIME SCALE!!
sca$bdate <- toTime(sca$BirthDate)
## We do not want to cut spells at calendar time limits:
##sca <- cal.window(sca, c(1710, 1812), surv = c("enter", "exit", "event", "bdate"))
sca <- sca[sca$year >= 1710, ] # Cohort selection
sca$Parish <- factor(sca$ParishName)
n4 <- NROW(sca)
cat("No. of rows left = ", n4, ", removed another ", n3-n4, "\n")
```

# Save data

```{r savedata, echo = TRUE}
sca$priest <- sca$Parish
levels(sca$priest) <- c("HalmSir", "HogKav", "Kage", "HogKav", "HalmSir")
saveRDS(sca, file = "Rdata/sca.rds")
```

# Analysis

See Table \@ref(tab:prelanal) and Figure \@ref(fig:anfig). For crude period statistics, see Figure \@ref(fig:period).

```{r prelanal, results = 'asis'}
fit <- coxreg(Surv(exit, event) ~ Sex + I(year - 1770) + Parish, data = sca)
ltx2(fit, caption = "Cox regression, full period.", label = "tab:prelanal")
```

```{r anfig, fig.cap = "Cumulative hazard function at baseline, boys and Halmstad 1770, model based."}
plot(fit, col = "blue", xlab = "Age (years)")
```

```{r period, fig.cap = "Crude IMR by time period start."}
cuts <- c(1710, 1725, 1750, 1775, 1800, 1812)
yy <- with(sca, cbind(enter/365, exit/365, event, toTime(BirthDate))) # NOTE: /365
xx <- perstat(yy, cuts, age = c(0, 1))
plot(cuts[-length(cuts)], xx$intensity, type = "b", ylim = c(0, 0.3), xlim = c(1710, 1811),
     xlab = "Time period", ylab = "IMR", axes = FALSE, col = "blue")
axis(1, at = cuts)
axis(2, las = 1)
box()
```

# Bourgeois-Pichat and BB

Time scale days and the log-cube transform. See Figure \@ref{fig:bpscale} for a comparison. Not very linear 
in either case.

```{r bpscale, fig.cap = "Bourgeois-Pichat in red, BB in blue."}
## sca$exit <- sca$exit * 365 # ALREADY "DAY SCALE"!
post <- sca
##post <- age.window(sca, c(29, 365))
#post$enter <- post$enter - 0.01
#post$exit = post$exit - post$enter
g <- function(x) (log(1 + x))^3
post <- post[post$year >= 1765, ]
post$exit <- g(post$exit)
post$enter <- g(post$enter)
gsurv <- with(post, Surv(enter, exit, event))
rs <- risksets(gsurv, members = FALSE)
bp <- cumsum(rs$n.events) / rs$size[1]
plot(rs$risktimes, bp, type = "l", ylim = c(0, 0.30), ##xlim = c(g(29), g(365)),
     xlab = "g(Age)", ylab = "", col = "red")
bb <- cumsum(rs$n.events) / rs$size
lines(rs$risktimes, bb, col = "blue", lty = 1)
n <- length(bb)
lines(rs$risktimes[c(29, n)], bb[c(29, n)], lty = 2, col = "blue")
lines(rs$risktimes[c(29, n)], bp[c(29, n)], lty = 2, col = "red")
```

# Conditional analysis

Condition on survival to 28:th day.

```{r cond}
p2 <- age.window(post, c(g(28), g(365)))
```

And BP vs. BB.

```{r bpbbcond, fig.cap = "Bourgeois-Pichat in red, BB in blue, conditional."}
gsurv <- with(p2, Surv(enter, exit, event))
rs <- risksets(gsurv, members = FALSE)
bp <- cumsum(rs$n.events) / rs$size[1]
plot(rs$risktimes, bp, type = "l", ylim = c(0, 0.30),
     xlab = "g(Age)", ylab = "", col = "red")
bb <- cumsum(rs$n.events) / rs$size
lines(rs$risktimes, bb, col = "blue", lty = 1)
n <- length(bb)
lines(rs$risktimes[c(1, n)], bb[c(1, n)], lty = 2, col = "blue")
lines(rs$risktimes[c(1, n)], bp[c(1, n)], lty = 2, col = "red")
```

## Gompertz

```{r gompertz, fig.cap = "Gompertz fit."}
fit.g <- phreg(Surv(exit - enter, event) ~ 1, data = p2, dist = "gompertz", param = "rate")
fit.c <- coxreg(Surv(exit - enter, event) ~ 1, data = p2)
check.dist(fit.g, fit.c)
```

Good fit, see Figure \@ref(fig:gompertz)!

## Weibull

```{r weibull, fig.cap = "Weibull fit."}
fit.w <- phreg(Surv(exit - enter, event) ~ 1, data = p2)
fit.c <- coxreg(Surv(exit - enter, event) ~ 1, data = p2)
check.dist(fit.w, fit.c)
```

Bad fit, see Figure \@ref(fig:weibull).

### 2023-03-06 desc.Rmd
first commit

---
title: "Old Scania data, summary statistics"
author: "GB & TB"
date: '`r Sys.time()`'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: ../Lund2205/lund.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 8)
library(eha)
source("R/tbl.R")
sca <- readRDS("Rdata/sca.rds")
```

# Introduction

Some statistics for Scania data and

# Cumulative hazards by parish

First the full period, Figure \@ref(fig:totcum).

```{r totcum, fig.cap = "Cumulative hazards by parish, full period."}
par(mfrow = c(3, 2))
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca[sca$Parish == "Halmstad",])
plot(fit, main = "Halmstad", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca[sca$Parish == "Hög",])
plot(fit, main = "Hög", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca[sca$Parish == "Kågeröd",])
plot(fit, main = "Kågeröd", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca[sca$Parish == "Kävlinge",])
plot(fit, main = "Kävlinge", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca[sca$Parish == "Sireköpinge",])
plot(fit, main = "Sireköpinge", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca)
plot(fit, main = "Alla", printLegend = "bottomright")
```

## Period 1710--1765

This period is depicted in Figure \@ref(fig:firstcum).

```{r firstcum, fig.cap = "Cumulative hazards by parish, 1710--1765."}
par(mfrow = c(3, 2))
sca1 <- sca[sca$year <= 1765, ]
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca1[sca1$Parish == "Halmstad",])
plot(fit, main = "Halmstad", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca1[sca1$Parish == "Hög",])
plot(fit, main = "Hög", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca1[sca1$Parish == "Kågeröd",])
plot(fit, main = "Kågeröd", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca1[sca1$Parish == "Kävlinge",])
plot(fit, main = "Kävlinge", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca1[sca1$Parish == "Sireköpinge",])
plot(fit, main = "Sireköpinge", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca1)
plot(fit, main = "Alla", printLegend = "bottomright")
```

## Period 1766--1811

This period is depicted in Figure \@ref(fig:lastcum).

```{r lastcum, fig.cap = "Cumulative hazards by parish, 1766--1811."}
par(mfrow = c(3, 2))
sca2 <- sca[sca$year > 1766, ]
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca2[sca2$Parish == "Halmstad",])
plot(fit, main = "Halmstad", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca2[sca2$Parish == "Hög",])
plot(fit, main = "Hög", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca2[sca2$Parish == "Kågeröd",])
plot(fit, main = "Kågeröd", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca2[sca2$Parish == "Kävlinge",])
plot(fit, main = "Kävlinge", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca2[sca2$Parish == "Sireköpinge",])
plot(fit, main = "Sireköpinge", printLegend = "bottomright")
fit <- coxreg(Surv(enter, exit, event) ~ strata(Sex), data = sca2)
plot(fit, main = "Alla", printLegend = "bottomright")
par(mfrow = c(1,1))
```

# Percentage deaths in the first month

```{r propd}
deaths <- sca[sca$event == 1, c("Parish", "exit")]
x <- round(prop.table(table(deaths$exit > 28)) * 100, 1)
names(x) = c("1st month", "later")
tbl(x, caption = "Percentage deaths first month, all parishes.")
```

# BP and BB

```{r bpbb, fig.cap = "Bourgeois-Pichat (BP) and Cumulative hazards (BB), full period."}
g <- function(x) {(log(1 + x))^3}
ginv <- function(x) {exp(x^(1/3)) - 1}
sca1 <- sca ## Future use!
sca$exit <- g(sca$exit) # NOTE!!
sca$enter <- g(sca$enter) # NOTE!!
sur <- with(sca, Surv(enter, exit, event))
source("../Lund2205/R/bbbp.R")
bbbp(sur)
```

## 1710--1765

```{r bpbb1, fig.cap = "Bourgeois-Pichat (BP) and Cumulative hazards (BB), born 1710--1765."}
sur <- with(sca[sca$year <= 1765, ], Surv(enter, exit, event))
bbbp(sur)
```

## 1766--1811

```{r bpbb2, fig.cap = "Bourgeois-Pichat (BP) and Cumulative hazards (BB), born 1766--1811."}
sur <- with(sca[sca$year > 1765, ], Surv(enter, exit, event))
bbbp(sur)
```

# Father's occupation (standard)

See Table \@ref(tab:occ).

```{r occ}
xx <- with(sca1, table(FatherOccupationStandard, event))
##tbl(xx, caption = "Father's occupation by survival.", rownames = TRUE)
yrk2 <- read.table("../Lund2205/yrkenklass2 utf.csv", header = TRUE, sep = ";", fileEncoding = "UTF-8")
yy <- cbind(rownames(xx), yrk2[, 2])
indx <- match(sca1$FatherOccupationStandard, yy[, 1])
sca1$socst <- yy[indx, 2]
sca1$socst <- as.factor(sca1$socst)
xy <- with(sca1, table(socst, event))

tbl(xy, caption = "socst by death, full period.", rownames = TRUE)
saveRDS(sca1, file = "Rdata/sca1.rds")
```

### 2023-03-06 getdata_bio.Rmd
first commit

---
title: "Create survival data frame for BP"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
#bibliography: snus2.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \usepackage{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
- \subsectionfont{\normalfont\itshape}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
requireNamespace("eha")
```

# Introduction

We are reading data from *Uttag U22003* from Cedar, containing infant mortality 
data from Västerbotten. The original is stored in *RDS* format in the folder *Data*, 
two files *individ* and *boende*.

The goal is th create one file suitable for *survival analysis*, meaning that the
building blocks for *episodes* are there in the form of three variables named *enter*
(start age of episode), *exit* (end age of episode), and *event* (indicator of death).
Key covariates are further *birthdate* and *place of birth*.

*Revision December 5, 2022:* Only infants with known birthdate to start with. Infants with incomplete death date are includeed as survivors, provided that it can be established that the 
infant survived one year of age.

# The individual file

Let's call it *infant*.

*NOTE:* Skip this and concentrate on *boende* for intervals.

We first separate the data into *survivors* **(is.na(deathdate))** and
*deaths* **(!is.na(deathdate))**, ad then we deal with the two groups separately.

Finally, we merge the two sets.

*END skip*

```{r read, individ}
ind <- readRDS("../Bio/Data/individ.rds")
infant <- data.frame(region = ind$REGPREFIX, 
                     fodfrs = ind$FODHEMFRS,
                     id = ind$UTTAGSID, 
                     sex = ind$KON, 
                     birthdate = ind$FODDAT_U, 
                     bdupd = ind$FODDAT_IND, 
                     bapdate = ind$DOPDAT_U, 
                     bapupd = ind$DOPDAT_IND, 
                     ab = ind$AB,
                     fb = ind$FB,
                     deathdate = ind$DODDAT_U, 
                     ddupd = ind$DODDAT_IND, 
                     cause1 = substr(ind$ICD10H_1, 1, 1), ## NOTE: Added 2023-02-09
                     cause2 = substr(ind$ICD10H_2, 1, 1), ## NOTE: Added 2023-02-09
                     cause3 = substr(ind$ICD10H_3, 1, 1), ## NOTE: Added 2023-02-09
                     cause4 = substr(ind$ICD10H_4, 1, 1), ## NOTE: Added 2023-02-09
                     cause5 = substr(ind$ICD10H_5, 1, 1), ## NOTE: Added 2023-02-09
                     cause6 = substr(ind$ICD10H_6, 1, 1), ## NOTE: Added 2022-02-09
                     outdate = ind$BOSDAT_U,
                     odupd = ind$BOSDAT_IND,
                     uttyp = ind$BOSTYP, 
                     parish = ind$FODHEMFRS,
                     m_parish = ind$MORBOORTKOD, 
                     hisco = ind$FAR_HISCOKOD_1, 
                     cause = ind$DODORSSTD_1,
                     icd10h = ind$ICD10H_1)
inf <- infant # Save a copy of the original.
infant <- infant[infant$bdupd < 0.5, ] # Known birthdate.
infant <- infant[infant$sex %in% 1:2, ] # Known sex
infant$sex <- factor(infant$sex, labels = c("boy", "girl"))
#
infant$birthdate <- as.Date(as.character(infant$birthdate), 
                            format = "%Y%m%d")
infant$bapdate <- as.Date(as.character(infant$bapdate), 
                            format = "%Y%m%d")
infant$deathdate <- as.Date(as.character(infant$deathdate), 
                            format = "%Y%m%d")
infant$outdate <- as.Date(as.character(infant$outdate), 
                          format = "%Y%m%d")
survs <- infant[is.na(infant$deathdate), ]
deaths <- infant[!is.na(infant$deathdate), ]
```

## Survivors

"Survivors" means here individuals with unknown death date.

We need to find a date *last seen* and the corresponding *exposure time*.

```{r survivors}
exp.time <- with(survs, as.numeric(outdate - birthdate))

survs$exit <- pmin(exp.time, 365) # Note 'p'!
survs <- survs[survs$exit > 0.5, ]
survs$event <- 0
survs$enter <- 0
##summary(exp.time)
```

## Deaths

```{r deaths}
exp.time <- with(deaths, as.numeric(deathdate - birthdate))
deaths$exit <- exp.time
deaths <- deaths[deaths$exit >= -0.0001, ] # Safety?!.....
deaths$exit[deaths$exit <= 0.01] <- 0.25 # Dies on birth day: Six hours of life...
deaths <- deaths[(deaths$ddupd < 0.5) | (deaths$exit > 365), ]
deaths$event <- 1
deaths$enter <- 0
deaths <- eha::age.window(deaths, c(0, 365))
```

## Put together and save

```{r together}
infant <- rbind(deaths, survs)
infant <- infant[order(infant$birthdate), ]
infant$icd.chapter <- substr(infant$icd10h, 1, 1)
is.na(infant$icd.chapter) <- infant$icd.chapter == ""
infant$icd.chapter <- factor(infant$icd.chapter)
infant <- infant[infant$birthdate > "1860-12-31", ]
infant <- infant[infant$birthdate < "1951-01-01", ]
infant$socst <- (infant$hisco %/% 10000) %in% 1:4
infant$socst <- factor(infant$socst, labels = c("Other", "Elite"))
saveRDS(infant, file = "Rdata/infant.rds")
summary(infant)
```

### 2023-03-06 paper.Rmd
first commit

--- 2023-03-06 GBTB23.Rmd
@@
      col = "blue", axes = FALSE)
-att <- c(0, 30, 183, 365)
+att <- c(0, 28, 183, 365)
 axis(1, at = att, labels = att)
-axis(2, las = 1, at = g(att), labels = c("0", "g(30)", "g(183)", "365"))
+axis(2, las = 1, at = g(att), labels = c("0", "g(28)", "g(183)", "365"))
 box()
@@
 abline(v = c(0, 365), h = c(0, 365))
-abline(v = 30, h = g(30), col = "darkgreen", lty = 3)
+abline(v = 28, h = g(28), col = "darkgreen", lty = 3)
 ```
@@
 Assume that a cohort of infants is followed over time from birth to age one. There are no drop-outs 
-(no right censoring). The exact age in days at each observed death is noted and transformed by $g$, 
+(no right censoring except at age 365 days). The exact age in days at each observed death is noted and transformed by $g$, 
 and due to imperfect time measuring there may be tied death ages. As an illustrative example, 
@@
 
-```{r first, fig.cap = "The periods 1861--1890 and 1921--1950, Bourgeois-Pichat method."}
+```{r first, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, Bourgeois-Pichat method."}
 source("R/drawBP.R")
@@
 corresponding death ages on the $g$ time scale, and as is seen, the fit to linearity
-after 30 days (72 on the $g$ scale) is excellent for both time periods.
+after 28 days (68 on the $g$ scale) is excellent for both time periods.
 
@@
 
-Instead of using the total number of births throughout in the denominator oas in  the
+Instead of using the total number of births throughout in the denominator as in  the
 Bourgeois-Pichat procedure, we suggest 
-using the present riskset size, that is, the the total number of infants still alive 
+using the present risk-set size, that is, the the total number of infants still alive 
 immediately prior to the death age under investigation. It is easily achieved by the use of 
@@
 It starts with `r as.integer(lt1[1, 3])` live births, of which `r lt1[1, 4]` dies on the day of birth, 
-on avareage at the age of six hours (a quarter of a day), and so on.
+on average at the age of six hours (a quarter of a day), and so on.
 
@@
 
-```{r second, fig.cap = "The periods 1861--1890 and 1921--1950, hazard method."}
+```{r second, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, hazards method."}
 source("R/drawBB.R")
@@
 
-```{r both, fig.cap = "Bourgeois-Pichat and 'hazard method' plots."}
-oldpar = par(mfrow = c(1, 2))
+```{r both, fig.cap = "Bourgeois-Pichat and 'hazards method' plots."}
+oldpar = par(mfrow = c(1, 2), cex = 0.7, las = 1)
 plot(r1$risktimes, cumsum(r1$n.events) / r1$size[1], type = "l", col = "blue",
@@
 only from an age later than birth). The hazards approach, on the other hand, allows
-left truncation and right censoring as long as they are *non-informative*. Often, though,
+left truncation and right censoring as long as they are *non-informative* in the usual sense. Often, though,
 this not very important, since new-born babies tend to be geographically stationary
@@
 
-## Post-neonatal mortality on the $g$ scale
+Throughout the rest of the paper, everything happens on the $g$ time scale.
 
-The following discussion assumes that age is measured on the $g$ scale.
-We concentrate on the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).
+## Post-neonatal mortality
 
-In order to see this, the data set from above is *left truncated* at age 30 days (at 72 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 30.
+ We concentrate on the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *Uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).
 
-```{r postneo, fig.cap = "Exponential fits to postneonatal data."}
-oldpar <- par(mfrow = c(1, 2))
-post <- age.window(ginfant, c(g(30), g(365)))
+In order to see this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 30.
+
+```{r postneo, fig.cap = "Exponential fits to postneonatal data, Västerbotten."}
+oldpar <- par(mfrow = c(1, 2), cex = 0.7)
+post <- age.window(ginfant, c(g(28), g(365)))
 p1 <- post[post$period == "1861-1890", ]
@@
 wfit1 <- phreg(Surv(exit - enter, event) ~ 1, data = p1, shape = 1)
-check.dist(fit1, wfit1, main = "1861-1890")
+##check.dist(fit1, wfit1, main = "1861-1890", printLegend = FALSE)
+plot(fit1, main = "1861-1890", printLegend = FALSE, xlab = "g(Day)", col = "blue")
+whaz1 <- hazards(wfit1)
+with(whaz1, lines(x, y, lty = 2, col = "red"))
+legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
+       col = c("blue", "red"), cex = 1)
 ##
@@
 wfit3 <- phreg(Surv(exit - enter, event) ~ 1, data = p3, shape = 1)
-check.dist(fit3, wfit3, main = "1921-1950")
+whaz <- hazards(wfit3)
+plot(fit3, main = "1921-1950", printLegend = FALSE, xlab = "g(Day)")
+with(whaz, lines(x, y, lty = 2, col = "red"))
+legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
+       col = c("blue", "red"), cex = 1)
 par(oldpar)
@@
 The *Exponential* fit is excellent, with a slight edge for the early period data, 
-where postneonatal mortality is high. However, one question pops up: If the exponential 
+where post-neonatal mortality is high. However, one question pops up: If the exponential 
 assumption is in fact true, will it also hold for subgroups of the data?
@@
 
-```{r joined, fig.cap = "Exponential fit for two periods."}
+```{r joined, fig.cap = "Exponential fit for two periods, Västerbotten."}
+oldpar = par(cex = 0.7)
 p13 <- post[post$period != "1891-1920", ]
@@
 wfit13 <- phreg(Surv(exit - enter, event) ~ 1, data = p13, shape = 1)
-check.dist(fit13, wfit13, main = "1861-1890, 1921-1950")
+plot(fit13, main = "1861-1890 and 1921-1950", xlab = "g(Days)", printLegend = FALSE)
+whaz13 <- hazards(wfit13)
+with(whaz13, lines(x, y, lty = 2, col = "red"))
+legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
+       col = c("blue", "red"), cex = 1)
 #
+par(oldpar)
 ```
@@
 
-## Exogeneous mortality on the $g$ scale
+## Exogeneous mortality
 
@@
 19.7 percent, independent of the overall level of infant mortality. This is not the case for the hazards model, see Figure \@ref(fig:exog).
-For $p = 0.2$, the fraction in question is 21.5 percent.
+For instance, for $p = 0.2$ the fraction in question is 21.5 percent.
 
@@
 
-## Endogeneous mortality on the $g$ scale
+## Endogeneous mortality
 
-In order to calculate the *endogeneous* infant mortality, a simple three-step procedure leads
+In order to calculate the *endogeneous* infant mortality, a simple two-step procedure leads
 to the goal.
@@
 
-2.   Translate the result from 1. to the natural time scale.
-We have found that the exogenous mortality distribution on the $g$ scale
-has a constant hazard function. Now we want to translate this back to the original, natural time scale.
-Apart from obvious reasons, we want to calculate the endogenous mortality by estimating 
-the total mortality in the usual way [@nelson72; @aalen78] and subtracting the exogenous mortality, 
-all takes of course place on the neonatal part on the natural time scale.
-First, it can be established that the relations between the natural and the $g$ time scales
-regarding hazard and cumulative hazard functions are given by Equations \@ref(eq:haz) and \@ref(eq:cumhaz).
-\begin{align} (\#eq:haz)
-h(t) &= \lambda \frac{d}{dt} g(t), \quad 0 \le t \le 365, \\
-(\#eq:cumhaz) H(t) &= \lambda g(t), \quad 0 \le t \le 365, 
-\end{align}
-where $\lambda$ is the constant exogenous hazard on the $g$ scale.
-See Figure \@ref(fig:hH).
-
-```{r hH, fig.cap = "Exogenous infant mortality, hazard and cumulative hazard functions."}
-lam <- 0.0002 ##rate
-oldpar <- par(mfrow = c(1, 2), cex.main = 0.8)
-x <- seq(0, 365, length = 1000)
-plot(x, lam * dg(x) * 1000, type = "l", main = "Hazard function", col = "blue", 
-     axes = FALSE, xlab = "Days", ylab = "per 1000")
-axis(1, at = c(0, 30, 183, 365))
-axis(2, las = 1)
-abline(h = 0, v = 0)
-abline(v = 30, lty = 3, col = "darkgreen")
-box()
-plot(x, lam * g(x) * 1000, type = "l", main = "Cumulative hazard function", col = "blue",
-     axes = FALSE, xlab = "Days", ylab = "per 1000")
-axis(1, at = c(0, 30, 183, 365))
-axis(2, las = 1)
-abline(h = 0, v = 0)
-abline(v = 30, lty = 3, col = "darkgreen")
-box()
-par(oldpar)
-```
-
-3.   On the full infant interval, estimate the total cumulative hazard rate $A(t)$ with the usual 
+2.   On the full infant interval, estimate the total cumulative hazard rate $A(t)$ with the usual 
 Nelson-Aalen estimator $\hat{A}(t)$ [@nelson72; @aalen78]. Then, subtract $H(t)$ from 
@@
 \begin{equation}
-\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t, 0 < t < g(30).
+\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t, 0 < t < g(28).
 \end{equation}
@@
 
-### Step 1: Estimate post-neonatal (exogenous) mortality on the $g$ scale
+### Step 1: Estimate post-neonatal (exogenous) mortality
 
@@
 hazard functions are the same on the post-neonatal age interval. So the first step
-involves data left truncated at age 30 days (or at `r round(g(30))`) on the $g$ scale.
+involves data left truncated at age 28 days (or at `r round(g(28))`) on the $g$ scale.
 We exemplify the numerical procedure with the aid of the **eha** [@eha; @ehar2].
@@
 ```{r retrunc, echo = TRUE}
-postneo <- eha::age.window(infant, c(30, 365))
-gpostneo <- eha::age.window(ginfant, c(g(30), g(365)))
+postneo <- eha::age.window(infant, c(28, 365))
+gpostneo <- eha::age.window(ginfant, c(g(28), g(365)))
 print(postneo[1:5, ], row.names = FALSE) # print first five rows
@@
 
-### Transform back to the natural time scale
+So we are done with the simple post-neonatal period.
 
-Use Equations \@ref(eq:haz) and \@ref(eq:cumhaz) to calculate the exogenous mortality distribution on 
-the natural time scale.
+### Step 2: Estimate total and endogenous mortality
 
-```{r back}
-H <- function(x, lambda = rate){
-   lambda * g(x)
-}
-
-S <- function(x, lambda = rate){
-   exp(-H(x, lambda))
-}
-
-getP <- function(lambda){
-   days <- c(0, 0.25, 1:365)
-   pd <- S(days, lambda)
-   -diff(pd)
-}
-```
-
-### Step 3: Estimate total and endogenous mortality
-
-Now, we are primarily interested in the neonatal period, so let us zoom in on the 
-neonatal age interval, Figure \@ref(fig:mag).
-
-```{r mag, fig.cap = "Exogenous neonatal mortality, hazard and cumulative hazard functions."}
-lam <- rate
-oldpar <- par(mfrow = c(1, 2), cex.main = 0.9)
-x <- seq(0, 30, length = 500)
-plot(x, lam * dg(x) * 1000, type = "l", main = "Hazard function", col = "blue", 
-     axes = FALSE, xlab = "Days", ylab = "per 1000")
-axis(1, at = c(0, 15, 30))
-axis(2, las = 1)
-abline(h = 0, v = 0)
-abline(v = 30, lty = 3, col = "darkgreen")
-box()
-plot(x, lam * g(x) * 1000, type = "l", main = "Cumulative hazard function", col = "blue",
-     axes = FALSE, xlab = "Days", ylab = "per 1000")
-axis(1, at = c(0, 15, 30))
-axis(2, las = 1)
-abline(h = 0, v = 0)
-abline(v = 30, lty = 3, col = "darkgreen")
-box()
-par(oldpar)
-```
+Here we focus on neonatal period.
 
@@
 ```{r hazards, fig.cap = "Cumulative hazard functions for neonatal mortality.", echo = TRUE}
-neo <- age.window(infant, c(0, 30))
+neo <- age.window(ginfant, c(0, g(28)))
 par(las = 1)
@@
 plot(x, y, type = "l", xlab = "Day", ylab = "by 1000")
-lines(x, 1000 * lam * g(x), col = "blue", lty = 2)
+lines(x, 1000 * rate * x, col = "blue", lty = 2)
 abline(h = 0, v = 0)
@@
 ```{r enof, fig.cap = "Cumulative hazard function for endogenous neonatal mortality.", fig.width = 7}
-yen <- y - 1000 * lam * g(x)
+yen <- y - 1000 * rate * x
 par(las = 1)
-plot(x, yen, type = "l", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 25.4))
+plot(x, yen, type = "l", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 36))
 abline(h = 0, v = 0)
-lines(x, 1000 * lam * g(x), col = "blue", lty = 3)
-lines(x, y, lty = 3)
-text(8, 15, "Total", cex = 0.8)
+lines(x, 1000 * rate * g(x), col = "blue", lty = 3)
+lines(x, y, lty = 1)
+text(8, 20, "Total", cex = 0.8)
 text(15, 14, "Endogenous", col = "red")
@@
 n <- length(yen)
-axis(4, at = c(0, 1000 * lam * g(x[n]), yen[n], y[n]), las = 1, cex.axis = 0.7)
+axis(4, at = c(0, 1000 * rate * g(x[n]), yen[n], y[n]), las = 1, cex.axis = 0.7)
 ```
 
-# The hazards approach in practice
+# Biometric analysis in practice
+
+The original procedure of Bourgeos-Pichat is compared to the hazard based procedure for some typical
+cases from the real world.
+
+```{r plotfun}
+pf <- function(x){
+   ## x is a dataframe on g scale
+   rs <- with(x, eha::risksets(Surv(enter, exit, event), members = FALSE))
+   oldpar <- par(mfrow = c(1, 2))
+   drawBB(rs, main = "Hazards method")
+   drawBP(rs, main = "B-P method")
+##
+par(oldpar)
+}
+```
 
@@
 ```{r lowume}
-source("R/biometric.R")
+##source("R/drawBP.R")
+##source("R/drawBB.R")
 ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
@@
 ##xx <- bio(ume, timescale = "day", graph = TRUE)
-##post <- age.window(ume, c(g(30), g(365)))
+##post <- age.window(ume, c(g(28), g(365)))
 ##rs.post <- risksets()
-oldpar <- par(mfrow = c(1, 2))
-biometric(ume, graph = TRUE, C = 365 / (log(366))^3)
-##
-par(oldpar)
+ume <- ginfant[ginfant$birthdate > "1920-12-31", ]
+pf(ume)
 ```
 
-There is a good model fit, and we see that the endogenous mortality clearly dominates 
-the early days of neonatal mortality, and almost vanishes towardsthe start of the post-neonatal period.
+There is a good hazards model fit, and we see that the endogenous mortality clearly dominates 
+the early days of neonatal mortality, and almost vanishes towards the start of the post-neonatal period.
+
+The B-P model fit is slightly worse, but not much to bother about.
 
@@
 ##xx <- bio(ume, timescale = "day", graph = TRUE)
-##post <- age.window(ume, c(g(30), g(365)))
+##post <- age.window(ume, c(g(28), g(365)))
 ##rs.post <- risksets()
-oldpar <- par(mfrow = c(1, 2))
-biometric(ume, graph = TRUE, C = 365 / (log(366))^3)
-##
-par(oldpar)
+ume$exit <- g(ume$exit)
+ume$enter <- g(ume$enter)
+pf(ume)
 ```
 
-The dominance of endogenous mortality in early life is still clear.
+The dominance of endogenous mortality in early life is still clear. The hazards fit is still slightly better.
 
@@
 ##xx <- bio(ume, timescale = "day", graph = TRUE)
-##post <- age.window(ume, c(g(30), g(365)))
+##post <- age.window(ume, c(g(28), g(365)))
 ##rs.post <- risksets()
+ume$exit <- g(ume$exit)
+ume$enter <- g(ume$enter)
+pf(ume)
+```
+
+The conclusion here is almost the same as  for the later time periods, good model
+fit and endogenous dominance in the very early days of life. However, the B-P method seems to have a slight upper hand.
+
+## Scania 1710--1800.
+
+```{r scahigh, fig.cap = "Scania 1710--1800."}
+sca1 <- readRDS("Rdata/sca1.rds")
+sca1 <- sca1[sca1$BirthDate < "1801-01-01", ]
+sca1$exit <- g(sca1$exit)
+sca1$enter <- g(sca1$enter)
+rs <- with(sca1, risksets(Surv(enter, exit, event), members = FALSE))
 oldpar <- par(mfrow = c(1, 2))
-biometric(ume, graph = TRUE, C = 365 / (log(366))^3)
-##
+drawBB(rs, main = "Hazards method")
+drawBP(rs, main = "B-P method")
 par(oldpar)
@@
 
-The conclusion here is almost the same as  for the later time periods, good model
-fit and endogenous dominance in the very early days of life.
+Very bad fits in both cases.  <!--- Let us look at the estimated cumulative hazards, total, 
+exogenous and endgenous, see Figure \@ref(fig:escu).
 
-# Detecting under-registration
-
-## Scania 1710--1800.
-
-This is an example of under-registration of early deaths.
-
-```{r scahigh, fig.cap = "Scania 1710--1800."}
-sca1 <- readRDS("Rdata/sca1.rds")
-sca1$exit <- sca1$exit * 366 / (log(366))^3
-sca1$enter <- sca1$enter * 366 / (log(366))^3
-oldpar <- par(mfrow = c(1, 2))
-biometric(sca1, graph = TRUE)
-par(oldpar)
+```{r escu, fig.cap = "Scania 1710-1800.", eval = FALSE}
+#source("R/biometric.R")
+#biometric(sca1, graph = TRUE)
+post <- age.window(sca1, c(g(28), 365))
+lambda <- with(post, sum(event) / sum(exit - enter))
+x <- c(0, rs$risktimes)
+y <- x * lambda
+hazo <- with(rs, c(0, n.events / size))
+hazo <- cumsum(hazo)
+plot(x, hazo, type = "s", col = "blue")
+lines(x, y)
+z <- hazo - y
+lines(x, z, col = "red")
 ```
 
-The estimation for the neonatal period breaks down totally, see Figure \@ref(fig:scahigh).
-A sign of severe under-registration of neonatal deaths.
+--->
 
@@
 
-Another example of severe under-registration, see Figure \@ref(fig:skum).
+An example of severe under-registration during the neonatal period, see Figure \@ref(fig:skum).
+There is no room for endogeneous deaths at all, but otherwise a reasonably good fit with both 
+methods.
 
@@
 births$event <- births$sluttyp == 2
-addup <- births$enter <= 0
-births$enter[addup] <- births$enter[addup] + 0.0001
-infa <- age.window(births, c(0, 1))
-biometric(infa, timescale = "year", graph = TRUE)
-```
-
-# Known causes of death
-
-In our data set U22003 we have some causes of death, distributed as in Table \@ref(tab:dcause).
-
-```{r dcause, results='asis'}
-deaths <- infant[infant$event == 1, ]
-xx <- with(deaths, table(icd.chapter))
-tt <- cbind(names(xx), as.vector(xx))
-tbl(tt, rownames = FALSE, caption = "ICD-10 Category, frequencies.")
-```
-
-## Cumulative incidence analysis
-
-Assuming that chapters XV, XVI, and XVII are endogenous and the rest is exogenous. 
-Except for "missing value", which is a separate category, named "not available"
-in Figure \@ref(fig:cuminc). This is a *competing risks analysis* [@ccr01].
-The R package *cmprsk* [@cmprsk] was used to produce  Figure \@ref(fig:cuminc).
-
-```{r cuminc, fig.cap = "Cumulative incidence functions, Västerbotten 1861-1950.", fig.height = 9}
-par(mfrow = c(2, 2), lwd = 2)
-library(survival)
-infant$fstatus <- 0
-endogen <- c("C", "D", "E", "G", "H", "O", "P", "Q")
-endo <- (infant$icd.chapter %in% endogen) & (infant$event == 1)
-exo <- (!(infant$icd.chapter %in% endogen)) & (infant$event == 1) &
-   (!is.na(infant$icd.chapter))
-unknown <- (is.na(infant$icd.chapter)) & (infant$event == 1)
-infant$fstatus[endo] <- 1
-infant$fstatus[exo] <- 2
-infant$fstatus[unknown] <- 3
-infant$fstatus <- factor(infant$fstatus, labels = c("censor", "endogenous", "exogenous", "not reported"))
-
-cols <- c("magenta", "blue", "black")
-att <- c(0, g(30), 365)
-labs <- c("0", "g(30)", "365")
-ltys <- c(1,2,3)
-## All
-fit <- survfit(Surv(exit, fstatus) ~ 1, data = infant)
-plot(fit, xlab = "g(Day)", col = cols, axes = FALSE, main = "1861-1950", lty = ltys, lwd = 2)
-axis(1, at = att, labels = labs)
-axis(2, las = 1)
-box()
-legend("topleft", legend = c("Endogenous", "Exogenous", "Not reported"), col = cols, lty = ltys)
-abline(h = 0, lwd = 1)
-## 1861-1890
-fit <- survfit(Surv(exit, fstatus) ~ 1, data = infant[infant$period == "1861-1890", ])
-plot(fit, xlab = "g(Day)", col = cols, axes = FALSE, main = "1861-1890", lty = ltys, lwd = 2)
-axis(1, at = att, labels = labs)
-axis(2, las = 1)
-box()
-legend("topleft", legend = c("Endogenous", "Exogenous", "Not reported"), col = cols, lty = ltys)
-abline(h = 0, lwd = 1)
-## 1891-1920
-fit <- survfit(Surv(exit, fstatus) ~ 1, data = infant[infant$period == "1891-1920", ])
-plot(fit, xlab = "g(Day)", col = cols, axes = FALSE, main = "1891-1920", lty = ltys, lwd = 2)
-axis(1, at = att, labels = labs)
-axis(2, las = 1)
-box()
-legend("topleft", legend = c("Endogenous", "Exogenous", "Not reported"), col = cols, lty = ltys)
-abline(h = 0, lwd = 1)
-## 1921-1950
-fit <- survfit(Surv(exit, fstatus) ~ 1, data = infant[infant$period == "1921-1950", ])
-plot(fit, xlab = "g(Day)", col = cols, axes = FALSE, main = "1921-1950", lty = ltys, lwd = 2)
-axis(1, at = att, labels = labs)
-axis(2, las = 1)
-box()
-legend("topleft", legend = c("Endogenous", "Exogenous", "Not reported"), col = cols, lty = ltys)
-abline(h = 0, lwd = 1)
-##
-par(oldpar)
+births$enter <- 365 * births$enter
+births$exit <- 365 * births$exit
+addup <- births$exit <= 0
+births$exit[addup] <- 0.25
+infa <- age.window(births, c(0, 365))
+infa$enter <- g(infa$enter)
+infa$exit <- g(infa$exit)
+pf(infa)
 ```
@@
 
+If anything, the hazards method never performs worse than the B-P method.
+But the real strength of the hazards method is that it fits naturally into general 
+modern survival analysis with censored and truncated data, and also proportional 
+hazards models with covariates.
+
 # References

### 2023-03-06 scania.Rmd
first commit

--- 2023-03-06 paper.Rmd
@@
 ---
-title: "A hazards approach to the biometric analysis of infant mortality"
+title: "Biometric analysis of infant mortality, Scania 1710--1812"
 author: "Göran Broström and Tommy Bengtsson"
@@
 - \subsectionfont{\normalfont\itshape}
-abstract: "A variation of the Bourgeois-Pichat biometric analysis of infant mortality is suggested.
-In the original model, cumulative mortality in the last eleven months of infancy is assumed to follow a uniform distribution given a log-cube transformation of age.
-Instead, we assume an exponential distribution.
-The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk.
-We argue that our assumption is more satisfactory from a theoretical point of view, since it focus on the conditional probability of dying.
-Our model gives different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is slightly better, especially in cases with higher levels of infant mortality."
+abstract: "A variation of the Bourgeois-Pichat biometric analysis of infant mortality is studied in connection to data from Scania in the very south of Sweden."
 ---
@@
 In addition, the model ﬁt is slightly better for our model, especially in cases with high level of infant mortality, common in pre-modern societies.
-The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.1 In addition, we give examples from eighteenth Sweden, where the curve do not follow the uniform distribution during the last eleven month of infancy, despite high levels of mortality in the first month of life.
+The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.
+In addition, we give examples from eighteenth century Sweden, where the curve do not follow the uniform distribution during the last eleven month of infancy, despite high levels of mortality in the first month of life.
 
@@
 ```{r readdata}
-infant <- readRDS("Rdata/infant.rds")
-infant$period <- cut(infant$birthdate, 
-                   breaks = as.Date(c("1860-12-31", "1890-12-31", 
-                                      "1920-12-31", "1951-01-01")),
-                   labels = c("1861-1890", "1891-1920", "1921-1950"))
-infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "icd.chapter")]
+infant <- readRDS("Rdata/sca1.rds")
+infant$period <- cut(infant$BirthDate, 
+                   breaks = as.Date(c("1709-12-31", "1734-12-31", 
+                                      "1759-12-31","1784-12-31",  "1812-12-12")),
+                   labels = c("1710-1734", "1735-1759", "1760-1784", "1785-1812"))
+infant <- infant[, c("BirthDate", "Sex", "enter", "exit", "event", "period", 
+                "socst", "ICD10")]
+
 ginfant <- infant
@@
 ginfant$exit <- g(infant$exit)
-```
-For the first and last 30 years in this data set, we have the results shown in Figure \@ref(fig:first).
-
-```{r first, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, Bourgeois-Pichat method."}
-source("R/drawBP.R")
-oldpar <- par(mfrow = c(1,2))
-##
-r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
-r1 <- eha::risksets(r1, members = FALSE)
-drawBP(r1, main = "1861-1890", ylim = c(0, 0.1))
-##
-##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
-##r2 <- eha::risksets(r2, members = FALSE)
-##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
-##
-r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
-r3 <- eha::risksets(r3, members = FALSE)
-drawBP(r3, main = "1921-1950", ylim = c(0, 0.1))
-par(oldpar)
-```
-
-The cumulative numbers of death are plotted against the
-corresponding death ages on the $g$ time scale, and as is seen, the fit to linearity
-after 28 days (68 on the $g$ scale) is excellent for both time periods.
-
-# The hazard-based procedure
-
-Instead of using the total number of births throughout in the denominator as in  the
-Bourgeois-Pichat procedure, we suggest 
-using the present risk-set size, that is, the the total number of infants still alive 
-immediately prior to the death age under investigation. It is easily achieved by the use of 
-the function `risksets` in the package `eha`. See Table \@ref(tab:risksets) for 
-an illustration.
-```{r risksets, echo = FALSE, results = 'asis'}
-source("R/tbl.R")
-lt1 <- with(r1, cbind(ginv(risktimes), round(risktimes, 1), size, n.events))
-lt1 <- data.frame(lt1)
-names(lt1) <- c("Day", "g(Day)", "Population", "Deaths")
-tbl(lt1[1:7, ], caption = "Life table, 1861-1890.")
-```
-It starts with `r as.integer(lt1[1, 3])` live births, of which `r lt1[1, 4]` dies on the day of birth, 
-on average at the age of six hours (a quarter of a day), and so on.
-
-The plot corresponding to Figure \@ref(fig:first) is shown in Figure \@ref(fig:second).
-
-```{r second, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, hazards method."}
-source("R/drawBB.R")
-oldpar <- par(mfrow = c(1,2))
-##
-##r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
-##r1 <- eha::risksets(r1, members = FALSE)
-drawBB(r1, main = "1861-1890", ylim = c(0, 0.115))
-##
-##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
-##r2 <- eha::risksets(r2, members = FALSE)
-##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
-##
-##r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
-##r3 <- eha::risksets(r3, members = FALSE)
-drawBB(r3, main = "1921-1950", ylim = c(0, 0.115))
-par(oldpar)
-```
-
-See Figure \@ref(fig:both) for a comparison of the two curves.
-
-```{r both, fig.cap = "Bourgeois-Pichat and 'hazards method' plots."}
-oldpar = par(mfrow = c(1, 2), cex = 0.7, las = 1)
-plot(r1$risktimes, cumsum(r1$n.events) / r1$size[1], type = "l", col = "blue",
-     main = "1861-1890", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
-lines(r1$risktimes, cumsum(r1$n.events / r1$size), type = "l", col = "red")
-abline(h = 0, v = 0)
-text(120, 0.08, "Hazard method", col = "red")
-text(200, 0.02,"Bourgeois-Pichat", col = "blue")
-##
-plot(r3$risktimes, cumsum(r3$n.events) / r3$size[1], type = "l", col = "blue",
-     main = "1921-1950", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
-lines(r3$risktimes, cumsum(r3$n.events / r3$size), type = "l", col = "red")
-abline(h = 0, v = 0)
-text(120, 0.05, "Hazard method", col = "red")
-text(200, 0.015,"Bourgeois-Pichat", col = "blue")
-##
-##
-par(oldpar)
-```
-
-The difference between the two curves increases as infant mortality increases, 
-but both methods show an excellent fit to linearity.
-
-# Theoretical considerations
-
-We note that the Bourgeois-Pichat method requires that no right censorings (infants lost 
-to follow-up before one year of age) are present, and no left truncations (infants under observation 
-only from an age later than birth). The hazards approach, on the other hand, allows
-left truncation and right censoring as long as they are *non-informative* in the usual sense. Often, though,
-this not very important, since new-born babies tend to be geographically stationary
-for their first year of life.
-
-Throughout the rest of the paper, everything happens on the $g$ time scale.
-
-## Post-neonatal mortality
-
- We concentrate on the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *Uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).
-
-In order to see this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 30.
-
-```{r postneo, fig.cap = "Exponential fits to postneonatal data, Västerbotten."}
-oldpar <- par(mfrow = c(1, 2), cex = 0.7)
-post <- age.window(ginfant, c(g(28), g(365)))
-p1 <- post[post$period == "1861-1890", ]
-p3 <- post[post$period == "1921-1950", ]
-fit1 <- coxreg(Surv(exit - enter, event) ~ 1, data = p1)
-wfit1 <- phreg(Surv(exit - enter, event) ~ 1, data = p1, shape = 1)
-##check.dist(fit1, wfit1, main = "1861-1890", printLegend = FALSE)
-plot(fit1, main = "1861-1890", printLegend = FALSE, xlab = "g(Day)", col = "blue")
-whaz1 <- hazards(wfit1)
-with(whaz1, lines(x, y, lty = 2, col = "red"))
-legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
-       col = c("blue", "red"), cex = 1)
-##
-fit3 <- coxreg(Surv(exit - enter, event) ~ 1, data = p3)
-wfit3 <- phreg(Surv(exit - enter, event) ~ 1, data = p3, shape = 1)
-whaz <- hazards(wfit3)
-plot(fit3, main = "1921-1950", printLegend = FALSE, xlab = "g(Day)")
-with(whaz, lines(x, y, lty = 2, col = "red"))
-legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
-       col = c("blue", "red"), cex = 1)
-par(oldpar)
-```
-
-The *Exponential* fit is excellent, with a slight edge for the early period data, 
-where post-neonatal mortality is high. However, one question pops up: If the exponential 
-assumption is in fact true, will it also hold for subgroups of the data?
-
-The theoretical ansver is "No", because a mixture of two exponential populations with
-different mortality rates cannot itself be exponential. A heuristic argument for that: Assume two equally-sized 
-groups at "birth", but with different levels of mortality. As time passes, there will be more deaths 
-in the high-mortality group, and that means that the proportion survivors in the low-level group
-will increase. Therefore, the population-level mortality will seem to decrease over time, 
-and not being constant, as the exponentiality prescribes.
-
-We can see what happens when the two periods in Figure \@ref(fig:postneo) are joined, see Figure
-\@ref(fig:joined).
-
-```{r joined, fig.cap = "Exponential fit for two periods, Västerbotten."}
-oldpar = par(cex = 0.7)
-p13 <- post[post$period != "1891-1920", ]
-p13$period <- factor(p13$period)
-fit13 <- coxreg(Surv(exit - enter, event) ~ 1, data = p13)
-wfit13 <- phreg(Surv(exit - enter, event) ~ 1, data = p13, shape = 1)
-plot(fit13, main = "1861-1890 and 1921-1950", xlab = "g(Days)", printLegend = FALSE)
-whaz13 <- hazards(wfit13)
-with(whaz13, lines(x, y, lty = 2, col = "red"))
-legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
-       col = c("blue", "red"), cex = 1)
-#
-par(oldpar)
-```
-
-Not what we expected: The fit seems to be extremely good in both figures! However,
-the rate is close to zero, and so differences are small and hard to notice. And for practical use, they are not interesting.
-
-What about the original Bourgeois-Pichat model, is it sensitive to taking subsets?
-Let us see.
-
-```{r letse, fig.cap = "The period 1861-1890, Bourgeouis-Pichat method by sex."}
-source("R/drawBP.R")
-oldpar <- par(mfrow = c(1,2))
-##
-r1 <- with(post[post$sex == "girl", ], eha::Surv(enter, exit, event))
-r1 <- eha::risksets(r1, members = FALSE)
-drawBP(r1, main = "Girls", ylim = c(0, 0.055))
-##
-##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
-##r2 <- eha::risksets(r2, members = FALSE)
-##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
-##
-r3 <- with(post[post$sex == "boy", ], eha::Surv(enter, exit, event))
-r3 <- eha::risksets(r3, members = FALSE)
-drawBP(r3, main = "Boys", ylim = c(0, 0.055))
-par(oldpar)
-```
-
-In Figure \@ref(fig:letse) we notice a weak tendency towards convexity for boys.
-Not a big deal.
-
-## Exogeneous mortality
-
-We now consider the whole infant age span (on the $g$ scale), and compare the uniform and exponential densities, especially of interest is the fraction of deaths that occur in the neonatal phase.
-@manfred04 argues that a less satisfactory property of the *B-P* model is that this fraction is constant,
-19.7 percent, independent of the overall level of infant mortality. This is not the case for the hazards model, see Figure \@ref(fig:exog).
-For instance, for $p = 0.2$ the fraction in question is 21.5 percent.
-
-```{r exog, fig.height = 8, fig.cap = "Densities for uniform and exponential distributions for varying death probability p."}
-source("R/denhaz.R")
-p <- c(0.01, 0.05, 0.1, 0.2)
-oldpar <- par(mfrow = c(2, 2))
-for (i in 1:4){
-   rr <- denhaz(p[i], header = paste("p =", p[i]))
-}
-par(oldpar)
-```
-
-## Endogeneous mortality
-
-In order to calculate the *endogeneous* infant mortality, a simple two-step procedure leads
-to the goal.
-
-1.   Estimate the post-neonatal mortality following the exponential route. It reduces to a
-simple occurrence/exposure calculation: The total number of post-neonatal deaths $D$ is divided by the
-total postneonatal exposure $E$ on the $g$ scale. So
-\begin{equation*}
-\hat{\lambda} = \frac{D}{E}
-\end{equation*}
-is the estimated hazard function (constant), and the cumulative hazard function is
-\begin{equation*}
-\hat{H}(t) = \hat{\lambda} t, \; t > 0
-\end{equation*}
-
-2.   On the full infant interval, estimate the total cumulative hazard rate $A(t)$ with the usual 
-Nelson-Aalen estimator $\hat{A}(t)$ [@nelson72; @aalen78]. Then, subtract $H(t)$ from 
-$\hat{A}(t)$ to get $\hat{E}(t)$.
-\begin{equation}
-\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t, 0 < t < g(28).
-\end{equation}
-Note that subtraction and addition of competing risks are okay on the hazards scale, 
-but not with probabilities, which is yet another argument in favor of the hazards approach.
-
-Let us do it with the given data, and the first period 1861--1890.
-
-### Step 1: Estimate post-neonatal (exogenous) mortality
-
-Post-neonatal mortality is the same as exogenous mortality in the sense that the
-hazard functions are the same on the post-neonatal age interval. So the first step
-involves data left truncated at age 28 days (or at `r round(g(28))`) on the $g$ scale.
-We exemplify the numerical procedure with the aid of the **eha** [@eha; @ehar2].
-The first lines of the data set are (Table \@ref(tab:firstlines))
-```{r firstlines, results='asis'}
-
-tbl(infant[1:5, ], caption = "First five rows of data frame.")
-nro <- nrow(infant)
-deat <- as.integer(sum(infant$event))
-```
-and the number of rows is `r nro`, and the number of deaths is `r deat`,
-resulting in a crude death probability of `r round(1000 * deat / nro)` per thousand
-live births.
-
-So, our first step is to truncate and rescale. We show how it is done in **R**:
-
-```{r retrunc, echo = TRUE}
-postneo <- eha::age.window(infant, c(28, 365))
-gpostneo <- eha::age.window(ginfant, c(g(28), g(365)))
-print(postneo[1:5, ], row.names = FALSE) # print first five rows
-print(gpostneo[1:5, ], row.names = FALSE) # first five rows after...
-```
-
-Now, the post-neonatal hazard function is *constant*, and its ML estimator is simply 
-the *occurrence/exposure* rate, or in **R**,
-
-```{r occexp, echo = TRUE}
-D <- sum(gpostneo$event) # No. of deaths
-E <- with(gpostneo, sum(exit - enter)) # exposure
-(rate <- D/E)
-```
-
-The rate is a very small number as a consequence of the very small time unit 
-(implying large total exposure time), originally *day*.
-
-So we are done with the simple post-neonatal period.
-
-### Step 2: Estimate total and endogenous mortality
-
-Here we focus on neonatal period.
-
-The estimation of the cumulative hazard function for the total neonatal mortality is
-easily achieved by calling the function `hazards` in the *eha* package, see also Figure \@ref(fig:hazards).
-
-```{r hazards, fig.cap = "Cumulative hazard functions for neonatal mortality.", echo = TRUE}
-neo <- age.window(ginfant, c(0, g(28)))
-par(las = 1)
-fit <- coxreg(Surv(enter, exit, event) ~ 1, data = neo)
-hneo <- hazards(fit, cum = TRUE)
-x <- c(0, hneo[[1]][, 1])
-y <- c(0, 1000 * cumsum(hneo[[1]][, 2]))
-plot(x, y, type = "l", xlab = "Day", ylab = "by 1000")
-lines(x, 1000 * rate * x, col = "blue", lty = 2)
-abline(h = 0, v = 0)
-text(8, 15, "Total", col = "black")
-text(8, 5, "Exogenous", col = "blue")
-```
-
-Next, take the difference between "Total" and "Exogenous" to get "Endogenous", Figure \@ref(fig:enof).
-
-```{r enof, fig.cap = "Cumulative hazard function for endogenous neonatal mortality.", fig.width = 7}
-yen <- y - 1000 * rate * x
-par(las = 1)
-plot(x, yen, type = "l", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 36))
-abline(h = 0, v = 0)
-lines(x, 1000 * rate * g(x), col = "blue", lty = 3)
-lines(x, y, lty = 1)
-text(8, 20, "Total", cex = 0.8)
-text(15, 14, "Endogenous", col = "red")
-text(8, 5, "Exogenous", col = "blue", cex = 0.8)
-n <- length(yen)
-axis(4, at = c(0, 1000 * rate * g(x[n]), yen[n], y[n]), las = 1, cex.axis = 0.7)
 ```
@@
 
-## Västerbotten 1921--1950
-
-```{r lowume}
-##source("R/drawBP.R")
-##source("R/drawBB.R")
-ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
-ume <- ume[ume$birthdate > "1920-12-31", ]
-crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
-crudprob <- round((1 - exp(-crudimr)) * 1000)
-```
-
-The crude IMR in Västerbotten 1921--1950 was around `r crudprob` per thousand, a rather 
-low figure in context. Let us perform the biometric analysis with these data, see 
-Figure \@ref(fig:lowumee).
-
-```{r lowumee, fig.cap = "Västerbotten 1921--1950."}
-##source("R/compare.R")
-##xx <- bio(ume, timescale = "day", graph = TRUE)
-##post <- age.window(ume, c(g(28), g(365)))
-##rs.post <- risksets()
-ume <- ginfant[ginfant$birthdate > "1920-12-31", ]
-pf(ume)
-```
-
-There is a good hazards model fit, and we see that the endogenous mortality clearly dominates 
-the early days of neonatal mortality, and almost vanishes towards the start of the post-neonatal period.
-
-The B-P model fit is slightly worse, but not much to bother about.
-
-## Västerbotten 1890--1920
-
-```{r medume}
-source("R/biometric.R")
-ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
-ume <- ume[(ume$birthdate > "1890-12-31") & (ume$birthdate <= "1920-12-31"), ]
-crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
-crudprob <- round((1 - exp(-crudimr)) * 1000)
-```
-
-The crude IMR in Umeå 1891--1920 was around `r crudprob` per thousand, clearly higher than 
-the later time period. Let us perform the biometric analysis with these data, see 
-Figure \@ref(fig:bioume9120).
-
-```{r bioume9120, fig.cap = "Västerbotten 1891--1920."}
-##source("R/compare.R")
-##xx <- bio(ume, timescale = "day", graph = TRUE)
-##post <- age.window(ume, c(g(28), g(365)))
-##rs.post <- risksets()
-ume$exit <- g(ume$exit)
-ume$enter <- g(ume$enter)
-pf(ume)
-```
-
-The dominance of endogenous mortality in early life is still clear. The hazards fit is still slightly better.
-
-## Västerbotten 1861--1890
-
-```{r highume}
-source("R/biometric.R")
-ume <- infant##[infant$fodfrs %in% c(82750, 82780), ]
-ume <- ume[(ume$birthdate <= "1890-12-31"), ]
-crudimr <- with(ume, sum(event) / sum(exit - enter)) * 365
-crudprob <- round((1 - exp(-crudimr)) * 1000)
-```
-
-The crude IMR in Umeå 1861--1890 was around `r crudprob` per thousand, highest of the three
-time periods.
-The biometric analysis is shown in 
-Figure \@ref(fig:bioume6190).
-
-```{r bioume6190, fig.cap = "Västerbotten 1861--1890."}
-##source("R/compare.R")
-##xx <- bio(ume, timescale = "day", graph = TRUE)
-##post <- age.window(ume, c(g(28), g(365)))
-##rs.post <- risksets()
-ume$exit <- g(ume$exit)
-ume$enter <- g(ume$enter)
-pf(ume)
-```
-
-The conclusion here is almost the same as  for the later time periods, good model
-fit and endogenous dominance in the very early days of life. However, the B-P method seems to have a slight upper hand.
-
-## Scania 1710--1800.
+## Scania 1710--1820.
 
 ```{r scahigh, fig.cap = "Scania 1710--1800."}
-sca1 <- readRDS("Rdata/sca1.rds")
+source("R/drawBB.R")
+source("R/drawBP.R")
+sca1 <- infant
 sca1 <- sca1[sca1$BirthDate < "1801-01-01", ]
@@
 
-Very bad fits in both cases.  <!--- Let us look at the estimated cumulative hazards, total, 
-exogenous and endgenous, see Figure \@ref(fig:escu).
+Very bad fits in both cases. Let us look at subsets of the data.
 
-```{r escu, fig.cap = "Scania 1710-1800.", eval = FALSE}
-#source("R/biometric.R")
-#biometric(sca1, graph = TRUE)
-post <- age.window(sca1, c(g(28), 365))
-lambda <- with(post, sum(event) / sum(exit - enter))
-x <- c(0, rs$risktimes)
-y <- x * lambda
-hazo <- with(rs, c(0, n.events / size))
-hazo <- cumsum(hazo)
-plot(x, hazo, type = "s", col = "blue")
-lines(x, y)
-z <- hazo - y
-lines(x, z, col = "red")
+### Hazards by season
+
+We define *season* as four three-months periods, *winter = December + January + February*,
+*spring = March + April + May*, and so on.
+
+```{r seasons, fig.cap = "BB by birth season."}
+sca1$month <- as.numeric(format(as.Date(sca1$BirthDate, format = "%Y-%m-%d"), "%m"))
+oldpar <- par(mfrow = c(2, 2))
+sca1$Q <- ""
+sca1$Q[sca1$month %in% c(1, 2, 12)] <- "winter"
+sca1$Q[sca1$month %in% c(3, 4, 05)] <- "spring"
+sca1$Q[sca1$month %in% c(6, 7, 08)] <- "summer"
+sca1$Q[sca1$month %in% c(9, 10, 11)] <- "fall"
+sca1$Q <- as.factor(sca1$Q)
+
+for (when in levels(sca1$Q)){
+   xxx <- with(sca1[sca1$Q %in% when, ], Surv(enter, exit, event))
+   rs <- risksets(xxx, members = FALSE)
+   drawBB(rs, main = when)
+}
+par(oldpar)
 ```
 
---->
+### Hazards without high-mortality years
 
-## Skellefteå 1820--1835
+```{r highmort, fig.cap = "Years with not so high IMR."}
+infant$year <- with(infant, format(BirthDate, "%Y"))
+oldpar <- par(mfrow = c(1, 2), cex = 0.7)
+years <- sort(unique(infant$year))
 
-An example of severe under-registration during the neonatal period, see Figure \@ref(fig:skum).
-There is no room for endogeneous deaths at all, but otherwise a reasonably good fit with both 
-methods.
+n <- length(years)
+imr <- numeric(n)
+i <- 0
+for (yr in years){
+   i <- i + 1 
+   imr[i] <- with(infant[infant$year == yr, ], sum(event) / (sum(exit - enter) / 365))
+}
+plot(years, imr, type = "b", las = 1, main = "IMR")
+abline(h = 0.6, lty = 2, col = "red")
 
-```{r skum, fig.cap = "Skellefteå 1821--1838."}
-library(skum)
-births <- obs[obs$starttyp == 2 &
-                 obs$foddat > as.Date("1820-12-31") &
-                 obs$foddat <= as.Date("1838-12-31"), ]
-births$event <- births$sluttyp == 2
-births$enter <- 365 * births$enter
-births$exit <- 365 * births$exit
-addup <- births$exit <= 0
-births$exit[addup] <- 0.25
-infa <- age.window(births, c(0, 365))
-infa$enter <- g(infa$enter)
+badyears <- years[imr > 0.6]
+goodinf <- infant[!(infant$year %in% badyears), ]
+goodinf$exit <- g(goodinf$exit)
+goodinf$enter <- g(goodinf$enter)
+rs <-  with(goodinf, risksets(Surv(enter, exit, event), members = FALSE))
+drawBB(rs, main = "IMR less than 60 percent")
+par(oldpar)
+```
+
+Tried a couple of cut points, no important differences.
+
+### Hazards by socst
+
+```{r socst, fig.cap = "Upper and lower classes."}
+infa <- infant
+oldpar <- par(mfrow = c(1, 2))
 infa$exit <- g(infa$exit)
-pf(infa)
+xxx <- with(infa[infa$socst %in% c("tjänsteman", "präst", "adel"), ], 
+            Surv(enter, exit, event))
+rs <- risksets(xxx, members = FALSE)
+drawBB(rs, main = "Upper")
+###
+xxx <- with(infa[!(infa$socst %in% c("tjänsteman", "präst", "adel")), ], 
+            Surv(enter, exit, event))
+rs <- risksets(xxx, members = FALSE)
+drawBB(rs, main = "Lower")
+par(oldpar)
 ```

### 2023-03-06 R/bio.R

@@ -1,0 +1,50 @@
+bio <- function(surv, timescale = "year", graph = FALSE){
+   library(eha)
+   ## surv is a data frame, containing a survival object in three columns:  
+   ## enter, exit, event.
+   ##
+   g <- function(x) (log(x + 1))^3
+   ginv <- function(x) exp(x^(1/3)) - 1
+   #####
+   
+   #####
+   if (timescale == "year") {
+      surv <- age.window(surv, c(0, 1))
+      surv[, "enter"]  <- 365 * surv[, "enter"]
+      surv[, "exit"] <- 365 * surv[, "exit"]
+   }else{
+      surv <- age.window(surv, c(0, 365)) # Time scale in days
+   }
+   g <- function(x) (log(x + 1))^3
+   ginv <- function(x) exp(x^(1/3)) - 1
+   #####
+  
+   gsurv <- surv
+   gsurv[, "enter"] <- g(surv[, "enter"])
+   gsurv[, "exit"] <- g(surv[, "exit"]) 
+   ######
+  
+   post <- age.window(surv, c(30, 365))
+   gpost <- age.window(gsurv, c(g(30), g(365)))
+   ##
+   neo <- age.window(surv, c(0, 30))
+   gneo <- age.window(gsurv, c(0, g(30)))
+   #####
+   ## postneo mortality, g-scale;
+   gprate <- with(gpost, sum(event) / sum(exit - enter))
+   gprate
+   ## neo cum. hazards, g-scale
+   x <- with(gneo, risksets(Surv(enter, exit, event), members = FALSE))
+   x$risktimes <- c(0, x$risktimes)
+   x$n.events <- c(0, x$n.events)
+   x$size <- c(NROW(gneo), x$size)
+   cumhaz <- cumsum(x$n.events / x$size)
+   if (graph){
+      plot(x$risktimes, cumhaz, type = "s", xlab = "Day", ylab = "Cum. hazards")
+      lines(c(0, 40), c(0, gprate * 40))
+   }
+   ##
+   endocum <- cbind(x$risktimes, cumhaz)
+   colnames(endocum) <- c("age", "cumhas")
+   list(exorate = gprate, endocum = endocum)
+}

### 2023-03-06 R/biometric.R

@@ -1,0 +1,137 @@
+biometric <- function(surv, timescale = "day", graph = FALSE, C = 1){
+   ## timescale can be "year", otherwise "day".
+   library(eha)
+   ## surv is a data frame, containing a survival object in three columns:  
+   ## enter, exit, event.
+   ##
+   ## NOTE that indata is NOT "g-transformed"!
+   ####
+   g <- function(x) C * (log(x + 1))^3
+   ginv <- function(x) exp((x/C)^(1/3)) - 1
+   dg <- function(x) C * (log(x + 1))^2 /(x + 1)
+   #####
+   
+   hexo <- function(x, lambda) {
+      lambda * dg(x)
+   }
+   Hexo <- function(x, lambda){
+      lambda * g(x)
+   }
+   Sexo <- function(x, lambda){
+      exp(-Hexo(x, lambda))
+   }
+   ######################################
+   
+   ##### All calculations are done on "day" time-scale, then translated back to
+   ##### the "original" time-scale ("year" or "day")
+   if (timescale == "year") {
+      ##surv <- age.window(surv, c(0, 1))
+      surv[, "enter"]  <- 365 * surv[, "enter"]
+      surv[, "exit"] <- 365 * surv[, "exit"]
+   }
+   nulls <- surv$exit <= 0
+   surv$exit[nulls] <- 0.25 # Note: 'age.window' would remove deaths at age 0.
+   surv <- age.window(surv, c(0, 365)) # Time scale in days.
+
+   ## Estimation of lambda = rate in exogenous mortality.
+   
+   exorate <- function(surv){
+      ## Use "postneo" data:
+      surv <- age.window(surv, c(29.5, 365))
+      ##pgrs <- with(surv, risksets(Surv(enter, exit, event), members = FALSE))
+      ## g- transformed data is only used in "postneo" estimation (exogenous mort).
+      ##gsurv <- surv
+      surv[, "enter"] <- g(surv[, "enter"])
+      surv[, "exit"] <- g(surv[, "exit"]) 
+      ######
+      lambda <- with(surv, sum(event) / sum(exit - enter)) # occ. / exp. rate
+      pgrs <- with(surv, risksets(Surv(enter, exit, event), members = FALSE))
+      list(lambda = lambda, pgrs = pgrs)
+   }   
+   ##
[…]

### 2023-03-06 R/compare.R

@@ -1,0 +1,25 @@
+compare <- function(infdat){
+   srv <- with(infdat, cbind(enter, exit, event))
+   rs <- risksets(srv, members = FALSE)
+   tid <- c(0, rs$risktime)
+   ## haz:
+   haz <- c(0, with(rs, cumsum(n.events/size)))
+   plot(tid, haz, type = "s", main = "Cum. hazards", axes = FALSE, xlab = "Day")
+   axis(1, at = c(0, g(30), g(365)), labels = c(0, 30, 365))
+   axis(2, las = 1)
+   abline(v = g(30), lty = 3, col = "darkgreen")
+   box()
+   n0 <- min(which(tid > g(30)))
+   n <- length(tid)
+   lines(tid[c(n0, n)], haz[c(n0, n)], lty = 2, col = "red")
+   abline(h = 0, v = 0)
+   ## BP:
+   prob <- c(0, with(rs, cumsum(n.events / size[1])))
+   plot(tid, prob, type = "s", main = "CDF", axes = FALSE, xlab = "Day")
+   axis(1, at = c(0, g(30), g(365)), labels = c(0, 30, 365))
+   axis(2, las = 1)
+   abline(v = g(30), lty = 3, col = "darkgreen")
+   box()
+   lines(tid[c(n0, n)], prob[c(n0, n)], lty = 2, col = "red")
+   abline(h = 0, v = 0)
+}

### 2023-03-06 R/denhaz.R

@@ -1,0 +1,43 @@
+denhaz <- function(p, header = ""){
+   ## p = Prob(T <= 365 days)
+   ##
+   ## Uniform:
+   ## S(365) = 1 - p; 
+   ## S(t) = p * t / 365
+   ## f * 365 = p, f = p / 365
+   ##
+   ## Same as in GBTB23:
+   g <- function(x) (log(1 + x))^3 * 365 / (log(366))^3
+   ##
+   funif <- function(x) rep(p / 365, length(x))
+   Sunif <- function(x) 1 - p * x / 365
+   hunif <- function(x) p / (365 - p * x)
+   ## Exponential:
+   ## S(365) = exp(-b*365) = 1 - p, b = - log(1 - p) / 365
+   ## f(t) = b * exp(-t * b), 0 < t < 365
+   ##
+   b <- -log(1 - p) / 365
+   fexp <- function(x) b * exp(-x * b)
+   Sexp <- function(x) exp(-x * b)
+   hexp <- function(x) rep(b, length(x))
+   ##
+   expneofrac <- (1 - Sexp(g(30))) / (1 - Sexp(g(365)))  
+   unifneofrac <- (1 - Sunif(g(30))) / (1 - Sunif(g(365)))
+   t <- seq(0, 365, length = 1000)
+   if (TRUE){
+   plot(t, 1000 * funif(t), type = "l", ylim = c(0, 1.5 * 1000 * max(fexp(t))), col = "blue",
+        axes = FALSE, xlab = "g(Days)", ylab = "1000 x density", main = header)
+   }
+   text(g(200), 1200 * max(fexp(t)), "Uniform", col = "blue")
+   text(g(200), 500 * max(fexp(t)), "Exponential", col = "red")
+  ## plot(t, Sunif(t), type = "l", ylim = c(1 - p, 1), col = "blue",
+    ##    axes = FALSE, xlab = "g(Days)", ylab = "")
+   axis(1, at = c(0, g(30), 365), labels = c("0", "g(30)", "365"))
+   axis(2, las = 1)
+   box()
+   lines(t, 1000 * fexp(t), col = "red")
+   ##lines(t, Sexp(t), col = "red")
+   abline(h = 0, v = 0)
+   abline(v = g(30), lty = 3, col = "darkgreen")
+   list(expneofrac = expneofrac, unifneofrac = unifneofrac)
+}

### 2023-03-06 R/drawBB.R

@@ -1,0 +1,35 @@
+drawBB <- function(rs, main = "", ylim){
+   # rs is the output from eha::risksets
+   # Assumed that age data are on the "g" scale (in days)
+   # Error on line 9 ("cumsum <- ...") fixed
+   oldpar = par(cex = 0.7)
+   
+   n <- rs$size[1]
+   x <- c(0, rs$risktimes)
+   y <- c(0, cumsum(rs$n.events / rs$size)) # Nominator differ from "BP"
+   if (missing(ylim)){
+      plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day", 
+           ylab = "Cum. death fraction", main = main)
+   }else{
+      plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day",
+           ylab = "Cumulative hazards", main = main, ylim = ylim)
+   }
+   att <- c(0, 28, 91, 183, 274, 365)
+   axis(1, at = g(att), labels = att)
+   
+   abline(h = 0, v = 0)
+   abline(v = g(28), lty = 2, col = "magenta", lwd = 1.5)
+   abline(v = c(g(91), g(183), g(274), g(365)), lty = 3, col = "red")
+   n2 <- length(x)
+   x2 <- x[n2]
+   y2 <- y[n2]
+   n1 <- min(which(x > g(28)))
+   x1 <- x[n1]
+   y1 <- y[n1]
+   x0 <- 0
+   y0 <- y1 - x1 * (y2 - y1) / (x2 - x1)
+   axis(2, at = round(c(0, y0, y1, y2), 3), las = 1)
+   box()
+   lines(c(0, x2), c(y0, y2), type = "l", lty = 2, col = "darkgreen")
+   
+}

### 2023-03-06 R/drawBP.R

@@ -1,0 +1,36 @@
+drawBP <- function(rs, main = "", ylim){
+   # rs is the output from eha::risksets
+   # Assumed that age data are on the "g" scale (in days)
+   
+   oldpar <- par(cex = 0.7)
+   n <- rs$size[1]
+   x <- c(0, rs$risktimes)
+   y <- c(0, cumsum(rs$n.events) / rs$size[1])
+   if (missing(ylim)){
+      plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day", 
+           ylab = "Cum. death fraction", main = main)
+   }else{
+      plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day",
+           ylab = "Cum. death fraction", main = main, ylim = ylim)
+   }
+   att <- c(0, 28, 183, 274, 365)
+   axis(1, at = g(att), labels = att)
+   
+   abline(h = 0, v = 0)
+   ##abline(v = g(28), lty = 3, col = "red")
+   
+   abline(v = g(28), lty = 2, col = "magenta", lwd = 1.5)
+   abline(v = c(g(91), g(183), g(274), g(365)), lty = 3, col = "red")
+   n2 <- length(x)
+   x2 <- x[n2]
+   y2 <- y[n2]
+   n1 <- min(which(x > g(28)))
+   x1 <- x[n1]
+   y1 <- y[n1]
+   x0 <- 0
+   y0 <- y1 - x1 * (y2 - y1) / (x2 - x1)
+   axis(2, at = round(c(0, y0, y1, y2), 3), las = 1)
+   box()
+   lines(c(0, x2), c(y0, y2), type = "l", lty = 2, col = "darkgreen")
+   par(oldpar)
+}

### 2023-03-06 R/exo.R

@@ -1,0 +1,28 @@
+## The 'exo' distribution, given by a model for exogeneous
+## infant mortality.
+
+tc <- 365 / (log(366))^2 # scaling constant
+
+g <- function(x, tc = 365 / (log(366))^2){
+    (log(1 + x))^3 * tc
+}
+dg <- function(x, tc = 365 / (log(366))^2){
+   3 * (log(1 + x))^2 * tc / (1 + x)
+}
+##
+## Survival function:
+Sexo <- function(x, lambda){
+   exp(-lambda * g(x))
+}
+## Density function:
+fexo <- function(x, lambda){
+   lambda * dg(x) * exp(-lambda * g(x))
+}
+## Hazard function:
+hexo <- function(x, lambda){
+   lambda * dg(x)
+}
+## Cumulative hazard function:
+Hexo <- function(x, lambda){
+   lambda * g(x)
+}

### 2023-03-06 R/drawBP.R
small modification

@@ -13,7 +13,7 @@ drawBP <- function(rs, main = "", ylim){
       plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day",
            ylab = "Cum. death fraction", main = main, ylim = ylim)
    }
-   att <- c(0, 28, 183, 274, 365)
+   att <- c(0, 28, 91, 183, 274, 365)
    axis(1, at = g(att), labels = att)
    
    abline(h = 0, v = 0)

### 2023-03-06 paper2.Rmd
first commit

--- 2023-03-06 paper.Rmd
@@
 # Introduction
+
+In this version the cumulative hazard functions are depicted as-is, but on the
+log-cube time scale. See the new **R** function *drawBB2*.
 
@@
 r1 <- with(ginfant[ginfant$period == "1861-1890", ], eha::Surv(enter, exit, event))
+g1 <- infant[infant$period == "1861-1890", ]
 r1 <- eha::risksets(r1, members = FALSE)
@@
 ##
+g3 <- infant[infant$period == "1921-1950", ]
 r3 <- with(ginfant[ginfant$period == "1921-1950", ], eha::Surv(enter, exit, event))
@@
 ```{r second, fig.cap = "The periods 1861--1890 and 1921--1950 in Västerbotten, hazards method."}
-source("R/drawBB.R")
+source("R/drawBB2.R")
 oldpar <- par(mfrow = c(1,2))
@@
 ##r1 <- eha::risksets(r1, members = FALSE)
-drawBB(r1, main = "1861-1890", ylim = c(0, 0.115))
+drawBB2(g1, main = "1861-1890")
 ##
@@
 ##r3 <- eha::risksets(r3, members = FALSE)
-drawBB(r3, main = "1921-1950", ylim = c(0, 0.115))
+drawBB2(g3, main = "1921-1950")
 par(oldpar)
@@
 pf <- function(x){
-   ## x is a dataframe on g scale
-   rs <- with(x, eha::risksets(Surv(enter, exit, event), members = FALSE))
+   ## x is a dataframe NOT on g scale
+    y <- x
+    y$enter <- g(y$enter)
+    y$exit <- g(y$exit)
+   rs <- with(y, eha::risksets(Surv(enter, exit, event), members = FALSE))
    oldpar <- par(mfrow = c(1, 2))
-   drawBB(rs, main = "Hazards method")
+   drawBB2(x, main = "Hazards method")
    drawBP(rs, main = "B-P method")
@@
 ##rs.post <- risksets()
-ume <- ginfant[ginfant$birthdate > "1920-12-31", ]
+ume <- infant[infant$birthdate > "1920-12-31", ]
 pf(ume)
@@
 ##rs.post <- risksets()
-ume$exit <- g(ume$exit)
-ume$enter <- g(ume$enter)
+##ume$exit <- g(ume$exit)
+##ume$enter <- g(ume$enter)
 pf(ume)
@@
 sca1 <- sca1[sca1$BirthDate < "1801-01-01", ]
+sca11 <- sca1
 sca1$exit <- g(sca1$exit)
@@
 oldpar <- par(mfrow = c(1, 2))
-drawBB(rs, main = "Hazards method")
+drawBB2(sca11, main = "Hazards method")
 drawBP(rs, main = "B-P method")
@@
 infa <- age.window(births, c(0, 365))
-infa$enter <- g(infa$enter)
-infa$exit <- g(infa$exit)
+##infa$enter <- g(infa$enter)
+##infa$exit <- g(infa$exit)
 pf(infa)

### 2023-03-06 R/drawBB2.R

@@ -1,0 +1,78 @@
+drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3){
+    
+    ## dat is a "survival data frame with enter, exit (natural scale) and event.
+    g <- function(x) {
+        C * (log(x + 1))^3
+    }
+    dat$enter <- g(dat$enter)
+    dat$exit <- g(dat$exit)
+    post <- eha::age.window(dat, c(g(28), g(365)))
+    rate <- with(post, sum(event) / sum(exit - enter))
+    ## Only change wrt drawBB is that all Cumulative hazards start at (0, 0).    
+    
+   # rs is the output from eha::risksets
+   # Assumed that age data are on the "g" scale (in days)
+   # Error on line 11 ("cumsum  ...") fixed
+   oldpar = par(cex = 0.7)
+   fit <- coxreg(Surv(enter, exit, event) ~ 1, data = dat)
+   xxx <- hazards(fit)
+   x <- c(0, xxx[[1]][, 1])
+   y <- cumsum(c(0, xxx[[1]][, 2]))
+   
+   plot(x, y, type = "s", main = main, xlab = "Day", axes = FALSE, 
+        ylab = "Cumulative hazards")
+   lines(c(0, g(365)), c(0, rate * g(365)), lty = 2, col = "blue")
+   att <- c(0, 28, 91, 183, 274, 365)
+   axis(1, at = g(att), labels = att)
+   abline(v = g(28), lty = 2, col = "darkgreen", lwd = 1.5)
+   y0 <- rate * g(28)
+   cat("rate = ", rate, "g(28) = ", g(28), "y0 = ", y0, "\n")
+   ##xxx <- max(xx[[1]])
+   n <- max(which(x <= g(28)))
+   y1 <- y[n]
+   red <- y[1:n] - x[1:n] * rate
+   lines(x[1:n], red, col = "red", type = "s", lty = 4)
+   y2 <- red[n]
+   y3 <- max(y)
+   y4 <- rate * g(365)
+   axis(2, at = round(c(0, y0, y1, y2, y3, y4), 3), las = 1)
+   box()
+   lines(c(0, g(28)), c(y0, y0), lty = 3, lwd = 0.5, col = "darkgreen")
+   lines(c(0, g(28)), c(y1, y1), lty = 3, lwd = 0.5, col = "darkgreen")
+   lines(c(0, g(28)), c(y2, y2), lty = 3, lwd = 0.5, col = "darkgreen")
[…]

### 2023-03-08 paper.Rmd
following Tommy's arrows

@@
 using the present risk-set size, that is, the the total number of infants still alive 
-immediately prior to the death age under investigation. It is easily achieved by the use of 
-the function `risksets` in the package `eha`. See Table \@ref(tab:risksets) for 
+and under observation immediately prior to the death age under investigation. Given 
+proper individual-based data, it is easily achieved. One example is the use of 
+the function `risksets` in the **R** package `eha`. See Table \@ref(tab:risksets) for 
 an illustration.
@@
 
-See Figure \@ref(fig:both) for a comparison of the two curves.
-
-```{r both, fig.cap = "Bourgeois-Pichat and 'hazards method' plots."}
-oldpar = par(mfrow = c(1, 2), cex = 0.7, las = 1)
-plot(r1$risktimes, cumsum(r1$n.events) / r1$size[1], type = "l", col = "blue",
-     main = "1861-1890", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
-lines(r1$risktimes, cumsum(r1$n.events / r1$size), type = "l", col = "red")
-abline(h = 0, v = 0)
-text(120, 0.08, "Hazard method", col = "red")
-text(200, 0.02,"Bourgeois-Pichat", col = "blue")
-##
-plot(r3$risktimes, cumsum(r3$n.events) / r3$size[1], type = "l", col = "blue",
-     main = "1921-1950", ylim = c(0, 0.1), xlab = "g(Day)", ylab = "Cum. hazards")
-lines(r3$risktimes, cumsum(r3$n.events / r3$size), type = "l", col = "red")
-abline(h = 0, v = 0)
-text(120, 0.05, "Hazard method", col = "red")
-text(200, 0.015,"Bourgeois-Pichat", col = "blue")
-##
-##
-par(oldpar)
-```
-
-The difference between the two curves increases as infant mortality increases, 
-but both methods show an excellent fit to linearity.
+Both methods show an excellent fit to linearity for this data set.
 
@@
 
-In order to see this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 30.
+In order to see this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 28 days.
 
@@
 
-The *Exponential* fit is excellent, with a slight edge for the early period data, 
-where post-neonatal mortality is high. However, one question pops up: If the exponential 
-assumption is in fact true, will it also hold for subgroups of the data?
+The *Exponential* fit is excellent, possibly with a slight edge for the early period, 
+where post-neonatal mortality is high.
 
-The theoretical ansver is "No", because a mixture of two exponential populations with
-different mortality rates cannot itself be exponential. A heuristic argument for that: Assume two equally-sized 
-groups at "birth", but with different levels of mortality. As time passes, there will be more deaths 
-in the high-mortality group, and that means that the proportion survivors in the low-level group
-will increase. Therefore, the population-level mortality will seem to decrease over time, 
-and not being constant, as the exponentiality prescribes.
-
-We can see what happens when the two periods in Figure \@ref(fig:postneo) are joined, see Figure
-\@ref(fig:joined).
-
-```{r joined, fig.cap = "Exponential fit for two periods, Västerbotten."}
-oldpar = par(cex = 0.7)
-p13 <- post[post$period != "1891-1920", ]
-p13$period <- factor(p13$period)
-fit13 <- coxreg(Surv(exit - enter, event) ~ 1, data = p13)
-wfit13 <- phreg(Surv(exit - enter, event) ~ 1, data = p13, shape = 1)
-plot(fit13, main = "1861-1890 and 1921-1950", xlab = "g(Days)", printLegend = FALSE)
-whaz13 <- hazards(wfit13)
-with(whaz13, lines(x, y, lty = 2, col = "red"))
-legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
-       col = c("blue", "red"), cex = 1)
-#
-par(oldpar)
-```
-
-Not what we expected: The fit seems to be extremely good in both figures! However,
-the rate is close to zero, and so differences are small and hard to notice. And for practical use, they are not interesting.
-
-What about the original Bourgeois-Pichat model, is it sensitive to taking subsets?
-Let us see.
-
-```{r letse, fig.cap = "The period 1861-1890, Bourgeouis-Pichat method by sex."}
-source("R/drawBP.R")
-oldpar <- par(mfrow = c(1,2))
-##
-r1 <- with(post[post$sex == "girl", ], eha::Surv(enter, exit, event))
-r1 <- eha::risksets(r1, members = FALSE)
-drawBP(r1, main = "Girls", ylim = c(0, 0.055))
-##
-##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
-##r2 <- eha::risksets(r2, members = FALSE)
-##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
-##
-r3 <- with(post[post$sex == "boy", ], eha::Surv(enter, exit, event))
-r3 <- eha::risksets(r3, members = FALSE)
-drawBP(r3, main = "Boys", ylim = c(0, 0.055))
-par(oldpar)
-```
-
-In Figure \@ref(fig:letse) we notice a weak tendency towards convexity for boys.
-Not a big deal.
+What about the original Bourgeois-Pichat model? See Figure \@ref(fig:first), 
+ages larger than 20 days. Very good fit.
 

### 2023-03-08 paper2.Rmd
following Tommy's arrows

@@
 - \renewcommand{\efloatseparator}{\mbox{}}
-- \subsectionfont{\normalfont\itshape}
+##- \subsectionfont{\normalfont\normalfont\itshape}
 abstract: "A variation of the Bourgeois-Pichat biometric analysis of infant mortality is suggested.
@@
 using the present risk-set size, that is, the the total number of infants still alive 
-immediately prior to the death age under investigation. It is easily achieved by the use of 
-the function `risksets` in the package `eha`. See Table \@ref(tab:risksets) for 
-an illustration.
+immediately prior to the death age under investigation. Given proper data, it is easily achieved.
+For instance,by the use of 
+the function `risksets` in the **R** package `eha` we can get Table \@ref(tab:risksets).
+
 ```{r risksets, echo = FALSE, results = 'asis'}
@@
 ```
+
 It starts with `r as.integer(lt1[1, 3])` live births, of which `r lt1[1, 4]` dies on the day of birth, 
@@
 
-In order to see this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 30.
+In order to see this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 28.
 
@@
 ##check.dist(fit1, wfit1, main = "1861-1890", printLegend = FALSE)
-plot(fit1, main = "1861-1890", printLegend = FALSE, xlab = "g(Day)", col = "blue")
+plot(fit1, main = "1861-1890", printLegend = FALSE, xlab = "g(Day)", col = "blue",
+     ylab = "Cumulative hazards")
 whaz1 <- hazards(wfit1)
@@
 The *Exponential* fit is excellent, with a slight edge for the early period data, 
-where post-neonatal mortality is high. However, one question pops up: If the exponential 
-assumption is in fact true, will it also hold for subgroups of the data?
+where post-neonatal mortality is high.
 
-The theoretical ansver is "No", because a mixture of two exponential populations with
-different mortality rates cannot itself be exponential. A heuristic argument for that: Assume two equally-sized 
-groups at "birth", but with different levels of mortality. As time passes, there will be more deaths 
-in the high-mortality group, and that means that the proportion survivors in the low-level group
-will increase. Therefore, the population-level mortality will seem to decrease over time, 
-and not being constant, as the exponentiality prescribes.
-
-We can see what happens when the two periods in Figure \@ref(fig:postneo) are joined, see Figure
-\@ref(fig:joined).
-
-```{r joined, fig.cap = "Exponential fit for two periods, Västerbotten."}
-oldpar = par(cex = 0.7)
-p13 <- post[post$period != "1891-1920", ]
-p13$period <- factor(p13$period)
-fit13 <- coxreg(Surv(exit - enter, event) ~ 1, data = p13)
-wfit13 <- phreg(Surv(exit - enter, event) ~ 1, data = p13, shape = 1)
-plot(fit13, main = "1861-1890 and 1921-1950", xlab = "g(Days)", printLegend = FALSE)
-whaz13 <- hazards(wfit13)
-with(whaz13, lines(x, y, lty = 2, col = "red"))
-legend("topleft", legend = c("Non-parametric", "Exponential"), lty = 1:2, 
-       col = c("blue", "red"), cex = 1)
-#
+The fit of the *Uniform* distribution is already shown in Figure \@ref(fig:first).
+Also a very good fits in the post-neonatal period.
+<!--
+```{r postneoBB, fig.cap = "Uniform fits to postneonatal data, Västerbotten.", eval = FALSE}
+oldpar <- par(mfrow = c(1, 2), cex = 0.7)
+##
+rs1 <- risksets(with(p1, Surv(enter, exit, event), members = FALSE))
+x <- c(g(28), rs1$risktimes)
+y <- c(0, rs1$n.events / rs1$size[1])
+y <- cumsum(y)
+plot(x, y, type = "s", axes = FALSE, xlab = "Day", ylab = "EDF", 
+     main = "1861-1890")
+axis(1, at = c(g(28), g(365)), labels = c(28, 365))
+axis(2, at = c(0, max(y)), labels = c(0, round(max(y), 2)))
+lines(c(g(28), g(365)), c(0, max(y)), lty = 2, col = "red")
+box()
+##
+rs3 <- risksets(with(p3, Surv(enter, exit, event), members = FALSE))
+x <- c(g(28), rs3$risktimes)
+y <- c(0, rs3$n.events / rs1$size[1])
+y <- cumsum(y)
+plot(x, y, type = "s", axes = FALSE, xlab = "Day", ylab = "EDF", 
+     main = "1921-1950")
+axis(1, at = c(g(28), g(365)), labels = c(28, 365))
+axis(2, at = c(0, max(y)), labels = c(0, round(max(y), 2)))
+lines(c(g(28), g(365)), c(0, max(y)), lty = 2, col = "red")
+box()
+##
 par(oldpar)
@@
 
-Not what we expected: The fit seems to be extremely good in both figures! However,
-the rate is close to zero, and so differences are small and hard to notice. And for practical use, they are not interesting.
-
-What about the original Bourgeois-Pichat model, is it sensitive to taking subsets?
-Let us see.
-
-```{r letse, fig.cap = "The period 1861-1890, Bourgeouis-Pichat method by sex."}
-source("R/drawBP.R")
-oldpar <- par(mfrow = c(1,2))
-##
-r1 <- with(post[post$sex == "girl", ], eha::Surv(enter, exit, event))
-r1 <- eha::risksets(r1, members = FALSE)
-drawBP(r1, main = "Girls", ylim = c(0, 0.055))
-##
-##r2 <- with(ginfant[ginfant$period == "1891-1920", ], eha::Surv(enter, exit, event))
-##r2 <- eha::risksets(r2, members = FALSE)
-##drawBP(r2, main = "1891-1920", ylim = c(0, 0.1))
-##
-r3 <- with(post[post$sex == "boy", ], eha::Surv(enter, exit, event))
-r3 <- eha::risksets(r3, members = FALSE)
-drawBP(r3, main = "Boys", ylim = c(0, 0.055))
-par(oldpar)
-```
-
-In Figure \@ref(fig:letse) we notice a weak tendency towards convexity for boys.
-Not a big deal.
+-->
 
@@
 @manfred04 argues that a less satisfactory property of the *B-P* model is that this fraction is constant,
-19.7 percent, independent of the overall level of infant mortality. This is not the case for the hazards model, see Figure \@ref(fig:exog).
-For instance, for $p = 0.2$ the fraction in question is 21.5 percent.
+18.56 percent, independent of the overall level of infant mortality $p$. This is not the case for the hazards model, see Table \@ref(tab:exog).
 
-```{r exog, fig.height = 8, fig.cap = "Densities for uniform and exponential distributions for varying death probability p."}
-source("R/denhaz.R")
-p <- c(0.01, 0.05, 0.1, 0.2)
-oldpar <- par(mfrow = c(2, 2))
-for (i in 1:4){
-   rr <- denhaz(p[i], header = paste("p =", p[i]))
+```{r exog}
+source("R/tbl.R")
+p <- c(0.005, 0.01, 0.05, 0.1, 0.2)
+lambda <- -log(1 - p) / 365
+n <- length(p)
+exo28 <- numeric(n)
+exo365 <- numeric(n)
+for (i in 1:n){
+    exo28[i] <- pexp(g(28), lambda[i]) * 1000
+    exo365[i] <- pexp(g(365), lambda[i]) * 1000
 }
-par(oldpar)
+utt <- cbind(exo28, exo365, round(100 * (exo28 / exo365), 2))
+utt <- round(utt, 3)
+colnames(utt) <- c("Neonatal", "Total", "Fraction")
+tbl(utt, caption = "Distribution over age periods of exogenous mortality (per 1000).")
 ```
@@
 \begin{equation}
-\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t, 0 < t < g(28).
+\hat{E}(t) = \hat{A}(t) - \hat{\lambda} t,\; 0 < t \le g(28).
 \end{equation}
@@
 
-Post-neonatal mortality is the same as exogenous mortality in the sense that the
-hazard functions are the same on the post-neonatal age interval. So the first step
-involves data left truncated at age 28 days (or at `r round(g(28))`) on the $g$ scale.
-We exemplify the numerical procedure with the aid of the **eha** [@eha; @ehar2].
-The first lines of the data set are (Table \@ref(tab:firstlines))
 ```{r firstlines, results='asis'}
-
-tbl(infant[1:5, ], caption = "First five rows of data frame.")
+tbl(infant[1:5, 1:6], caption = "First five rows of data frame.")
 nro <- nrow(infant)
@@
 ```
-and the number of rows is `r nro`, and the number of deaths is `r deat`,
-resulting in a crude death probability of `r round(1000 * deat / nro)` per thousand
-live births.
 
-So, our first step is to truncate and rescale. We show how it is done in **R**:
-
-```{r retrunc, echo = TRUE}
+```{r retrunc, echo = FALSE}
 postneo <- eha::age.window(infant, c(28, 365))
 gpostneo <- eha::age.window(ginfant, c(g(28), g(365)))
-print(postneo[1:5, ], row.names = FALSE) # print first five rows
-print(gpostneo[1:5, ], row.names = FALSE) # first five rows after...
+days <- sum(gpostneo$event)
+expo <- with(gpostneo, sum(exit - enter))
+rate <- days / expo
 ```
 
-Now, the post-neonatal hazard function is *constant*, and its ML estimator is simply 
-the *occurrence/exposure* rate, or in **R**,
-
-```{r occexp, echo = TRUE}
-D <- sum(gpostneo$event) # No. of deaths
-E <- with(gpostneo, sum(exit - enter)) # exposure
-(rate <- D/E)
-```
+Post-neonatal mortality is the same as exogenous mortality in the sense that the
+hazard functions are the same on the post-neonatal age interval. So the first step
+involves data left truncated at age 28 days. The maximum-likelihood estimator of the rate 
+in the exponential distribution is simply the number of observed post-neonatal deaths
+`r days`, 
+divided by the total post-neonatal exposure time `r expo` (still on the $g$ scale), 
+giving a rate of `r rate`.
 

### 2023-03-08 R/drawBB2.R

@@ -26,12 +26,14 @@ drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3){
    axis(1, at = g(att), labels = att)
    abline(v = g(28), lty = 2, col = "darkgreen", lwd = 1.5)
    y0 <- rate * g(28)
-   cat("rate = ", rate, "g(28) = ", g(28), "y0 = ", y0, "\n")
+   ##cat("rate = ", rate, "g(28) = ", g(28), "y0 = ", y0, "\n")
    ##xxx <- max(xx[[1]])
    n <- max(which(x <= g(28)))
    y1 <- y[n]
    red <- y[1:n] - x[1:n] * rate
+   redall <- y - x * rate
    lines(x[1:n], red, col = "red", type = "s", lty = 4)
+   lines(x, redall, col ="magenta", type = "s", lty = 3)
    y2 <- red[n]
    y3 <- max(y)
    y4 <- rate * g(365)

### 2023-03-08 paper2.Rmd
cleaning and adding reference

@@
 biblio-style: apalike
+setcitestyle: authoryear
 documentclass: article
@@
 header-includes:
+- \usepackage{natbib}
+- \setcitestyle{authoryear}
 - \usepackage{a4wide}
@@
 A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec.
+@knoki77 argues that breastfeeding patterns have a strong impact on the linearity.
 The curve for Sweden 1910--1946 show, however, no such downward bend [@bp51b, Figure 9; @bp52].
@@
 fit1 <- coxreg(Surv(exit - enter, event) ~ 1, data = p1)
+haz1 <- hazards(fit1)
+x <- c(0, haz1[[1]][, 1])
+y <- c(0, cumsum(haz1[[1]][, 2]))
+##       
 wfit1 <- phreg(Surv(exit - enter, event) ~ 1, data = p1, shape = 1)
 ##check.dist(fit1, wfit1, main = "1861-1890", printLegend = FALSE)
-plot(fit1, main = "1861-1890", printLegend = FALSE, xlab = "g(Day)", col = "blue",
-     ylab = "Cumulative hazards")
+plot(x, y, main = "1861-1890", xlab = "Day", col = "blue",
+     ylab = "Cumulative hazards", axes = FALSE, type = "s")
+axis(2, las = 1)
+axis(1, at = c(0, g(365) - g(28)), labels = c(28, 365))
+abline(h = 0)
+box()
 whaz1 <- hazards(wfit1)
@@
 fit3 <- coxreg(Surv(exit - enter, event) ~ 1, data = p3)
+haz3 <- hazards(fit3)
+x <- c(0, haz3[[1]][, 1])
+y <- c(0, cumsum(haz3[[1]][, 2]))
 wfit3 <- phreg(Surv(exit - enter, event) ~ 1, data = p3, shape = 1)
 whaz <- hazards(wfit3)
-plot(fit3, main = "1921-1950", printLegend = FALSE, xlab = "g(Day)")
+plot(x, y, main = "1921-1950", xlab = "Day", col = "blue", 
+     ylab = "Cumulativa hazards", axes = FALSE, type = "s")
+axis(2, las = 1)
+axis(1, at = c(0, g(365) - g(28)), labels = c(28, 365))
+box()
+abline(h = 0)
 with(whaz, lines(x, y, lty = 2, col = "red"))
@@
 $\hat{A}(t)$ to get $\hat{E}(t)$.
-\begin{equation}
+\begin{equation} (\#eq:endog)
 \hat{E}(t) = \hat{A}(t) - \hat{\lambda} t,\; 0 < t \le g(28).
@@
 in the exponential distribution is simply the number of observed post-neonatal deaths
-`r days`, 
-divided by the total post-neonatal exposure time `r expo` (still on the $g$ scale), 
+`r as.integer(days)`, 
+divided by the total post-neonatal exposure time `r format(round(expo), scientific = FALSE)` (still on the $g$ scale), 
 giving a rate of `r rate`.
@@
 
-Here we focus on neonatal period.
-
-The estimation of the cumulative hazard function for the total neonatal mortality is
+The estimation of the cumulative hazard function for the total neonatal mortality is standard. <!--
 easily achieved by calling the function `hazards` in the *eha* package, see also Figure \@ref(fig:hazards).
-
-```{r hazards, fig.cap = "Cumulative hazard functions for neonatal mortality.", echo = TRUE}
+-->
+```{r hazards, eval = TRUE}
 neo <- age.window(ginfant, c(0, g(28)))
@@
 y <- c(0, 1000 * cumsum(hneo[[1]][, 2]))
-plot(x, y, type = "l", xlab = "Day", ylab = "by 1000")
-lines(x, 1000 * rate * x, col = "blue", lty = 2)
-abline(h = 0, v = 0)
-text(8, 15, "Total", col = "black")
-text(8, 5, "Exogenous", col = "blue")
+##plot(x, y, type = "s", xlab = "Day", ylab = "by 1000", axes = FALSE)
+##axis(2, las = 1)
+##axis(1, at = c(g(0), g(7), g(14), g(21), g(28)), labels = c(0, 7, 14, 21, 28))
+##axis(2)
+##box()
+##lines(x, 1000 * rate * x, col = "blue", lty = 2)
+##abline(h = 0, v = 0)
+##text(8, 15, "Total", col = "black")
+##text(8, 5, "Exogenous", col = "blue")
 ```
 
-Next, take the difference between "Total" and "Exogenous" to get "Endogenous", Figure \@ref(fig:enof).
+Next, take the difference between "Total" and "Exogenous", as in Equation \@ref(eq:endog) to get "Endogenous",
+the result is shown in Figure \@ref(fig:enof).
 
-```{r enof, fig.cap = "Cumulative hazard function for endogenous neonatal mortality.", fig.width = 7}
+```{r enof, fig.cap = "Cumulative hazard functions for neonatal mortality, Västerbotten.", fig.width = 7}
 yen <- y - 1000 * rate * x
 par(las = 1)
-plot(x, yen, type = "l", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 36))
+plot(x, yen, type = "s", col = "red", xlab = "Day", ylab = "by 1000", ylim = c(0, 36), axes = FALSE)
+axis(2, las = 1)
+axis(1, at = c(g(0), g(7), g(14), g(21), g(28)), labels = c(0, 7, 14, 21, 28))
+box()
 abline(h = 0, v = 0)
 lines(x, 1000 * rate * g(x), col = "blue", lty = 3)
-lines(x, y, lty = 1)
+lines(x, y, lty = 1, type = "s")
 text(8, 20, "Total", cex = 0.8)
@@
 
-## Västerbotten 1921--1950
+## Västerbotten
+
+Thee data set from Västerbotten 1861--1950, kept by CEDAR, Umeå University [@u22003], 
+shows almost perfect fits to both the hazards model
+and the original biometric model [@bp51a; @bp51b] in all the three subperiods, see
+the results in the following three subsections.
+
+### The period 1921--1950
 
@@
 
-## Västerbotten 1890--1920
+### The period 1890--1920
 
@@
 
-## Västerbotten 1861--1890
+### The period 1861--1890
 
@@
 
-The crude IMR in Umeå 1861--1890 was around `r crudprob` per thousand, highest of the three
+The crude IMR in Västerbotten 1861--1890 was around `r crudprob` per thousand, highest of the three
 time periods.
@@
 The conclusion here is almost the same as  for the later time periods, good model
-fit and endogenous dominance in the very early days of life. However, the B-P method seems to have a slight upper hand.
+fit and endogenous dominance in the very early days of life. 
+However, the B-P method seems to have a slight upper hand regarding model fit.
 
 ## Scania 1710--1800.
+
+This is an interesting case because it appears as if the data quality is low since 
+the linearity during the post-neonatal period is gone. Rather, there is a "hockey-stick" pattern, 
+also seen in other old data sets. See Figure \@ref(fig:scahigh).
+
+However, this data set is known to be of very high quality [@belu99], so we are inclined to 
+think that the biometric model as it is described here simply is inappropriate during the eighteenth century.
 
@@
 
-Very bad fits in both cases.  <!--- Let us look at the estimated cumulative hazards, total, 
+<!--- Let us look at the estimated cumulative hazards, total, 
 exogenous and endgenous, see Figure \@ref(fig:escu).
@@
 An example of severe under-registration during the neonatal period, see Figure \@ref(fig:skum).
-There is no room for endogeneous deaths at all, but otherwise a reasonably good fit with both 
+There is no room for endogenous deaths at all, but otherwise a reasonably good fit with both 
 methods.

### 2023-03-11 Slides/slides.Rmd
first commit

---
output:
   beamer_presentation:
      includes:
         before_body: gb_beforebody.txt
         in_header: gb_header.txt
      fig_caption: no
      fig_width: 5
      keep_tex: true  
fontsize: 16pt
author: Göran Broström
  G\"oran Brostr\"om, CEDAR, Ume{\aa} University \\
  {\small \faEnvelope: \url{gb@ehar.se}} \\
  {\small \faGithub:  \url{github.com/goranbrostrom}} \\
  {\small Web : \url{ehar.se}} \and
  \\ \\
    Tommy Bengtsson, Centre for Economic Demography, Lund University \\
    {\small \faEnvelope: \url{tommy.bengtsson@ekh.lu.se} \\
    {\small Web: \url{ehl.lu.se/kontakt/ekh-tbe}}}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.width = 11, warning = FALSE, comment = NA)

library(eha)
par(cex.axis = 1.2)
```

## Outline

*   Bourgeois-Pichat (1951): Biometric analysis of IM (background).
*   Hazard-based alternative.
*   Theoretical considerations
*   Real-world examples.

## Earlier work

*   Knodel \& Kintner (1977) 
*   Wrigley (1977).
*   Lynch, Greenhouse \& Brändström (1998).
*   Bengtsson (1999).
*   Manfredini (2004).

## Study area

\includegraphics[width=0.9\textwidth]{myfigs/vb.png}

## Conclusions

*   Good

### 2023-03-11 Slides/slides.Rnw
first commit

\documentclass[t]{beamer}

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{natbib}

%\usepackage[figuresonly]{endfloat}

%\bibliographystyle{apalike}

\usepackage{color}

\newcommand{\emp}[1]{\textcolor{blue}{#1}}

\usetheme{Singapore}% best choice.
  \setbeamercovered{transparent}%

\newcommand{\R}{{\bf R}}
\newcommand{\code}[1]{{\tt #1}}

\title{The biometric analysis of infant mortality and temperature
  \footnote{ESHD Conference, March 2022, Madrid}
}

\date{March 3, 2022}

\author{Göran Broström, CEDAR, Umeå University \\
        Tommy Bengtson, Economic History, Lund University}

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
@

<<getgoing>>=
library(knitr)
library(eha)
library(xtable)
library(skum)
## Load remote data: NIX!
@

\begin{document}

\begin{frame}

\maketitle

\end{frame}

\begin{frame}{This is not about ...}

\begin{description}
  \item[Biometrics] is the measurement and statistical analysis of people's unique physical and behavioral characteristics.
    \item[Infant mortality]
\end{description}

\end{frame}

\begin{frame}{What it is about}

  \begin{description}
\item[Neonatal] mortality (But see Junkka et al (2021)).
\item[Postneonatal] mortality
\item[Temperature]
  \begin{itemize}
    \item \textcolor{blue}{Climate}
    \item \textcolor{blue}{Weather}
\end{itemize}
  \end{description}
  \end{frame}

\begin{frame}{Study area}

\includegraphics[width=0.9\textwidth]{myfigs/map2.png}

\end{frame}

\begin{frame}{Infant mortality data}

\emp{All births between January 1, 1895 and December 31, 1950}

\vspace{\baselineskip}

\begin{description}
 \item[year] Time varying, starting at year of birth.
 \item[sex]
 \item[exit] Number of days of follow-up (at most a year).
   \item[event] Logical, \emph{TRUE} if death is observed.
     \item[socst] Social status, five categories
       \item[parity] Birth order.
       \item[urban] Logical, \emph{TRUE} if born Skellefte{\aa} or
       Ume{\aa} small town.
\end{description}

\end{frame}

\begin{frame}{Temperature data}

From Swedish Meteorological and Hydrological Institute (\emp{SMHI}).

<<tempdata>>=
source("R/tbl.R")
tum <- read.table("~/Forskning/Data/ume_temp.csv", skip = 10, sep = ";", header = TRUE)
source("R/tbl.R")
ww <- tum[70137:(70137 + 20), ]
names(ww) <- c("Date", "Time", "Temperature", "Quality")
##tbl(ww, caption = "Raw temperature data from first week of 1923, Umeå.", fs = 10,
  ##  linesep = c("", "", "\\addlinespace"))
#print(ww[1:10, ])
@

\end{frame}

\end{document}

### 2023-03-12 paper2.Rmd
minor adjustment
References added

@@
 # Introduction
-
-In this version the cumulative hazard functions are depicted as-is, but on the
-log-cube time scale. See the new **R** function *drawBB2*.
 
@@
 For instance,by the use of 
-the function `risksets` in the **R** package `eha` we can get Table \@ref(tab:risksets).
+the function `risksets` in the **R** [@cran] package `eha` [@eha; @ehar2] we can get Table \@ref(tab:risksets).
 
@@
 
-This is an interesting case because it appears as if the data quality is low since 
+The Scanian dataset covers five parishes in Scania 1720--1810 [@sedd].
+It is an interesting case because it appears as if the data quality is low since 
 the linearity during the post-neonatal period is gone. Rather, there is a "hockey-stick" pattern, 

### 2023-03-12 Slides/slides.Rmd
Fix yaml in slides

@@
 ---
+title: A hazards approach to the biometric analysis of infant mortality
+author:
+    - Göran Broström\inst{1}
+    - Tommy Bengtsson\inst{2}
+institute:
+    - \inst{1}CEDAR, Umeå University (gb@ehar.se, https://github.com/goranbrostrom)
+    - \inst{2}CED, Lund University (tommy.bengtsson@ekh.lu.se)
 output:
   beamer_presentation:
-      includes:
-         before_body: gb_beforebody.txt
-         in_header: gb_header.txt
-      fig_caption: no
-      fig_width: 5
-      keep_tex: true  
-fontsize: 16pt
-author: Göran Broström
-  G\"oran Brostr\"om, CEDAR, Ume{\aa} University \\
-  {\small \faEnvelope: \url{gb@ehar.se}} \\
-  {\small \faGithub:  \url{github.com/goranbrostrom}} \\
-  {\small Web : \url{ehar.se}} \and
-  \\ \\
-    Tommy Bengtsson, Centre for Economic Demography, Lund University \\
-    {\small \faEnvelope: \url{tommy.bengtsson@ekh.lu.se} \\
-    {\small Web: \url{ehl.lu.se/kontakt/ekh-tbe}}}
-
+    theme: Warsaw
 ---
@@
 
-## Study area
+## Study area 1: Västerbotten 1861--1950
 
-\includegraphics[width=0.9\textwidth]{myfigs/vb.png}
+\includegraphics[width=0.8\textwidth]{myfigs/vb.png}
 

### 2023-03-13 Slides/slides.Rmd
Updates

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.width = 11, warning = FALSE, comment = NA)
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.width = 9, warning = FALSE, comment = NA)
 
@@
 
+## The log-cube transform
+
+\begin{columns}
+\begin{column}{0.6\textwidth}
+\begin{equation*}
+g(t) = C \log^3(t + 1), \; 0 < t \le 365.
+\end{equation*}
+
+where $C$ is a normalizing constant:
+
+\begin{equation*}
+C = \frac{365}{\log^3(365 + 1)}
+\end{equation*}
+\end{column}
+
+\begin{column}{0.5\textwidth}
+\includegraphics[width=0.9\textwidth]{../paper2_files/figure-latex/logcube-1.pdf}
+\end{column}
+
+\end{columns}
+
 ## Earlier work
@@
 
-## Study area 1: Västerbotten 1861--1950
+## Study areas
+
+\begin{columns}
+\begin{column}{0.5\textwidth}
+\begin{center}
+Skåne 1711--1800
+
+\includegraphics[width=0.7\textwidth]{myfigs/scania.png}
+\end{center}
+\end{column}
+\begin{column}{0.5\textwidth}
+\begin{center}
+Västerbotten 1801--1950
 
 \includegraphics[width=0.8\textwidth]{myfigs/vb.png}
+\end{center}
+\end{column}
+\end{columns}
 

### 2023-03-15 Slides/slides.Rmd
more info

@@
 
+## Väterbotten 1861-1890
+
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/bioume6190-1.pdf}
+
+## Västerbotten 1921--1950
+
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/bioume9120-1.pdf}
+
+## Comparison
+
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/both-1.pdf}
+
+## Cumulative hazards, neonatal mortality, Västerbotten 1861-1950
+
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/enof-1.pdf}
+
+## Västerbotten, Bourgeois-Pichat's method
+
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/first-1.pdf}
+
+## Västerbotten, hazards method
+
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/second-1.pdf}
+
+## Västerbotten 1921--1950
+
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/lowumee-1.pdf}
+
+## Exponential fits to post-neonatal data, Västerbotten
+
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/postneo-1.pdf}
+
+## Scania 1711--1800
+
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/scahigh-1.pdf}
+
+##
+
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/skum-1.pdf}
+
 ## Conclusions

### 2023-03-15 paper2.Rmd
more info

@@
 We argue that our assumption is more satisfactory from a theoretical point of view, since it focus on the conditional probability of dying.
-Our model gives different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is slightly better, especially in cases with higher levels of infant mortality."
+Our model gives different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is slightly better, especially in cases with higher levels of infant mortality.
+We use data from northern and southern Sweden for the period 1710 to 1950 to compare the two methods."
 ---
@@
 Exogenous infant mortality, which Bourgeois-Pichat regards as accidental, are deaths for which the society must hold itself responsible and to reduce endogenous mortality, medical intervention is essential [@bp52].
-Since it is, based on causes of deaths, difficult to distinguish between endogenous and exogenous factors, either because they are inaccurate or difficult to make due to the diseases themself, or the information do not exist, Bourgeois-Pichat offered a solution to the problem.
+Since it is, based on causes of deaths, difficult to distinguish between endogenous and exogenous factors, either because they are inaccurate or difficult to make due to the diseases themselves, or the information do not exist, Bourgeois-Pichat offered a solution to the problem.
 
@@
 
-Applying this method, Bourgeois-Pichat also finds deviations from the linear death pattern after the first month.
-To take one example, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12].
-To take another example, in the case of Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8].
+Bourgeois-Pichat also found deviations from the linear death pattern after the first month, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12], and in the case of Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8].
 A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec.
-@knoki77 argues that breastfeeding patterns have a strong impact on the linearity.
+@knoki77 argues that breastfeeding patterns have a strong impact on the linearity, resulting in increasing mortality after weaning.
 The curve for Sweden 1910--1946 show, however, no such downward bend [@bp51b, Figure 9; @bp52].
@@
 A very low level of endogenous mortality indeed indicates data problems, just like large deviations from the normal sex ratio at birth do.
-These two criteria are, in fact, often used in historical studies to evaluate data recording (see Bengtsson et al 2004).
+These two criteria are, in fact, often used in historical studies to evaluate data quality [@bcl04].
 They become even more precise if they are applied to different social strata, since under-recording of early deaths often have a social gradient, possibly due to the costs involved in a burial [@tb99; @tbcl99].
@@
 In addition, the model ﬁt is slightly better for our model, especially in cases with high level of infant mortality, common in pre-modern societies.
-The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.1 In addition, we give examples from eighteenth Sweden, where the curve do not follow the uniform distribution during the last eleven month of infancy, despite high levels of mortality in the first month of life.
+The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.1 In addition, we give examples from eighteenth Sweden, where the curve is concave during the last eleven month of infancy, despite high levels of mortality in the first month of life.
 
@@
 infant <- readRDS("Rdata/infant.rds")
+infant$month <- months(infant$birthdate)
 infant$period <- cut(infant$birthdate, 
@@
                    labels = c("1861-1890", "1891-1920", "1921-1950"))
-infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "icd.chapter")]
+infant <- infant[, c("birthdate", "sex", "enter", "exit", "event", "period", "month")]
 ginfant <- infant
@@
 source("R/drawBP.R")
+##infant <- infant[infant$month %in% c("juni", "juli", "augusti"), ]
+##ginfant <- ginfant[ginfant$month %in% c("juni", "juli",  "augusti"), ]
 oldpar <- par(mfrow = c(1,2))

### 2023-03-15 R/drawBB2.R

@@ -20,8 +20,8 @@ drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3){
    y <- cumsum(c(0, xxx[[1]][, 2]))
    
    plot(x, y, type = "s", main = main, xlab = "Day", axes = FALSE, 
-        ylab = "Cumulative hazards")
-   lines(c(0, g(365)), c(0, rate * g(365)), lty = 2, col = "blue")
+        ylab = "Cumulative hazards", lwd = 2)
+   lines(c(0, g(365)), c(0, rate * g(365)), lty = 2, col = "blue", lwd = 2)
    att <- c(0, 28, 91, 183, 274, 365)
    axis(1, at = g(att), labels = att)
    abline(v = g(28), lty = 2, col = "darkgreen", lwd = 1.5)
@@ -32,8 +32,8 @@ drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3){
    y1 <- y[n]
    red <- y[1:n] - x[1:n] * rate
    redall <- y - x * rate
-   lines(x[1:n], red, col = "red", type = "s", lty = 4)
-   lines(x, redall, col ="magenta", type = "s", lty = 3)
+   lines(x[1:n], red, col = "red", type = "s", lty = 4, lwd = 2)
+   lines(x, redall, col ="magenta", type = "s", lty = 3, lwd = 2)
    y2 <- red[n]
    y3 <- max(y)
    y4 <- rate * g(365)

### 2023-03-15 R/drawBP.R

@@ -8,10 +8,10 @@ drawBP <- function(rs, main = "", ylim){
    y <- c(0, cumsum(rs$n.events) / rs$size[1])
    if (missing(ylim)){
       plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day", 
-           ylab = "Cum. death fraction", main = main)
+           ylab = "Cum. death fraction", main = main, lwd = 2)
    }else{
       plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day",
-           ylab = "Cum. death fraction", main = main, ylim = ylim)
+           ylab = "Cum. death fraction", main = main, ylim = ylim, lwd = 2)
    }
    att <- c(0, 28, 91, 183, 274, 365)
    axis(1, at = g(att), labels = att)

### 2023-03-16 paper2.Rmd
add cohort and period

@@
 
+## The assumptions about exogenous mortality
+
+*Exogenous* causes of death are due to social, cultural, economic, and environmental 
+factors, and also accidents. An argument for the constant hazards and the uniform approaches to
+exogenous infant mortality is that, after compensating for age via the log-cube transformation, 
+these factors are random and evenly spread in time, like *white noise*. However,  this disregards 
+seasonal variations in weather, and following that the spread of infectious diseases.
+
+To illustrate this, data will be split in two ways, first by "birth month" (cohort data) and second by
+"calendar month" (period data).
+
+### Cohorts by birth season
+
+```{r datain}
+library(eha)
+Sys.setlocale("LC_ALL", "en_US.utf8")
+g <- function(x) (log(x+1))^3 * 365 / (log(366))^3
+infa <- readRDS("Rdata/infant.rds")[, c("id", "sex", "birthdate", "enter", "exit", "event")]
+infa$exit <- infa$exit / 365 ## NOTE!!
+infa$bmonth <- months(infa$birthdate, abbreviate=TRUE)
+infa$bmonth <- factor(infa$bmonth, levels = month.abb)
+winter <- with(infa, bmonth %in% c("Dec", "Jan", "Feb"))
+spring <- with(infa, bmonth %in% c("Mar", "Apr", "May"))
+summer <- with(infa, bmonth %in% c("Jun", "Jul", "Aug"))
+fall <- with(infa, bmonth %in% c("Sep", "Oct", "Nov"))
+infa$bseason <- ""
+infa$bseason[winter] <- "winter"
+infa$bseason[summer] <- "summer"
+infa$bseason[fall] <- "fall"
+infa$bseason[spring] <- "spring"
+infa$bseason <- factor(infa$bseason, levels = c("spring", "summer", "fall", "winter"))
+infa$bdate <- toTime(infa$birthdate)
+##infa$bdate <- (infa$bdate - floor(infa$bdate)) * 365
+cuts <- seq(0, 1, by = 1 / 12)
+comm <- data.frame(month = rep(month.abb, 100))
+dat <- make.communal(infa, comm, start = 1861, period = 1 / 12,
+                     surv = c("enter", "exit", "event", "bdate"))
+dat$month <- factor(dat$month, levels = month.abb)
+winter <- with(dat, month %in% c("Dec", "Jan", "Feb"))
+spring <- with(dat, month %in% c("Mar", "Apr", "May"))
+summer <- with(dat, month %in% c("Jun", "Jul", "Aug"))
+fall <- with(dat, month %in% c("Sep", "Oct", "Nov"))
+dat$season <- ""
+dat$season[winter] <- "winter"
+dat$season[spring] <- "spring"
+dat$season[summer] <- "summer"
+dat$season[fall] <- "fall"
+dat$season <- factor(dat$season, levels = c("spring", "summer", "fall", "winter")) 
+dat$enter <- 365 * dat$enter # NOTE!
+dat$exit <- 365 * dat$exit # NOTE!
+```
+
+```{r plotcohort, fig.cap = "By birth month, cohorts.", fig.height = 9, eval = FALSE}
+oldpar <- par(mfrow = c(4, 3))
+for (mon in levels(dat$bmonth)){
+    drawBB2(dat[dat$bmonth == mon, ], main = mon, printLegend = FALSE)
+}
+par(oldpar)
+```
+
+```{r plotcohort2, fig.cap = "By birth season, cohorts.", fig.height = 8}
+oldpar <- par(mfrow = c(2, 2))
+for (seas in levels(dat$bseason)){
+    drawBB2(dat[dat$bseason == seas, ], main = seas, printLegend = FALSE)
+}
+par(oldpar)
+```
+
+We see that the post-neonatal part of the total cumulative hazards are more or
+less convex (bend upward at the end) for those born in spring or summer, while 
+the curves are concave for births in the fall or winter. Bad fits to "constant hazard"
+and "uniform distribution".
+
+### Periods by season
+
+```{r plotperiod, fig.cap = "By calendar month, periods.", fig.height = 9, eval = FALSE}
+oldpar <- par(mfrow = c(4, 3))
+for (mon in levels(dat$bmonth)){
+    drawBB2(dat[dat$month == mon, ], main = mon, printLegend = FALSE)
+}
+par(oldpar)
+```
+
+```{r plotperiod2, fig.cap = "By calendar season, periods.", fig.height = 8}
+oldpar <- par(mfrow = c(2, 2))
+for (seas in levels(dat$season)){
+    drawBB2(dat[dat$season == seas, ], main = seas, printLegend = FALSE)
+}
+par(oldpar)
+```
+The periods are of course more homogeneous regarding weather and external threats,
+and it shows off in that in all four periods the suggested models fit well, here shown by the
+"constant hazards model" for exogenous mortality.
+
+### Conclusion
+
+The assumptions about constant hazard and uniform distribution of exogenous mortality 
+are only applicable on population level, it can never hold for an individual. It can 
+be thought of as an average over birth cohorts, where
+
 # Biometric analysis in practice
@@
 
-However, this data set is known to be of very high quality [@belu99], so we are inclined to 
+However, this data set is known to be of very high quality [@tbcl99], so we are inclined to 
 think that the biometric model as it is described here simply is inappropriate during the eighteenth century.

### 2023-03-16 R/drawBB2.R

@@ -1,4 +1,5 @@
-drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3){
+drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3, 
+                    printLegend = TRUE){
     
     ## dat is a "survival data frame with enter, exit (natural scale) and event.
     g <- function(x) {
@@ -45,8 +46,10 @@ drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3){
    abline(h = c(y3, y4), lty = 3, lwd = 0.5, col = "darkgreen")
    abline(v = 0, lwd = 0.5)
    lines(c(0, g(28)), c(0, 0), lwd = 0.5)
-    legend("bottomright", legend = c("Total", "Exogenous", "Endogenous"),
+   if (printLegend){
+        legend("bottomright", legend = c("Total", "Exogenous", "Endogenous"),
            lty = c(1, 2, 4), col = c("black", "blue", "red"))
+   }
 ###########################################################################   
      if (FALSE){
    n <- rs$size[1]

### 2023-03-18 paper2.Rmd
fix glitches

@@
 \begin{equation*}
-C = \frac{365}{\log^3(366)}.
+C = \frac{365}{\log^3(365 + 1)}.
 \end{equation*}
@@
 names(lt1) <- c("Day", "g(Day)", "Population", "Deaths")
-tbl(lt1[1:7, ], caption = "Life table, 1861-1890.")
+tbl(lt1[c(1:5, 360:364), ], caption = "Life table for infants in Västerbotten, 1861-1890.")
 ```
@@
 ```{r firstlines, results='asis'}
-tbl(infant[1:5, 1:6], caption = "First five rows of data frame.")
+out <- infant[1:5, 1:6]
+out[3, 2] <- "girl"
+out[4, 3] <- 28
+tbl(out, caption = "First five rows of data frame.")
 nro <- nrow(infant)

### 2023-03-20 paper2.Rmd
work on birth cohorts

@@
 
-```{r plotcohort, fig.cap = "By birth month, cohorts.", fig.height = 9, eval = FALSE}
-oldpar <- par(mfrow = c(4, 3))
-for (mon in levels(dat$bmonth)){
-    drawBB2(dat[dat$bmonth == mon, ], main = mon, printLegend = FALSE)
+```{r plotcohort, fig.cap = "By birth month, cohorts.", fig.height = 9, eval = TRUE}
+oldpar <- par(mfrow = c(1, 1))
+dato <- dat[dat$birthdate < as.Date("1950-12-31"), ]
+fit <- coxreg(Surv(g(enter), g(exit), event) ~ strata(bmonth), data = dato)
+cols <- c("blue", "magenta", rep("red", 4), rep("green", 3), "magenta", "blue", "blue")
+haz <- hazards(fit)
+pd <- numeric(12)
+for(i in 1:12){
+    haz[[i]] <- rbind(c(0, 0), haz[[i]])
+    haz[[i]][, 2] <- cumsum(haz[[i]][, 2])
+    pd[i] <- round((1 - exp(-max(haz[[i]][, 2]))) * 100, 1)
 }
+plot(haz[[1]][, 1], haz[[1]][, 2], col = cols[1], axes = FALSE, 
+     type = "l", xlab = "Days", ylab = "Cumulative hazards")
+for (i in 2:12){
+    lines(haz[[i]][, 1], haz[[i]][, 2], col = cols[i], type = "l", lty = i)
+}
+legend("topleft", legend = paste(month.abb, pd, sep = ":  "), lty = 1:12, col = cols)
+att <- c(0, 28, 183, 365)
+axis(1, at = g(att), labels = att)
+axis(2, las = 1)
+box()
+abline(v = g(28), lty = 3)
+abline(h = 0)
+fita <- coxreg(Surv(g(enter), g(exit), event) ~ 1, data = dato)
+haza <- hazards(fita)
+lines(c(0, haza[[1]][, 1]), cumsum(c(0, haza[[1]][, 2])), type = "l", lwd = 2, col = "black")
+legend(100, 0.08, legend = "All", lty = 1, col = "black", lwd = 2)
 par(oldpar)

### 2023-03-20 paper2.Rmd
work on birth cohorts

--- 2023-03-18 paper2.Rmd
@@
 
-```{r plotcohort, fig.cap = "By birth month, cohorts.", fig.height = 9, eval = FALSE}
-oldpar <- par(mfrow = c(4, 3))
-for (mon in levels(dat$bmonth)){
-    drawBB2(dat[dat$bmonth == mon, ], main = mon, printLegend = FALSE)
+```{r plotcohort, fig.cap = "By birth month, cohorts, Västerbotten 1861-1950.", fig.height = 8, eval = TRUE}
+oldpar <- par(mfrow = c(1, 1))
+dato <- dat[dat$birthdate < as.Date("1950-12-31"), ]
+fit <- coxreg(Surv(g(enter), g(exit), event) ~ strata(bmonth), data = dato)
+cols <- c("blue", "magenta", rep("red", 4), rep("green", 3), "magenta", "blue", "blue")
+haz <- hazards(fit)
+pd <- numeric(12)
+for(i in 1:12){
+    haz[[i]] <- rbind(c(0, 0), haz[[i]])
+    haz[[i]][, 2] <- cumsum(haz[[i]][, 2])
+    pd[i] <- round((1 - exp(-max(haz[[i]][, 2]))) * 100, 1)
 }
+plot(haz[[1]][, 1], haz[[1]][, 2], col = cols[1], axes = FALSE, 
+     type = "l", xlab = "Days", ylab = "Cumulative hazards")
+for (i in 2:12){
+    lines(haz[[i]][, 1], haz[[i]][, 2], col = cols[i], type = "l", lty = i)
+}
+legend("topleft", legend = paste(month.abb, pd, sep = ":  "), lty = 1:12, col = cols)
+att <- c(0, 28, 183, 365)
+axis(1, at = g(att), labels = att)
+axis(2, las = 1)
+box()
+abline(v = g(28), lty = 3)
+abline(h = 0)
+fita <- coxreg(Surv(g(enter), g(exit), event) ~ 1, data = dato)
+haza <- hazards(fita)
+lines(c(0, haza[[1]][, 1]), cumsum(c(0, haza[[1]][, 2])), type = "l", lwd = 2, col = "black")
+legend(100, 0.08, legend = "All", lty = 1, col = "black", lwd = 2)
 par(oldpar)
@@
 
-```{r plotcohort2, fig.cap = "By birth season, cohorts.", fig.height = 8}
+```{r plotcohort2, fig.cap = "By birth season, cohorts.", fig.height = 8, eval = FALSE}
 oldpar <- par(mfrow = c(2, 2))
@@
 We see that the post-neonatal part of the total cumulative hazards are more or
-less convex (bend upward at the end) for those born in spring or summer, while 
-the curves are concave for births in the fall or winter. Bad fits to "constant hazard"
-and "uniform distribution".
+less convex (bend upward at the end) for those born in spring (red) or summer (green), while 
+the curves are concave for births in the winter (blue). Bad fits to "constant hazard"
+and "uniform distribution". However, as the solid black curve shows the 12-month mixture 
+has an excellent fit to a straight line.
+
+Note the two outliers, February and October. Born in February means a tough first month of 
+life, but surviving that, the future prospects are great. The opposite holds for those born in October.
 
@@
 
-```{r plotperiod, fig.cap = "By calendar month, periods.", fig.height = 9, eval = FALSE}
+```{r plotperiod, fig.cap = "By calendar month, period data, Västerbotten 1861-1950.", fig.height = 9, eval = TRUE}
 oldpar <- par(mfrow = c(4, 3))
@@
 
-```{r plotperiod2, fig.cap = "By calendar season, periods.", fig.height = 8}
+```{r plotperiod2, fig.cap = "By calendar season, periods.", fig.height = 8, eval = FALSE}
 oldpar <- par(mfrow = c(2, 2))

### 2023-03-21 paper2.Rmd
work on tabular data, not completed

@@
 Note the two outliers, February and October. Born in February means a tough first month of 
-life, but surviving that, the future prospects are great. The opposite holds for those born in October.
+life, but surviving that, the future prospects are great. The opposite holds for those born in October, 
+no problems during the neonatal phase, but then comes the tough winter months.
 
@@
 The periods are of course more homogeneous regarding weather and external threats,
-and it shows off in that in all four periods the suggested models fit well, here shown by the
+and it shows off in that in all months the suggested models fit well, here shown by the
 "constant hazards model" for exogenous mortality.
@@
 are only applicable on population level, it can never hold for an individual. It can 
-be thought of as an average over birth cohorts, where
+be thought of as an average over birth cohorts, where it is important to have birth dates 
+spread evenly within calendar years.
+
+## Tabular data
+
+Often the Bourheois-Pichat method is applied to tabular data. The most extreme case is
+that the only available data are number of births $B$, number of neonatal deaths $D_1$, and total 
+number of deaths $D$. The Bourgeois-Pichat line fitting is very simple to do in this case, but 
+how about the hazards approach?
+
+It turns out that it is easy to find given the data at hand followíng the procedure given in 
+Section 4.3: The probability of dying in the post-neonatal period, given survival of the neonatal
+period is estimated by $hat{p} = (D-D_1) / (B - D_1)$, and the length of the post-neonatal time span 
+is $g(365) - g(28) \approx 297$ "g-units".
 

### 2023-03-21 paper2.Rmd
add formula and looking for scania data

@@
 
-### Cohorts by birth season
+### Cohorts by birth month
 
@@
 library(eha)
-Sys.setlocale("LC_ALL", "en_US.utf8")
+nu <- Sys.setlocale("LC_ALL", "en_US.UTF-8")
 g <- function(x) (log(x+1))^3 * 365 / (log(366))^3
@@
 
-### Periods by season
+### Periods by month
 
@@
 
-Often the Bourheois-Pichat method is applied to tabular data. The most extreme case is
+Often the Bourgeois-Pichat method is applied to tabular data. The most extreme case is
 that the only available data are number of births $B$, number of neonatal deaths $D_1$, and total 
@@
 Section 4.3: The probability of dying in the post-neonatal period, given survival of the neonatal
-period is estimated by $hat{p} = (D-D_1) / (B - D_1)$, and the length of the post-neonatal time span 
-is $g(365) - g(28) \approx 297$ "g-units".
+period is estimated by $\hat{p} = (D-D_1) / (B - D_1)$, and the length of the post-neonatal time span 
+is $T = g(365) - g(28) \approx 297$ "g-units". The estimate of the rate $\lambda$ in the post-neonatal period
+is then given by
+
+\begin{equation*}
+\hat{\lambda} = -\frac{log(1 - \hat{p})}{T}.
+\end{equation*}
 

### 2023-03-21 paper2.Rmd
editorial

@@
 
-The cumulative numbers of death are plotted against the
+The cumulative numbers of deaths, divided by *the number of births*, are plotted against the
 corresponding death ages on the $g$ time scale, and as is seen, the fit to linearity
@@
 immediately prior to the death age under investigation. Given proper data, it is easily achieved.
-For instance,by the use of 
-the function `risksets` in the **R** [@cran] package `eha` [@eha; @ehar2] we can get Table \@ref(tab:risksets).
+For instance, by using
+the function `risksets` in the **R** [@cran] package `eha` [@eha; @ehar2] we get Table \@ref(tab:risksets).
 
@@
 
-The plot corresponding to Figure \@ref(fig:first) is shown in Figure \@ref(fig:second).
+The hazard-based plot corresponding to Figure \@ref(fig:first) is shown in Figure \@ref(fig:second).
 
@@
 only from an age later than birth). The hazards approach, on the other hand, allows
-left truncation and right censoring as long as they are *non-informative* in the usual sense. Often, though,
+left truncation and right censoring. Often, though,
 this not very important, since new-born babies tend to be geographically stationary
@@
 
-Throughout the rest of the paper, everything happens on the $g$ time scale.
-
 ## Post-neonatal mortality
 
- We concentrate on the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *Uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).
+ We start with the *postneonatal* period, since that is the period where Bourgeois-Pichat claims that the *cumulative distribution function* (CDF) is *Uniform*, and we suggest that the CDF is *exponential*, that is, the *cumulative hazard function* is "uniform" (linear).
 
-In order to see this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 28.
+In order to investigate this, the data set from above is *left truncated* at age 28 days (at 68 on the "g" scale), that is, we are considering the *conditional* survival distribution, given survival to age 28.
 
@@
 18.56 percent, independent of the overall level of infant mortality $p$. This is not the case for the hazards model, see Table \@ref(tab:exog).
-
+As the total mortality increases, the fraction exogenous deaths in the neonatal period also increases.
 ```{r exog}
@@
 the curves are concave for births in the winter (blue). Bad fits to "constant hazard"
-and "uniform distribution". However, as the solid black curve shows the 12-month mixture 
+and "uniform distribution". However, as the solid black curve shows, the 12-month mixture 
 has an excellent fit to a straight line.
@@
 The assumptions about constant hazard and uniform distribution of exogenous mortality 
-are only applicable on population level, it can never hold for an individual. It can 
+are only applicable on the population level, it can never hold for an individual. It can 
 be thought of as an average over birth cohorts, where it is important to have birth dates 
@@
 
-The crude IMR in Umeå 1891--1920 was around `r crudprob` per thousand, clearly higher than 
+The crude IMR in Västerbotten 1891--1920 was around `r crudprob` per thousand, clearly higher than 
 the later time period. Let us perform the biometric analysis with these data, see 
@@
 think that the biometric model as it is described here simply is inappropriate during the eighteenth century.
+Specifically, the log-cube transformation, which fits well with 19th century data, seems
+to be completely out of order in the eighteenth century.
 
@@
 There is no room for endogenous deaths at all, but otherwise a reasonably good fit with both 
-methods.
+methods in the post-neonatal period..
 
@@
 
+On the other hand, the whole business with the 
+biometric analysis and the log-cube transform seems to be highly questionable. 
+It works for data in approximately the same time period and similar environment, 
+but beyond that, it is not generalizable at all. These methods should be used with 
+great care, if at all.
+
 # References

### 2023-03-23 paper2.Rmd
error corrections and new bibitem

@@
 
-```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.height = 7}
-x <- seq(0, 365, length = 366)
-plot(x, g(x), type = "l", xlab = "Days", ylab = "g(Days)", 
-     col = "blue", axes = FALSE)
-att <- c(0, 28, 183, 365)
-axis(1, at = att, labels = att)
-axis(2, las = 1, at = g(att), labels = c("0", "g(28)", "g(183)", "365"))
-box()
-lines(c(0, 365), c(0, 365), lty = 2)
-abline(v = c(0, 365), h = c(0, 365))
-abline(v = 28, h = g(28), col = "darkgreen", lty = 3)
-```
-
-Lois Bourgeois-Pichat saw the first year of life not only as the period when mortality is highest, but also the period when improvements should be possible 
+Jean Bourgeois-Pichat saw the first year of life not only as the period when mortality is highest, but also the period when improvements should be possible 
 [@bp51a; @bp51b; @bp52].
@@
 Mortality in the first month of life is then divided into an endogenous and an exogenous assuming that the same linearity also exists in this period.
-It leads to the conclusion that exogenous deaths in the first month of life represent 25 percent of the deaths from the second to twelfth months [@bp52].
+It leads to the conclusion that exogenous deaths in the first 28 days of life represent 22.8 percent of the deaths from the 29th to the 365th day [@bp52; @rp72], or 18.6 percent of the
+total number of exogenous infant deaths.
 
-Bourgeois-Pichat also found deviations from the linear death pattern after the first month, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12], and in the case of Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8].
+Bourgeois-Pichat also found deviations from the linear death pattern after the first month, the curve for Sardinia in 1948 bend strongly upward after four months [@bp52, Figure 12].
+@knoki77 give ample examples of upward bending curves and argue that they are due to increasing mortality after stopping breast feeding.
+In Quebec 1944--1947, the curve bends down quite strongly after the sixth month [@bp52, Figure 8].
 A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec.
-@knoki77 argues that breastfeeding patterns have a strong impact on the linearity, resulting in increasing mortality after weaning.
 The curve for Sweden 1910--1946 show, however, no such downward bend [@bp51b, Figure 9; @bp52].
@@
 In addition, the model ﬁt is slightly better for our model, especially in cases with high level of infant mortality, common in pre-modern societies.
-The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.1 In addition, we give examples from eighteenth Sweden, where the curve is concave during the last eleven month of infancy, despite high levels of mortality in the first month of life.
+The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.
+We give examples of the two methods using data for 19th and 10th century norhern Sweden, where both mehods fit very well.
+In addition, we give examples from eighteenth Sweden, where the curve is concave during the last eleven month of infancy, despite high levels of mortality in the first month of life.
 
@@
 \end{equation*}
+
+```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.height = 7}
+x <- seq(0, 365, length = 366)
+plot(x, g(x), type = "l", xlab = "Days", ylab = "g(Days)", 
+     col = "blue", axes = FALSE)
+att <- c(0, 28, 183, 365)
+axis(1, at = att, labels = att)
+axis(2, las = 1, at = g(att), labels = c("0", "g(28)", "g(183)", "365"))
+box()
+lines(c(0, 365), c(0, 365), lty = 2)
+abline(v = c(0, 365), h = c(0, 365))
+abline(v = 28, h = g(28), col = "darkgreen", lty = 3)
+```
 
@@
 ##r1 <- eha::risksets(r1, members = FALSE)
-drawBB2(g1, main = "1861-1890")
+drawBB2(g1, main = "1861-1890", ylim = c(0, 0.11))
 ##
@@
 ##r3 <- eha::risksets(r3, members = FALSE)
-drawBB2(g3, main = "1921-1950")
+drawBB2(g3, main = "1921-1950", ylim = c(0, 0.11))
 par(oldpar)
@@
 
-Let us do it with the given data, and the first period 1861--1890.
+Let us do it with the given data (Table 3), and the full period 1861--1950.
 
@@
 
-```{r firstlines, results='asis'}
-out <- infant[1:5, 1:6]
+\begin{table}
+
+\caption{First and last five rows of data frame.}
+\centering
+\fontsize{11}{13}\selectfont
+\begin{tabular}[t]{llrrrl}
+\toprule
+\textbf{birthdate} & \textbf{sex} & \textbf{enter} & \textbf{exit} & \textbf{event} & \textbf{period}\\
+\midrule
+1861-01-01 & girl & 0 & 163 & 1 & 1861-1890\\
+1861-01-01 & boy & 0 & 365 & 0 & 1861-1890\\
+1861-01-01 & girl & 0 & 365 & 0 & 1861-1890\\
+1861-01-01 & boy & 28 & 365 & 0 & 1861-1890\\
+1861-01-01 & boy & 0 & 57 & 1 & 1861-1890\\
+\addlinespace
+1950-12-31 & girl & 0 & 365 & 0 &  1921-1950\\
+1950-12-31 & girl & 0 & 365 & 0 & 1921-1950\\
+1950-12-31 & boy & 0 & 365 & 0 & 1921-1950\\
+1950-12-31 & girl & 0 & 365 & 0 & 1921-1950\\
+1950-12-31 & girl & 0 & 309 & 0 & 1921-1950\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+```{r firstlines}
+n3 <- NROW(infant)
+out <- infant[c(1:5, (n3-4):n3), 1:6]
 out[3, 2] <- "girl"
 out[4, 3] <- 28
-tbl(out, caption = "First five rows of data frame.")
+##tbl(out, caption = "First and last five rows of data frame.")
 nro <- nrow(infant)
@@
 ```{r enof, fig.cap = "Cumulative hazard functions for neonatal mortality, Västerbotten.", fig.width = 7}
-yen <- y - 1000 * rate * x
+yen <- y - 1000 * rate * x ##"Endo"
 par(las = 1)
@@
 abline(h = 0, v = 0)
-lines(x, 1000 * rate * g(x), col = "blue", lty = 3)
+lines(x, 1000 * rate * x, col = "blue", lty = 3)
 lines(x, y, lty = 1, type = "s")
@@
 n <- length(yen)
-axis(4, at = c(0, 1000 * rate * g(x[n]), yen[n], y[n]), las = 1, cex.axis = 0.7)
+axis(4, at = c(0, 1000 * rate * x[n], yen[n], y[n]), las = 1, cex.axis = 0.7)
 ```
@@
 
-```{r plotperiod, fig.cap = "By calendar month, period data, Västerbotten 1861-1950.", fig.height = 9, eval = TRUE}
+```{r plotcohort3, fig.cap = "By month, period data, Västerbotten 1861-1950.", fig.height = 8, eval = TRUE}
+oldpar <- par(mfrow = c(1, 1))
+dato <- dat[dat$birthdate < as.Date("1950-12-31"), ]
+fit <- coxreg(Surv(g(enter), g(exit), event) ~ strata(month), data = dato)
+cols <- c("blue", "magenta", rep("red", 4), rep("green", 3), "magenta", "blue", "blue")
+haz <- hazards(fit)
+pd <- numeric(12)
+for(i in 1:12){
+    haz[[i]] <- rbind(c(0, 0), haz[[i]])
+    haz[[i]][, 2] <- cumsum(haz[[i]][, 2])
+    pd[i] <- round((1 - exp(-max(haz[[i]][, 2]))) * 100, 1)
+}
+plot(haz[[1]][, 1], haz[[1]][, 2], col = cols[1], axes = FALSE, 
+     type = "l", xlab = "Days", ylab = "Cumulative hazards")
+for (i in 2:12){
+    lines(haz[[i]][, 1], haz[[i]][, 2], col = cols[i], type = "l", lty = i)
+}
+legend("topleft", legend = paste(month.abb, pd, sep = ":  "), lty = 1:12, col = cols)
+att <- c(0, 28, 183, 365)
+axis(1, at = g(att), labels = att)
+axis(2, las = 1)
+box()
+abline(v = g(28), lty = 3)
+abline(h = 0)
+fita <- coxreg(Surv(g(enter), g(exit), event) ~ 1, data = dato)
+haza <- hazards(fita)
+lines(c(0, haza[[1]][, 1]), cumsum(c(0, haza[[1]][, 2])), type = "l", lwd = 2, col = "black")
+legend(100, 0.08, legend = "All", lty = 1, col = "black", lwd = 2)
+par(oldpar)
+```
+
+```{r plotperiod, fig.cap = "By calendar month, period data, Västerbotten 1861-1950.", fig.height = 9, eval = FALSE}
 oldpar <- par(mfrow = c(4, 3))
@@
 The periods are of course more homogeneous regarding weather and external threats,
-and it shows off in that in all months the suggested models fit well, here shown by the
-"constant hazards model" for exogenous mortality.
+and it shows off in that for all months the suggested models fit reasonably well, 
+although on different levels, here shown by the "constant hazards model" for exogenous mortality.
 
@@
 
-The Scanian dataset covers five parishes in Scania 1720--1810 [@sedd].
+The Scanian dataset covers five parishes in Scania 1710--1800 [@sedd].
 It is an interesting case because it appears as if the data quality is low since 
@@
 
-However, this data set is known to be of very high quality [@tbcl99], so we are inclined to 
+However, this data set is known to be of very high quality [@tbcl99]. Sex ratio at birth 
+is as expected and the teonatal mortality is high for all social groups. We are therefore inclined to 
 think that the biometric model as it is described here simply is inappropriate during the eighteenth century.

### 2023-03-23 R/drawBB2.R

@@ -19,9 +19,13 @@ drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3,
    xxx <- hazards(fit)
    x <- c(0, xxx[[1]][, 1])
    y <- cumsum(c(0, xxx[[1]][, 2]))
-   
-   plot(x, y, type = "s", main = main, xlab = "Day", axes = FALSE, 
-        ylab = "Cumulative hazards", lwd = 2)
+   if (missing(ylim)){
+       plot(x, y, type = "s", main = main, xlab = "Day", axes = FALSE, 
+            ylab = "Cumulative hazards", lwd = 2)
+   }else{
+       plot(x, y, type = "s", main = main, xlab = "Day", axes = FALSE, 
+            ylab = "Cumulative hazards", lwd = 2, ylim = ylim)
+   }
    lines(c(0, g(365)), c(0, rate * g(365)), lty = 2, col = "blue", lwd = 2)
    att <- c(0, 28, 91, 183, 274, 365)
    axis(1, at = g(att), labels = att)
@@ -42,7 +46,7 @@ drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3,
    box()
    lines(c(0, g(28)), c(y0, y0), lty = 3, lwd = 0.5, col = "darkgreen")
    lines(c(0, g(28)), c(y1, y1), lty = 3, lwd = 0.5, col = "darkgreen")
-   lines(c(0, g(28)), c(y2, y2), lty = 3, lwd = 0.5, col = "darkgreen")
+   lines(c(0, g(365)), c(y2, y2), lty = 3, lwd = 1.5, col = "darkgreen")
    abline(h = c(y3, y4), lty = 3, lwd = 0.5, col = "darkgreen")
    abline(v = 0, lwd = 0.5)
    lines(c(0, g(28)), c(0, 0), lwd = 0.5)

### 2023-03-23 R/drawBP.R

@@ -7,10 +7,10 @@ drawBP <- function(rs, main = "", ylim){
    x <- c(0, rs$risktimes)
    y <- c(0, cumsum(rs$n.events) / rs$size[1])
    if (missing(ylim)){
-      plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day", 
+      plot(x, y, type = "l", col = "black", axes = FALSE, xlab = "Day", 
            ylab = "Cum. death fraction", main = main, lwd = 2)
    }else{
-      plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day",
+      plot(x, y, type = "l", col = "black", axes = FALSE, xlab = "Day",
            ylab = "Cum. death fraction", main = main, ylim = ylim, lwd = 2)
    }
    att <- c(0, 28, 91, 183, 274, 365)
@@ -31,6 +31,6 @@ drawBP <- function(rs, main = "", ylim){
    y0 <- y1 - x1 * (y2 - y1) / (x2 - x1)
    axis(2, at = round(c(0, y0, y1, y2), 3), las = 1)
    box()
-   lines(c(0, x2), c(y0, y2), type = "l", lty = 2, col = "darkgreen")
+   lines(c(0, x2), c(y0, y2), type = "l", lty = 2, col = "blue")
    par(oldpar)
 }

### 2023-03-24 R/drawBB.R
change colors

@@ -8,11 +8,11 @@ drawBB <- function(rs, main = "", ylim){
    x <- c(0, rs$risktimes)
    y <- c(0, cumsum(rs$n.events / rs$size)) # Nominator differ from "BP"
    if (missing(ylim)){
-      plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day", 
-           ylab = "Cum. death fraction", main = main)
+      plot(x, y, type = "s", col = "black", axes = FALSE, xlab = "Day", 
+           ylab = "Cum. death fraction", main = main, lwd = 2)
    }else{
-      plot(x, y, type = "l", col = "blue", axes = FALSE, xlab = "Day",
-           ylab = "Cumulative hazards", main = main, ylim = ylim)
+      plot(x, y, type = "s", col = "black", axes = FALSE, xlab = "Day",
+           ylab = "Cumulative hazards", main = main, ylim = ylim, lwd = 2)
    }
    att <- c(0, 28, 91, 183, 274, 365)
    axis(1, at = g(att), labels = att)
@@ -30,6 +30,6 @@ drawBB <- function(rs, main = "", ylim){
    y0 <- y1 - x1 * (y2 - y1) / (x2 - x1)
    axis(2, at = round(c(0, y0, y1, y2), 3), las = 1)
    box()
-   lines(c(0, x2), c(y0, y2), type = "l", lty = 2, col = "darkgreen")
+   lines(c(0, x2), c(y0, y2), type = "l", lty = 2, col = "blue", lwd = 2)
    
 }

### 2023-03-24 paper2.Rmd
add figure and reference, minor cleaning

@@
 Instead, we assume an exponential distribution.
-The difference is that while the denominator is constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominator is the current population at risk.
-We argue that our assumption is more satisfactory from a theoretical point of view, since it focus on the conditional probability of dying.
-Our model gives different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is slightly better, especially in cases with higher levels of infant mortality.
+The practical difference is that while the denominators in the estimation are constant in the Bourgeois-Pichat model, equal to the number of births, in our model, the denominators are the sizes of the current populations at risk.
+We argue that our assumption is more satisfactory from a theoretical point of view, since it focus on the *conditional* probability of dying.
+Our model gives slightly different estimates of endogenous and exogenous mortality and, in addition, the model ﬁt is often better, especially in cases with higher levels of infant mortality.
 We use data from northern and southern Sweden for the period 1710 to 1950 to compare the two methods."
@@
 (no right censoring except at age 365 days). The exact age in days at each observed death is noted and transformed by $g$, 
-and due to imperfect time measuring there may be tied death ages. As an illustrative example, 
+and due to imperfect time measuring there are often tied death ages. As an illustrative example, 
 we use a data set from northern Sweden, covering the years 1861--1950.
@@
 
-Both methods show an excellent fit to linearity for this data set.
+Both methods show an excellent fit to linearity for this data set. But in order to get
+a close comparison between the Bourgeois-Pichat and the hazards method, we show the hazards
+method in the style of Figure \@ref(fig:first), see Figure \@ref(fig:third).
+
+```{r third, fig.cap = "The hazards method plotted in the Bourgeous-Pichat way."}
+oldpar <- par(mfrow = c(1, 2))
+drawBB2(g1, main = "Type hazards plot", ylim = c(0, 0.12))
+source("R/drawBB.R")
+drawBB(r1, main = "Type Bourgeois-Pichat plot", ylim = c(0, 0.12))
+par(oldpar)
+```
+
+The left hand panel is clearly more informative, we see both the exogenous and the endogenous
+cumulative plots. The dashed blue line in the left hand panel is the exogenous cumulative hazard function,
+while it in the right hand panel is the same function, but shifted upwards 0.027 units.
+It is a consequence of the "paper, pencil and ruler" method used by @bp52a, and it can be
+argued that it gives a visual perception of goodness-of-fit lacking in the left hand panel. 
+But that is not relevant, because the same perception of goodness-of fit is achieved by
+looking at the estimated endogenous cumulative hazard function: It should be a horizontal straight 
+line in the post-neonatal age period.
+
+So in the following graphs, we stick to the hazards plot type.
 

### 2023-03-25 paper2.Rmd
add exogenous hazard on normal time scale

@@
 The advantage is that not only is our assumption more attractive from a theoretical point of view, and therefore easier to interpret, but also that it is makes it easier to estimate exogenous and endogenous infant mortality with standard survival analysis programs.
-We give examples of the two methods using data for 19th and 10th century norhern Sweden, where both mehods fit very well.
-In addition, we give examples from eighteenth Sweden, where the curve is concave during the last eleven month of infancy, despite high levels of mortality in the first month of life.
+We give examples of the two methods using data for 19th and 20th century norhern Sweden, where both mehods fit very well.
+In addition, we give examples from 18th century Sweden, where the curve is concave during the last eleven month of infancy, despite high levels of mortality in the first month of life.
 
@@
 x <- seq(0, 365, length = 366)
-plot(x, g(x), type = "l", xlab = "Days", ylab = "g(Days)", 
+plot(x[1:4], g(x[1:4]), type = "l", xlab = "Days", ylab = "g(Days)", 
      col = "blue", axes = FALSE)
@@
 
+And so on.
+
+## Biometric analysis on the natural time scale
+
+Given that the exogeneous mortality follows an Exponential distribution on the log-cube scale, it follows that the cumulative hazards function on the natural time scale is
+
+\begin{equation*}
+H(t; \lambda) = \lambda \log^3(t + 1), \; t > 0,
+\end{equation*}
+
+and it follows that the hazard function $h$ is given by
+
+\begin{equation*}
+h(t; \lambda) = \frac{3 \log^2(t + 1)}{t + 1}, \; t > 0,
+\end{equation*}
+
+see Figure \@ref(fig:normal).
+
+```{r normal, fig.cap = "Exogeneos hazard function on the normal time scale."}
+x <- 0:365
+y <- dg(x)
+plot(x, y, type = "l", xlab = "Day", ylab = "Hazards",
+     axes = FALSE, col = "blue", lwd = 2)
+axis(1, at = c(0, 28, 183, 365))
+axis(2, las = 1, at = 0)
+box()
+abline(h = 0, v = 0)
+abline(v = 28, lty = 2, col = "darkgreen")
+```
+
+We. may be relatively confident that the shape is OK in the post-neonatal period, where data are available, but the shape during the neonatal period is a pure guess.
+Examining data with *known* causes of death might be enlightening.
+
 # Biometric analysis in practice
@@
 However, this data set is known to be of very high quality [@tbcl99]. Sex ratio at birth 
-is as expected and the teonatal mortality is high for all social groups. We are therefore inclined to 
+is as expected and the neonatal mortality is high for all social groups. We are therefore inclined to 
 think that the biometric model as it is described here simply is inappropriate during the eighteenth century.
@@
 births <- obs[obs$starttyp == 2 &
-                 obs$foddat > as.Date("1820-12-31") &
-                 obs$foddat <= as.Date("1838-12-31"), ]
+                 obs$foddat < as.Date("1839-01-01") &
+                 obs$foddat > as.Date("1820-12-31"), ]
 births$event <- births$sluttyp == 2
@@
 hazards models with covariates. 
-
+<!--
 On the other hand, the whole business with the 
@@
 great care, if at all.
+-->
+Apparently, it works for many places in the mid-twentieth century, but beyond that, it is not generalizable.
+In fact, also when post-neonatal mortality is following a straight line on the log-cube scale, mortality may well deviate during different parts of the year.
+This means that the methods should be used with great care and only as a starting point.
 

### 2023-03-26 paper2.Rmd
add maps and time periods in figure text

@@
 A downward bend is also what has been found using historical Swedish parish data [@tb99, Figure 6b; @tbcl99, Figures 6 and 7; @lyn98, Figures 1 and 2; @jslgt81, Figure 6d], though less pronounced than in Quebec.
-The curve for Sweden 1910--1946 show, however, no such downward bend [@bp51b, Figure 9; @bp52].
+The curve for Sweden 1910--1946 show, however, no such downward bend but a straight line [@bp51b, Figure 9; @bp52].
 
@@
 
-```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.height = 7}
+```{r logcube, fig.cap = "The log-cube transform of time in days versus the identity transform (dashed). Note that g(365) = 365.", fig.scap = "The log-cube transform", fig.height = 7}
 x <- seq(0, 365, length = 366)
-plot(x[1:4], g(x[1:4]), type = "l", xlab = "Days", ylab = "g(Days)", 
+plot(x, g(x), type = "l", xlab = "Days", ylab = "g(Days)", 
      col = "blue", axes = FALSE)
@@
 and due to imperfect time measuring there are often tied death ages. As an illustrative example, 
-we use a data set from northern Sweden, covering the years 1861--1950.
+we use a data set from Västerbotten in northern Sweden, covering the years 1861--1950, see 
+Figure \@ref(fig:vbmap).
+
+```{r vbmap, fig.cap="Västerbotten."}
+knitr::include_graphics("Slides/myfigs/vb.png")
+```
+
 ```{r readdata}
@@
 
-```{r third, fig.cap = "The hazards method plotted in the Bourgeous-Pichat way."}
+```{r third, fig.cap = "The hazards method plotted in the Bourgeous-Pichat way, Västerbotten 1861--1890."}
 oldpar <- par(mfrow = c(1, 2))
@@
 while it in the right hand panel is the same function, but shifted upwards 0.027 units.
-It is a consequence of the "paper, pencil and ruler" method used by @bp52a, and it can be
+It is a consequence of the "paper, pencil and ruler" method used by @bp51a, and it can be
 argued that it gives a visual perception of goodness-of-fit lacking in the left hand panel. 
@@
 
-```{r enof, fig.cap = "Cumulative hazard functions for neonatal mortality, Västerbotten.", fig.width = 7}
+```{r enof, fig.cap = "Cumulative hazard functions for neonatal mortality, Västerbotten 1861--1950.", fig.width = 7}
 yen <- y - 1000 * rate * x ##"Endo"
@@
 
-```{r plotcohort, fig.cap = "By birth month, cohorts, Västerbotten 1861-1950.", fig.height = 8, eval = TRUE}
+```{r plotcohort, fig.cap = "By birth month, cohorts, Västerbotten 1861--1950.", fig.height = 8, eval = TRUE}
 oldpar <- par(mfrow = c(1, 1))
@@
 
-```{r plotcohort3, fig.cap = "By month, period data, Västerbotten 1861-1950.", fig.height = 8, eval = TRUE}
+```{r plotcohort3, fig.cap = "By month, period data, Västerbotten 1861--1950.", fig.height = 8, eval = TRUE}
 oldpar <- par(mfrow = c(1, 1))
@@
 
-```{r plotperiod, fig.cap = "By calendar month, period data, Västerbotten 1861-1950.", fig.height = 9, eval = FALSE}
+```{r plotperiod, fig.cap = "By calendar month, period data, Västerbotten 1861--1950.", fig.height = 9, eval = FALSE}
 oldpar <- par(mfrow = c(4, 3))
@@
 
-## Scania 1710--1800.
+## Skåne 1710--1800.
 
-The Scanian dataset covers five parishes in Scania 1710--1800 [@sedd].
+Skåne is the southernmost province of Sweden, see the map in Figure \@ref(fig:scaniamap).
+
+```{r scaniamap, fig.cap = "Skåne."}
+knitr::include_graphics("Slides/myfigs/scania.png")
+```
+
+This dataset covers five parishes i n Skåne 1710--1800 [@sedd].
 It is an interesting case because it appears as if the data quality is low since 
@@
 
-```{r scahigh, fig.cap = "Scania 1710--1800."}
+```{r scahigh, fig.cap = "Skåne 1710--1800."}
 sca1 <- readRDS("Rdata/sca1.rds")

### 2023-03-27 paper2.Rmd
add No of births and deaths

@@
 and due to imperfect time measuring there are often tied death ages. As an illustrative example, 
-we use a data set from Västerbotten in northern Sweden, covering the years 1861--1950, see 
+we use a data set from Västerbotten in northern Sweden, covering $221\,773$ births with $17\,675$ infant deaths over the years 1861--1950, see 
 Figure \@ref(fig:vbmap).
@@
 
-This dataset covers five parishes i n Skåne 1710--1800 [@sedd].
+This dataset covers five parishes and 7461 births with 1654 infant deaths in Skåne 1710--1800 [@sedd].
 It is an interesting case because it appears as if the data quality is low since 
@@
 Specifically, the log-cube transformation, which fits well with 19th century data, seems
-to be completely out of order in the eighteenth century.
+to be completely out of order in the eighteenth century, maybe due to the high infant mortality,
+which in this dataset is 222 per thousand.
 

### 2023-03-29 Slides/slides.Rmd
new path to both
add basics

@@
 ---
-title: A hazards approach to the biometric analysis of infant mortality
+title: "A hazards approach to the biometric analysis of infant mortality"
 author:
@@
 institute:
+- \inst{2}CED, Lund University (tommy.bengtsson@ekh.lu.se)
 - \inst{1}CEDAR, Umeå University (gb@ehar.se, https://github.com/goranbrostrom)
-    - \inst{2}CED, Lund University (tommy.bengtsson@ekh.lu.se)
 output:
@@
     theme: Warsaw
+classoption: t
 ---
@@
 
-## The log-cube transform
+## Bourgeois-Pichat and causes of death
 
-\begin{columns}
+*   Two categories: 
+    *   Endogenous: *inherited*, *delivery*, etc.
+    *   Exogenous: *accidental*, *infectious diseases*, etc.
+*   How to differentiate between the two categories without infornation of causes of death?
+
+Bourgeois-Pichat's Biometric model.
+
+## The biometric model
+
+Two postulates:
+
+1.  Endogenous deaths only occur during the neonatal period (0--28 days). 
+2.  On a specific time scale, exogenous infant mortality is uniformly distributed.
+
+## The log-cube transform {.t}
+
+\begin{columns}[T]
 \begin{column}{0.6\textwidth}
@@
 \end{equation*}
+
+\tiny
+\emph{Note:} $C$ is not part of B-P's original definition.
+
 \end{column}
@@
 \end{columns}
+
+## Demonstration of the B-P plot
+
+Assume 100 births, of which 90 survives infancy, ten death ages:
+
+2, 6, 10, 15, 25, 41, 90, 120, 180, 319.
+
+```{r bpex, fig.height = 4}
+C <- 365 / (log(365 + 1))^3
+g <- function(x) C * (log(x + 1))^3
+d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 319)
+h <- seq_along(d) / 100
+plot(g(d), h, type = "b", lty = 3, xlab = "Day", axes = FALSE, ylab = "Deaths / 100", 
+     ylim = c(0, 0.11), xlim = c(0, 365))
+axis(2, las = 1)
+abline(h = 0, v = c(0, 365))
+axis(1, at = g(d), labels = d)
+axis(3, at = c(0, g(28), 365), labels = c(0, 28, 365))
+box()
+abline(v = g(28), lty = 2, col = "magenta")
+text(g(12), 0.08, "Neonatal", col = "red")
+text(g(130), 0.04, "Post-neonatal", col = "blue")
+```
+
+Calculation (no ties): $\frac{1}{100}, \frac{2}{100}, \frac{3}{100}, \ldots, \frac{10}{100}.$
+
+<!--
+\begin{equation*}
+p_n = \frac{n}{100}, \; n = 1, \ldots 10
+\end{equation*}
+-->
+
+## Demonstration of the B-P plot II
+
+Assume 100 births, of which 90 survives infancy, ten death ages:
+
+2, 6, 10, 15, 25, 41, 90, 120, 180, 319.
+
+```{r bpex2, fig.height = 4}
+C <- 365 / (log(365 + 1))^3
+g <- function(x) C * (log(x + 1))^3
+d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 319)
+h <- seq_along(d) / 100
+plot(g(d), h, type = "b", lty = 3, xlab = "Day", axes = FALSE, ylab = "Deaths / 100", 
+     ylim = c(0, 0.11), xlim = c(0, 365))
+
+abline(h = 0, v = c(0, 365))
+axis(1, at = g(d), labels = d)
+axis(3, at = c(0, g(28), 365), labels = c(0, 28, 365))
+box()
+abline(v = g(28), lty = 2, col = "magenta")
+text(g(12), 0.08, "Exogenous", col = "blue")
+text(g(12), 0.03, "Endogenous", col = "red")
+points(g(c(41, 319)), c(6/100, 10/100), col = "blue", lwd = 2)
+##
+fy <- function(x, xc, yc){
+    yc[1] + (yc[2] - yc[1]) / (xc[2] - xc[1]) * (x - xc[1])
+}
+x1 <- g(41)
+y1 <- 6/100
+x2 <- g(319)
+y2 <- 10/100
+x0 <- 0
+x3 <- g(319)
+y0 <- fy(x0, c(x1, x2), c(y1, y2))
+y3 <- fy(x3, c(x1, x2), c(y1, y2))
+lines(c(x0, x3), c(y0, y3), lty = 2)
+axis(2, las = 1, at = c(0, y0, y3), labels = round(c(0, y0, y3), 3))
+lines(c(0, g(319)), c(y3, y3), lty = 2)
+lines(c(0, 0), c(0, y0), col = "red", lwd = 3)
+lines(c(0, 0), c(y0, y3), col = "blue", lwd = 3)
+```
+
+Calculation (no ties): $\frac{1}{100}, \frac{2}{100}, \frac{3}{100}, \ldots, \frac{10}{100}.$
+
+<!--
+\begin{equation*}
+p_n = \frac{n}{100}, \; n = 1, \ldots 10
+\end{equation*}
+-->
+
+## Explanation of the hazards plot
+
+Assume 100 births, of which 90 survives infancy, ten death ages:
+
+2, 6, 10, 15, 25, 41, 90, 120, 180, 319 (same as before).
+
+```{r hazex, fig.height = 4}
+C <- 365 / (log(365 + 1))^3
+g <- function(x) C * (log(x + 1))^3
+d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 319)
+h <- seq_along(d) / (100:91)
+plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Cumulative hazards", 
+     ylim = c(0, 0.11), xlim = c(0, 365))
+h2 <-  seq_along(d) / 100
+lines(g(d), h2, type = "b", lty = 3) 
+axis(2, las = 1)
+abline(h = 0, v = c(0, 365))
+axis(1, at = g(d), labels = d)
+axis(3, at = c(0, g(28), 365), labels = c(0, 28, 365))
+box()
+abline(v = g(28), lty = 2, col = "magenta")
+text(g(12), 0.08, "Neonatal", col = "red")
+text(g(130), 0.04, "Post-neonatal", col = "blue")
+ttt <- sum(g(c(41, 90, 120, 180, 319))) + 90 * (g(365) - g(28))
+
+```
+
+Calculation (no ties): $\frac{1}{100}, \frac{2}{99}, \frac{3}{98}, \ldots, \frac{10}{91}.$
+
+## Explanation of the hazards plot II
+
+Assume 100 births, of which 90 survives infancy, ten death ages:
+
+2, 6, 10, 15, 25, 41, 90, 120, 180, 319 (same as before).
+
+```{r hazex2, fig.height = 4}
+C <- 365 / (log(365 + 1))^3
+g <- function(x) C * (log(x + 1))^3
+d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 319)
+h <- seq_along(d) / (100:91)
+plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Cumulative hazards", 
+     ylim = c(0, 0.11), xlim = c(0, 365))
+
+axis(2, las = 1)
+abline(h = 0, v = c(0, 365))
+axis(1, at = g(d), labels = d)
+axis(3, at = c(0, g(28), 365), labels = c(0, 28, 365))
+box()
+abline(v = g(28), lty = 2, col = "magenta")
+text(g(12), 0.08, "Neonatal", col = "red")
+text(g(130), 0.04, "Exogenous", col = "blue")
+ttt <- sum(g(c(41, 90, 120, 180, 319))) + 90 * (g(365) - g(28))
+lambda <- 5 / ttt
+x <- g(0:365)
+y <- Hweibull(x, shape = 1, scale = 1 / lambda)
+lines(x, y, lty = 2, col = "blue")
+```
 
@@
 
+Remove?
+
 \includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/both-1.pdf}

### 2023-03-30 Slides/slides.Rmd
educational part complete

--- 2023-03-15 Slides/slides.Rmd
@@
 \end{columns}
+
+## Demonstration of the B-P plot
+
+Assume 100 births, of which 90 survives infancy, ten death ages:
+
+2, 6, 10, 15, 25, 41, 90, 120, 180, 309.
+
+```{r bpex, fig.height = 5}
+par(lwd = 2)
+C <- 365 / (log(365 + 1))^3
+g <- function(x) C * (log(x + 1))^3
+d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 309)
+h <- seq_along(d) / 100
+plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Deaths / 100", 
+     ylim = c(0, 0.12), xlim = c(0, 365))
+axis(2, las = 1)
+abline(h = 0, v = c(0, 365))
+axis(1, at = g(d), labels = d)
+axis(3, at = c(0, g(28), 365), labels = c(0, 28, 365))
+box()
+abline(v = g(28), lty = 2, col = "magenta")
+text(30, 0.11, "Neonatal", col = "red")
+text(110, 0.11, "Post-neonatal", col = "blue")
+points(g(c(41, 309)), c(6/100, 10/100), col = "red", lwd = 2, pch = 4)
+```
+
+Calculation (no ties): $\frac{1}{100}, \frac{2}{100}, \frac{3}{100}, \ldots, \frac{10}{100}.$
+
+<!--
+\begin{equation*}
+p_n = \frac{n}{100}, \; n = 1, \ldots 10
+\end{equation*}
+-->
+
+## Demonstration of the B-P plot II
+
+Assume 100 births, of which 90 survives infancy, ten death ages:
+
+2, 6, 10, 15, 25, 41, 90, 120, 180, 309.
+
+```{r bpex2, fig.height = 5}
+par(lwd = 2)
+C <- 365 / (log(365 + 1))^3
+g <- function(x) C * (log(x + 1))^3
+d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 309)
+h <- seq_along(d) / 100
+plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Deaths / 100", 
+     ylim = c(0, 0.11), xlim = c(0, 365))
+
+abline(h = 0, v = c(0, 365))
+axis(1, at = g(d), labels = d)
+axis(3, at = c(0, g(28), 365), labels = c(0, 28, 365))
+box()
+abline(v = g(28), lty = 2, col = "magenta")
+text(g(12), 0.08, "Exogenous", col = "blue", cex = 0.8)
+text(g(12), 0.03, "Endogenous", col = "red", cex = 0.8)
+points(g(c(41, 309)), c(6/100, 10/100), col = "blue", lwd = 2)
+##
+fy <- function(x, xc, yc){
+    yc[1] + (yc[2] - yc[1]) / (xc[2] - xc[1]) * (x - xc[1])
+}
+x1 <- g(41)
+y1 <- 6/100
+x2 <- g(309)
+y2 <- 10/100
+x0 <- 0
+x3 <- g(309)
+y0 <- fy(x0, c(x1, x2), c(y1, y2))
+y3 <- fy(x3, c(x1, x2), c(y1, y2))
+lines(c(x0, x3), c(y0, y3), lty = 2, lwd = 1)
+axis(2, las = 1, at = c(0, y0, y3), labels = round(c(0, y0, y3), 3))
+lines(c(0, g(309)), c(y3, y3), lty = 2, lwd = 1)
+lines(c(0, 0), c(0, y0), col = "red", lwd = 3)
+lines(c(0, 0), c(y0, y3), col = "blue", lwd = 3)
+```
+
+Calculation (no ties): $\frac{1}{100}, \frac{2}{100}, \frac{3}{100}, \ldots, \frac{10}{100}.$
+
+<!--
+\begin{equation*}
+p_n = \frac{n}{100}, \; n = 1, \ldots 10
+\end{equation*}
+-->
+
+## The hazards plot
+
+Assume 100 births, of which 90 survives infancy, ten death ages:
+
+2, 6, 10, 15, 25, 41, 90, 120, 180, 309 (same as before).
+
+```{r hazex, fig.height = 5}
+par(lwd = 2)
+C <- 365 / (log(365 + 1))^3
+g <- function(x) C * (log(x + 1))^3
+d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 309)
+h <- seq_along(d) / (100:91)
+plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Cumulative hazards", 
+     ylim = c(0, 0.12), xlim = c(0, 365))
+h2 <-  seq_along(d) / 100
+lines(g(d), h2, type = "b", lty = 3, lwd = 1) 
+axis(2, las = 1)
+abline(h = 0, v = c(0, 365))
+axis(1, at = g(d), labels = d)
+axis(3, at = c(0, g(28), 365), labels = c(0, 28, 365))
+box()
+abline(v = g(28), lty = 2, col = "magenta")
+text(g(12), 0.11, "Neonatal", col = "red")
+text(g(110), 0.11, "Post-neonatal", col = "blue")
+ttt <- sum(g(c(41, 90, 120, 180, 309))) + 90 * (g(365) - g(28))
+
+```
+
+Calculation (no ties): $\frac{1}{100}, \frac{2}{99}, \frac{3}{98}, \ldots, \frac{10}{91}.$
+
+## The hazards plot II
+
+<!-- Assume 100 births, of which 90 survives infancy, ten death ages:
+
+2, 6, 10, 15, 25, 41, 90, 120, 180, 319 (same as before).
+-->
+
+```{r hazex2, fig.height = 5}
+par(mai = c(1,1,1,1), lwd = 2)
+C <- 365 / (log(365 + 1))^3
+g <- function(x) C * (log(x + 1))^3
+d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 309)
+h <- seq_along(d) / (100:91)
+plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Cumulative hazards", 
+     ylim = c(0, 0.13), xlim = c(0, 365))
+
+axis(2, las = 1)
+abline(h = 0, v = c(0, 365))
+axis(1, at = g(c(d, 365)), labels = c(d, 365))
+tabb <- g(c(0, 28, 41, 90,120, 180, 309, 365))
+axis(3, at = tabb, labels = round(tabb))
+box()
+abline(v = g(28), lty = 2, col = "magenta")
+##text(g(12), 0.08, "Neonatal", col = "red")
+text(g(130), 0.11, "Total")
+text(g(130), 0.05, "Exogenous", col = "blue")
+ttt <- sum(g(c(41, 90, 120, 180, 319) - 28)) + 90 * (g(365) - g(28))
+lambda <- 5 / ttt
+x <- g(0:365)
+y <- Hweibull(x, shape = 1, scale = 1 / lambda)
+lines(x, y, lty = 2, col = "blue")
+axis(4, at = c(0, y[366], h[10]), labels = round(c(0, y[366], h[10]), 3), las = 1)
+lines(c(365, 365), c(0, y[366]), col = "blue", lwd = 3)
+lines(c(365, 365), c(y[366], h[10]), col = "red", lwd = 3)
+text(30, 0.12, "Neonatal", col = "red", cex = 0.8)
+text(110, 0.12, "Post-neonatal", col = "blue", cex = 0.8)
+```
+
+Exogenous slope = $\frac{\#\text{(post deaths)}}{\text{post exposure}} = \frac{5}{27626} \approx 0.00018$
+
+Post exposure = $90 (365 - 68) + (93 - 68) + (163 - 68) + \cdots + (335 - 68) \approx 27626$
+
+## Adding the endogenous mortality
+
+```{r hazex3, fig.height = 5}
+par(mai = c(1,1,1,1), lwd = 2)
+C <- 365 / (log(365 + 1))^3
+g <- function(x) C * (log(x + 1))^3
+d <- c(2, 6, 10, 15, 25, 41, 90, 120, 180, 309)
+h <- seq_along(d) / (100:91)
+plot(g(d), h, type = "b", lty = 1, xlab = "Day", axes = FALSE, ylab = "Cumulative hazards", 
+     ylim = c(0, 0.13), xlim = c(0, 365))
+
+axis(2, las = 1)
+abline(h = 0, v = c(0, 365))
+axis(1, at = g(c(d, 365)), labels = c(d, 365))
+tabb <- g(c(0, 28, 41, 90,120, 180, 309, 365))
+##axis(3, at = tabb, labels = round(tabb))
+box()
+abline(v = g(28), lty = 2, col = "magenta")
+##text(g(12), 0.08, "Neonatal", col = "red")
+text(g(130), 0.11, "Total")
+text(g(130), 0.02, "Exogenous", col = "blue")
+text(g(130), 0.06, "Endogenous", col = "red")
+ttt <- sum(g(c(41, 90, 120, 180, 319) - 28)) + 90 * (g(365) - g(28))
+lambda <- 5 / ttt
+x <- g(0:365)
+y <- Hweibull(x, shape = 1, scale = 1 / lambda)
+lines(x, y, lty = 2, col = "blue")
+axis(4, at = c(0, y[366], h[10]), labels = round(c(0, y[366], h[10]), 3), las = 1)
+lines(c(365, 365), c(0, y[366]), col = "blue", lwd = 3)
+lines(c(365, 365), c(y[366], h[10]), col = "red", lwd = 3)
+text(30, 0.12, "Neonatal", col = "red", cex = 0.8)
+text(110, 0.12, "Post-neonatal", col = "blue", cex = 0.8)
+## Add neonatal:
+xn <- g(d)
+yn <- h - Hweibull(xn, shape = 1, scale = 1 / lambda)
+lines(xn, yn, type = "l", col = "red", lty = 4)
+```
 

### 2023-03-30 Slides/slides.Rmd
finish

@@
 
+\tiny
+
 Post exposure = $90 (365 - 68) + (93 - 68) + (163 - 68) + \cdots + (335 - 68) \approx 27626$
@@
 text(110, 0.12, "Post-neonatal", col = "blue", cex = 0.8)
-## Add neonatal:
+## Add endogenous:
 xn <- g(d)
@@
 lines(xn, yn, type = "l", col = "red", lty = 4)
+lines(c(g(28), g(309)), c(yn[6], yn[6]), lty = 4, lwd = 1)
 ```
@@
 
-*   Knodel \& Kintner (1977) 
+*   Knodel \& Kintner (1977).
+    *    Discusses effects of breastfeeding habits.
 *   Wrigley (1977).
+    *    Problems with data recording.
 *   Lynch, Greenhouse \& Brändström (1998).
+    *    Mentions "constant hazard".
+    *    Suggest linear regression methods.
 *   Bengtsson (1999).
+    *    Data quality.
 *   Manfredini (2004).
+    *    Deals with effect of climate variation.
 
@@
 
+## Cohort effect of birth month, Västerbotten
+
+\begin{columns}
+\begin{column}{0.6\textwidth}
+
+\includegraphics[width=0.8\textwidth,height=0.7\textheight]{../paper2_files/figure-latex/plotcohort-1.pdf}
+
+\end{column}
+
+\begin{column}{0.4\textwidth}
+
+Conclusion:
+\vspace{2\baselineskip}
+
+Post-neonatal linearity is a population property, not an individual.
+\end{column}
+
+\end{columns}
+
+## Period effect of month, Västerbotten
+
+\begin{columns}
+
+\begin{column}{0.6\textwidth}
+
+\includegraphics[width=0.8\textwidth,height=0.7\textheight]{../paper2_files/figure-latex/plotcohort3-1.pdf}
+
+\end{column}
+
+\begin{column}{0.4\textwidth}
+
+Looks quite proportional!
+
+\end{column}
+
+\end{columns}
+
+<!--
 ## Comparison
@@
 
+-->
+
 ## Scania 1711--1800

### 2023-04-02 Slides/slides.Rmd
add normal time

@@
 
-##
+## Skellefteå 1801--1820
 
@@
 
-*   Good
+*   Advantages of the hazards method.
+       *    Fits naturally into modern survival analysis, allows censored and truncated data, and 
+       *   The exogenous component forms a class of proportional hazards distributions.
+
+*   Restriction for both models.
+    *   Models for populations, not for individuals.
+    *   Less useful for traditional survival analysis.
+    *.  The log-cube transform is not universal.

### 2023-04-03 paper2.Rmd
add normal time
add Weibull, Waloddi to references

@@
 
-We. may be relatively confident that the shape is OK in the post-neonatal period, where data are available, but the shape during the neonatal period is a pure guess.
+We may be relatively confident that the shape is OK in the post-neonatal period, where data are available, but the shape during the neonatal period is a pure guess.
 Examining data with *known* causes of death might be enlightening.
+
+We repeat the hazards method for Västerbotten 1961--1890 on both the log-cube 
+and the natural time scales for comparison, see Figure \@ref(fig:natlogcube2).
+
+```{r natlogcube2, fig.cap = "Västerbotten 1861-1890, log-cube and natural time scales."}
+par(mfrow = c(1, 2), las = 1, lwd = 2)
+rate <- drawBB2(g1, main = "Log-cube", ylim = c(0, 0.11))
+fit <- coxreg(Surv(enter, exit, event) ~ 1, data = g1)
+##plot(fit, main = "Natural", lwd = 3)
+hz <- hazards(fit)
+x <- c(0, hz[[1]][, 1])
+y1 <- c(0, cumsum(hz[[1]][, 2]))
+plot(x, y1, type = "l", axes = FALSE, xlab = "Days", main = "Natural",
+     ylab = "Cumulative hazards", ylim = c(0, 0.11))
+axis(1, at = c(0, 28, 91, 183, 274, 365))
+axis(2, at = c(0, 0.015, 0.027, 0.042, 0.080, 0.108))
+box()
+
+abline(v = 28, lty = 2, lwd = 1)
+abline(h = 0, lwd = 1)
+y <- g(x) * rate
+lines(x, y, lty = 2, col = "blue")
+## And the Endogenous:
+y2 <- y1 - y
+lines(x, y2, lty = 4, col = "red")
+```
+
+The picture on the natural time scale is of course what we prefer, so why did Bourgeois-Pichat use the log-cube time transform instead of the log-cube cumulative hazards function?
+The answer is of course that he had no access to a computer back in the fifties, and his way of compensating for that was in fact very clever.
+Similar pencil-and-ruler techniques have earlier been used, for instance in *Weibull probability plotting* [@ww51].
 
@@
 
-## Skellefteå 1820--1835
+## Skellefteå 1821--1838
 

### 2023-04-03 Slides/slides.Rmd
improve hazard plot

@@
 
+## Biometric analysis on the natural time scale
+
+Given that the exogenous mortality follows an Exponential distribution on the log-cube scale, it follows that the cumulative hazards function on the natural time scale is
+
+\begin{equation*}
+H(t; \lambda) = \lambda \log^3(t + 1), \; t > 0,
+\end{equation*}
+
+and it follows that the hazard function $h$ is given by
+
+\begin{equation*}
+h(t; \lambda) = \frac{3 \log^2(t + 1)}{t + 1}, \; t > 0,
+\end{equation*}
+
+## Exogenous mortality on the normal time scale
+
+```{r normal, fig.height = 4}
+dg <- function(x) 3 * (log(1 + x))^2 / (1 + x) * 365 / (log(366))^3
+oldpar <- par(mfrow = c(1, 2))
+x <- 0:365
+y <- dg(x)
+plot(x, y, type = "l", xlab = "Day", ylab = "Hazards",
+     axes = FALSE, col = "blue", lwd = 2, main = "Hazard function")
+axis(1, at = c(0, 28, 183, 365))
+axis(2, las = 1, at = 0)
+box()
+abline(h = 0, v = 0)
+abline(v = 28, lty = 2, col = "darkgreen")
+y <- g(x)
+plot(x, y, type = "l", xlab = "Day", ylab  = "Cum. hazards", main = "Cumulative hazards",
+     col = "blue")
+par(oldpar)
+```
+
+## Natural vs. log-cube time scales
+
+Västerbotten 1861--1890.
+
+\includegraphics[width=\textwidth]{../paper2_files/figure-latex/natlogcube2-1.pdf}
+
 ## Conclusions

### 2023-04-03 R/drawBB2.R

@@ -37,8 +37,8 @@ drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3,
    y1 <- y[n]
    red <- y[1:n] - x[1:n] * rate
    redall <- y - x * rate
-   lines(x[1:n], red, col = "red", type = "s", lty = 4, lwd = 2)
-   lines(x, redall, col ="magenta", type = "s", lty = 3, lwd = 2)
+   lines(x[1:n], red, col = "red", type = "l", lty = 4, lwd = 2)
+   lines(x, redall, col ="red", type = "l", lty = 4, lwd = 2)
    y2 <- red[n]
    y3 <- max(y)
    y4 <- rate * g(365)
@@ -46,7 +46,7 @@ drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3,
    box()
    lines(c(0, g(28)), c(y0, y0), lty = 3, lwd = 0.5, col = "darkgreen")
    lines(c(0, g(28)), c(y1, y1), lty = 3, lwd = 0.5, col = "darkgreen")
-   lines(c(0, g(365)), c(y2, y2), lty = 3, lwd = 1.5, col = "darkgreen")
+   lines(c(0, g(28)), c(y2, y2), lty = 3, lwd = 1.5, col = "darkgreen")
    abline(h = c(y3, y4), lty = 3, lwd = 0.5, col = "darkgreen")
    abline(v = 0, lwd = 0.5)
    lines(c(0, g(28)), c(0, 0), lwd = 0.5)
@@ -83,5 +83,6 @@ drawBB2 <- function(dat, main = "", ylim, C = 365 / (log(366))^3,
    axis(2, at = round(c(0, y0, y1, y2), 3), las = 1)
    box()
    lines(c(0, x2), c(y0, y2), type = "l", lty = 2, col = "darkgreen")
-   }
+     }
+   invisible(rate)
 }

### 2023-04-04 paper2.Rmd
fix errors and add on natural time scale

@@
 
-```{r vbmap, fig.cap="Västerbotten."}
+```{r vbmap, fig.cap="Västerbotten.", out.width="6in"}
 knitr::include_graphics("Slides/myfigs/vb.png")
@@
 
-Given that the exogeneous mortality follows an Exponential distribution on the log-cube scale, it follows that the cumulative hazards function on the natural time scale is
+Given that the exogenous mortality follows an Exponential distribution on the log-cube scale, 
+it follows that the cumulative hazards function on the natural time scale is
 
 \begin{equation*}
-H(t; \lambda) = \lambda \log^3(t + 1), \; t > 0,
+H(t; \lambda) = \lambda \log^3(t + 1), \; t > 0.
 \end{equation*}
 
-and it follows that the hazard function $h$ is given by
+Note the duality: $\text{log-cube transform} \Longleftrightarrow  \text{cumulative hazards function}$!
+
+It follows that the hazard function $h$ is given by
 
 \begin{equation*}
-h(t; \lambda) = \frac{3 \log^2(t + 1)}{t + 1}, \; t > 0,
+h(t; \lambda) = \lambda \frac{3 \log^2(t + 1)}{t + 1}, \; t > 0,
 \end{equation*}
@@
 
-```{r scaniamap, fig.cap = "Skåne."}
+```{r scaniamap, fig.cap = "Skåne.", out.width="6in"}
 knitr::include_graphics("Slides/myfigs/scania.png")
@@
 to be completely out of order in the eighteenth century, maybe due to the high infant mortality,
-which in this dataset is 222 per thousand.
+which in this data set is 222 per thousand.
 
@@
 
-<!--- Let us look at the estimated cumulative hazards, total, 
-exogenous and endgenous, see Figure \@ref(fig:escu).
+Let us look at what happens on the natural time scale and with the Gompertz distribution 
+replacing the log-cube time transformation, see Figure \@ref(fig:escu).
 
-```{r escu, fig.cap = "Scania 1710-1800.", eval = FALSE}
+```{r escu, fig.cap = "Skåne 1710--1800, natural time scale, Gompertz exogenous distribution."}
 #source("R/biometric.R")
 #biometric(sca1, graph = TRUE)
-post <- age.window(sca1, c(g(28), 365))
-lambda <- with(post, sum(event) / sum(exit - enter))
-x <- c(0, rs$risktimes)
-y <- x * lambda
-hazo <- with(rs, c(0, n.events / size))
-hazo <- cumsum(hazo)
-plot(x, hazo, type = "s", col = "blue")
-lines(x, y)
-z <- hazo - y
-lines(x, z, col = "red")
+par(mfrow = c(1, 1), las = 1, lwd = 2, cex = 0.7)
+fit.c <- coxreg(Surv(enter, exit, event) ~ 1, data = sca11)
+hc <- hazards(fit.c)
+x <- c(0, hc[[1]][, 1])
+y <- c(0, cumsum(hc[[1]][, 2]))
+post <- age.window(sca11, c(28, 365))
+fit.g <- phreg(Surv(enter, exit, event) ~ 1, data = post, dist = "gompertz", param = "rate")
+hg <- hazards(fit.g)
+plot(x, y, type = "l", xlab = "Day", ylab = "Cumulative hazards", axes = FALSE)
+axis(1, at = c(0, 28, 183, 365))
+att <-  c(0, 0.041, 0.063, 0.104, 0.216, 0.278)
+axis(2, at = att)
+abline(h = att, lty = 3, lwd = 1)
+abline(h = 0, lwd = 1) 
+abline(v = 28, lty = 2, lwd = 1, col = "darkgreen")
+abline(v = 0, lwd = 1)
+box()
+y1 <- Hgompertz(x, shape = exp(-6.4115), rate = -0.007017,  param = "rate")
+lines(x, y1, col = "blue", lty = 2)
+xp <- x[c(0, 31:320)]
+yp <- y - y1
+lines(x, yp, col = "red", lty = 4)
 ```
 
---->
+Quite okay, if not perfect,
 
@@
 -->
-Apparently, it works for many places in the mid-twentieth century, but beyond that, it is not generalizable.
-In fact, also when post-neonatal mortality is following a straight line on the log-cube scale, mortality may well deviate during different parts of the year.
-This means that the methods should be used with great care and only as a starting point.
+Apparently, it works for many places in the mid-twentieth century, but beyond that, 
+it is not generalisable. In fact, also when post-neonatal mortality is following a 
+straight line on the log-cube scale, mortality may well deviate during different 
+parts of the year. This means that the methods should be used with great care and 
+only as a starting point.
 

### 2023-04-04 Slides/slides.Rmd
updates from paper

--- 2023-03-30 Slides/slides.Rmd
@@
   beamer_presentation:
+#      includes: 
+#         header: gb_header.txt
+#         before_body: gb_beforebody.txt
       theme: Warsaw
 classoption: t
+header-includes:
+- \newcommand{\emp}[1]{\textcolor{blue}{#1}}
 ---
@@
 *   Two categories: 
-    *   Endogenous: *inherited*, *delivery*, etc.
-    *   Exogenous: *accidental*, *infectious diseases*, etc.
+    *   \emp{Endogenous}: *inherited*, *delivery*, etc.
+    *   \emp{Exogenous}: *accidental*, *infectious diseases*, etc.
 *   How to differentiate between the two categories without infornation of causes of death?
 
-Bourgeois-Pichat's Biometric model.
+Bourgeois-Pichat's \emp{biometric model}.
 
@@
 
-1.  Endogenous deaths only occur during the neonatal period (0--28 days). 
-2.  On a specific time scale, exogenous infant mortality is uniformly distributed.
+1.  \emp{Endogenous} deaths only occur during the \emp{neonatal} period (0--28 days). 
+2.  On a specific time scale, \emp{exogenous} infant mortality is \emp{uniformly} distributed.
 
@@
 \tiny
-\emph{Note:} $C$ is not part of B-P's original definition.
+\emph{Note:} $C$ is \emp{not} part of B-P's original definition.
 
@@
 *   Knodel \& Kintner (1977).
-    *    Discusses effects of breastfeeding habits.
+    *    Discusses effects of \emp{breastfeeding} habits.
 *   Wrigley (1977).
-    *    Problems with data recording.
+    *    Problems with \emp{data recording}.
 *   Lynch, Greenhouse \& Brändström (1998).
-    *    Mentions "constant hazard".
-    *    Suggest linear regression methods.
+    *    Mentions \emp{"constant hazard"}.
+    *    Suggest \emp{linear regression} methods.
 *   Bengtsson (1999).
-    *    Data quality.
+    *    Data \emp{quality}.
 *   Manfredini (2004).
-    *    Deals with effect of climate variation.
+    *    Deals with effect of \emp{climate} variation.
 
@@
 
+## Biometric analysis on the natural time scale
+
+Given that the exogenous mortality follows an Exponential distribution on the log-cube scale, 
+it follows that the \emp{cumulative hazards} function \emp{$H$} on the \emp{natural} time scale is
+
+\begin{equation*}
+H(t; \lambda) = \lambda \log^3(t + 1), \; t \ge 0,
+\end{equation*}
+
+and it follows that the \emp{hazard} function \emp{$h$} is given by
+
+\begin{equation*}
+h(t; \lambda) = \lambda \frac{3 \log^2(t + 1)}{t + 1}, \; t \ge 0,
+\end{equation*}
+
+As $\lambda$ varies, this constitutes a \emp{proportional hazards} family of distributions.
+
+## Exogenous mortality on the normal time scale
+
+```{r normal, fig.height = 4}
+dg <- function(x) 3 * (log(1 + x))^2 / (1 + x) * 365 / (log(366))^3
+oldpar <- par(mfrow = c(1, 2), las = 1)
+x <- 0:365
+y <- dg(x)
+plot(x, y, type = "l", xlab = "Day", ylab = "Hazards",
+     axes = FALSE, col = "blue", lwd = 2, main = "Hazard function")
+axis(1, at = c(0, 28, 183, 365))
+axis(2, las = 1, at = 0)
+box()
+abline(h = 0, v = 0)
+abline(v = 28, lty = 2, col = "darkgreen")
+y <- g(x)
+plot(x, y, type = "l", xlab = "Day", ylab  = "Cum. hazards", main = "Cumulative hazards",
+     col = "blue", axes = FALSE)
+axis(1, at = c(0, 28, 183, 365))
+axis(2, las = 1, at = 0)
+abline(h = 0, v = 0)
+box()
+par(oldpar)
+```
+
+## Natural vs. log-cube time scales
+
+\emp{Västerbotten} 1861--1890.
+
+\includegraphics[width=\textwidth]{../paper2_files/figure-latex/natlogcube2-1.pdf}
+
+## Natural time scale and Gompertz exogenous distribution
+
+\emp{Skåne} 1711--1800
+
+\includegraphics[width = \textwidth]{../paper2_files/figure-latex/escu-1.pdf}
+
 ## Conclusions
 
-*   Good
+*   \emp{Advantages} of the hazards method.
+       *    Fits naturally into modern survival analysis, allows \emp{censored} and \emp{truncated data}, and 
+       *    The \emp{exogenous} component forms a class of \emp{proportional hazards} distributions.
+
+*   \emp{Restrictions} for both models.
+    *   Models for \emp{populations}, not for individuals.
+    *   \emp{Less useful} for traditional \emp{survival analysis}.
+    *   The \emp{log-cube} transform is \emp{not} universal.
+
+*   \emp{Suggestions}.
+    *   Work on the \emp{natural} time scale, if access to a computer.
+    *   Try \emp{standard} distributions for \emp{exogenous}  mortality. 
+    *   \emp{Remember} that we are dealing with \emp{competing risks}!
+        *   Requires special care.

### 2023-04-05 Slides/slides.Rmd
polishing

@@
 
-*   Bourgeois-Pichat (1951): Biometric analysis of IM (background).
-*   Hazard-based alternative.
+*   Jean \emp{Bourgeois-Pichat} (1951): \emp{Biometric} analysis of infant mortality (background).
+*   \emp{Hazard}-based \emp{alternative}.
 *   Theoretical considerations
@@
     *   \emp{Exogenous}: *accidental*, *infectious diseases*, etc.
-*   How to differentiate between the two categories without infornation of causes of death?
+*   How to \emp{differentiate} between the two categories \emp{without information} of causes of death?
 
@@
 2.  On a specific time scale, \emp{exogenous} infant mortality is \emp{uniformly} distributed.
+    *   The \emp{log-cube} time scale.
 
@@
 
-\tiny
+$t$ is age in days.
+
+\vspace{\baselineskip}
+
+\tiny{
 \emph{Note:} $C$ is \emp{not} part of B-P's original definition.
+}
 
@@
 
-Assume 100 births, of which 90 survives infancy, ten death ages:
+Assume \emp{100 births}, of which \emp{90 survives} infancy, \emp{ten death ages}:
 
@@
 
+By \emp{subtraction}: $\text{Endogenous} = \text{Total} - \text{Exogenous}$
+
 ```{r hazex3, fig.height = 5}
@@
 
-## Väterbotten 1861-1890
+## Västerbotten 1861-1890
 
@@
 
-\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/bioume9120-1.pdf}
+\includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/lowumee-1.pdf}
 
@@
 
-Post-neonatal linearity is a population property, not an individual.
+Post-neonatal \emp{linearity} is a \emp{population property}, not an individual.
+
 \end{column}
@@
        *    The \emp{exogenous} component forms a class of \emp{proportional hazards} distributions.
+       *    So, theoretically more appropriate.
 
@@
 
-*   \emp{Suggestions}.
-    *   Work on the \emp{natural} time scale, if access to a computer.
-    *   Try \emp{standard} distributions for \emp{exogenous}  mortality. 
+## Suggestions for future work
+
+*   Work on the \emp{natural} time scale, if 
+    *   access to a computer, and
+    *   suitable software.
+*   Try \emp{standard} distributions for \emp{exogenous}  mortality, eg.
+    *   \emp{Weibull},
+    *   \emp{Gompertz}.
 *   \emp{Remember} that we are dealing with \emp{competing risks}!
     *   Requires special care.
+
+[https://github.com/goranbrostrom/ESSHC_23](https://github.com/goranbrostrom/ESSHC_23)
+
+**R** code for this paper and talk.

### 2023-04-05 Slides/slides.Rmd
polishing slides

@@
 
-$t$ is age in days.
+$t$ is \emp{age in days}.
 
@@
 
-Calculation (no ties): $\frac{1}{100}, \frac{2}{100}, \frac{3}{100}, \ldots, \frac{10}{100}.$
+Calculation (no ties): $(2, \frac{1}{100}), (6, \frac{2}{100}), (10, \frac{3}{100}), \ldots, (309, \frac{10}{100}).$
 
@@
 
-\tiny
-
 Post exposure = $90 (365 - 68) + (93 - 68) + (163 - 68) + \cdots + (335 - 68) \approx 27626$
@@
 
+Total mortality: $1 - \exp(-0.106) \approx 0.101$.
+
+Excellent fits!
+
 ## Västerbotten 1921--1950
@@
 \includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/lowumee-1.pdf}
+
+Total mortality: $1 - \exp(-0.056) \approx 0.054$.
+
+Excellent fits!
 
@@
 
+The log-cube model \emp{does not fit} at all, despite \emp{high-quality} data!
+
+\emp{Too old} time period?
+
 ## Skellefteå 1801--1820
@@
 \includegraphics[width=0.8\textwidth]{../paper2_files/figure-latex/skum-1.pdf}
+
+*   Reasonable linear post-neonatal fit.
+*   But \emp{no} room for \emp{endogenous} mortality.
+*   Many \emp{early deaths missing}.
 
@@
 
-## Exogenous mortality on the normal time scale
+\emp{NOTE:} $H(t; \lambda)$ is the \emp{log-cube transform!}
+
+## Exogenous mortality on the natural time scale
 
@@
 
-\emp{Skåne} 1711--1800
+\emp{Skåne} 1711--1800.
 
 \includegraphics[width = \textwidth]{../paper2_files/figure-latex/escu-1.pdf}
+
+\emp{Decent fit} if we skip the log-cube distribution!
 
@@
 
+**R** code for this paper:
+
 [https://github.com/goranbrostrom/ESSHC_23](https://github.com/goranbrostrom/ESSHC_23)
-
-**R** code for this paper and talk.
